---
title: "CRIME-Q critical-appraisal assessment"
subtitle: "Prenatal chronic-noise exposure and offspring anxiety- and depression-like behaviours"
format:
  html:
    theme: litera
    toc: true
    toc-depth: 2
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    code-copy: true
    embed-resources: true
    page-layout: full
    smooth-scroll: true
    fig-format: svg
execute:
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup-crimeq
#| echo: true

library(dplyr)
library(ggplot2)
library(here)
library(knitr)
library(readxl)
library(stringr)
library(tibble)
library(tidyr)

theme_set(theme_minimal(base_size = 12))

final_file <- here("Data", "CRIME-Q_Appraisal_Template.xlsx")
reviewer_1_file <- here("Data", "CRIME-Q_Appraisal_Template_Anna.xlsx")
input_sheet <- "CRIME-Q_Appraisal"
codebook_sheet <- "Codebook"

item_spec <- tribble(
  ~Item,   ~data_column,
  "1X",    "1X_peer_review",
  "3X",    "3X_animals_reporting",
  "3Y",    "3Y_animals_methodology",
  "3Z",    "3Z_animals_selection_bias_baseline_characteristics_syrcle_item_2",
  "4Y",    "4Y_sample_size_calculation",
  "5X",    "5X_in_vivo_reporting",
  "5Y",    "5Y_in_vivo_methodology",
  "5Z(1)", "5Z1_sequence_generation",
  "5Z(2)", "5Z2_random_housing",
  "5Z(3)", "5Z3_random_outcome_assessment",
  "6X",    "6X_welfare_compliance",
  "7X",    "7X_blinding",
  "8X",    "8X_methods_results_congruency",
  "8Z(1)", "8Z1_attrition_bias_incomplete_outcome_data",
  "8Z(2)", "8Z2_reporting_bias_selective_outcome",
  "9X",    "9X_limitations",
  "10X",   "10X_statement_conflict_of_interest",
  "10Z",   "10Z_publication_bias",
  "11X",   "11X_authors_contributions",
  "12X",   "12X_analysis_software",
  "13X",   "13X_data_availability"
)

rating_levels <- c("Yes", "Partly", "Unclear", "No", "NA")

read_appraisal <- function(path) {
  read_excel(
    path = path,
    sheet = input_sheet,
    col_types = "text",
    na = ""
  ) |>
    rename_with(str_trim) |>
    mutate(across(everything(), ~ if_else(is.na(.x), NA_character_, str_trim(.x))))
}

validate_appraisal <- function(dat, source_name) {
  required_cols <- c("study_id", "study_title", item_spec$data_column)
  missing_cols <- setdiff(required_cols, names(dat))

  if (length(missing_cols) > 0L) {
    stop(
      source_name,
      " is missing required column(s): ",
      paste(missing_cols, collapse = ", ")
    )
  }

  if (anyDuplicated(dat$study_id) > 0L) {
    duplicated_ids <- unique(dat$study_id[duplicated(dat$study_id)])
    stop(
      source_name,
      " contains duplicated study_id value(s): ",
      paste(duplicated_ids, collapse = ", ")
    )
  }

  rating_dat <- dat |>
    select(study_id, all_of(item_spec$data_column)) |>
    pivot_longer(
      cols = all_of(item_spec$data_column),
      names_to = "data_column",
      values_to = "Rating"
    )

  if (anyNA(rating_dat$Rating)) {
    missing_ratings <- rating_dat |>
      filter(is.na(Rating)) |>
      transmute(location = paste(study_id, data_column, sep = " / ")) |>
      pull(location)
    stop(
      source_name,
      " contains missing CRIME-Q rating(s): ",
      paste(missing_ratings, collapse = "; ")
    )
  }

  unexpected <- setdiff(unique(rating_dat$Rating), rating_levels)
  if (length(unexpected) > 0L) {
    stop(
      source_name,
      " contains unexpected rating value(s): ",
      paste(unexpected, collapse = ", ")
    )
  }

  invisible(dat)
}

consensus <- read_appraisal(final_file)
reviewer_1 <- read_appraisal(reviewer_1_file)

validate_appraisal(consensus, basename(final_file))
validate_appraisal(reviewer_1, basename(reviewer_1_file))

if (!setequal(consensus$study_id, reviewer_1$study_id)) {
  stop("The reviewer workbooks do not contain the same study_id values.")
}

normalise_item <- function(x) str_replace_all(x, "[^[:alnum:]]", "")

codebook_raw <- read_excel(
  path = final_file,
  sheet = codebook_sheet,
  skip = 7,
  col_types = "text",
  na = ""
) |>
  rename_with(str_trim) |>
  mutate(across(everything(), ~ if_else(is.na(.x), NA_character_, str_squish(.x))))

codebook <- codebook_raw |>
  transmute(
    Item_raw = Items,
    Item_key = normalise_item(Items),
    Type,
    Definition = Questions
  ) |>
  inner_join(
    item_spec |>
      mutate(Item_key = normalise_item(Item)),
    by = "Item_key"
  ) |>
  mutate(
    Ratings = case_when(
      str_detect(Definition, fixed("Yes/Partly/No")) ~ "Yes; Partly; No",
      str_detect(Definition, fixed("Yes/No/Unclear")) ~ "Yes; No; Unclear",
      str_detect(Definition, fixed("Yes/No/NA")) ~ "Yes; No; NA",
      str_detect(Definition, fixed("Yes/No")) ~ "Yes; No",
      TRUE ~ NA_character_
    ),
    Definition = Definition |>
      str_remove("\\s+Yes/Partly/No\\s*$") |>
      str_remove("\\s+Yes/No/Unclear\\s*$") |>
      str_remove("\\s+Yes/No/NA\\s*$") |>
      str_remove("\\s+Yes/No\\s*$") |>
      str_squish()
  ) |>
  arrange(match(Item, item_spec$Item)) |>
  select(Item, Type, Definition, Ratings)

if (!identical(codebook$Item, item_spec$Item)) {
  stop("The Codebook sheet does not match the expected CRIME-Q item set and order.")
}

if (anyNA(codebook$Ratings)) {
  stop("Allowed ratings could not be identified for every codebook item.")
}

study_label <- function(study_id) {
  parts <- str_split(study_id, "_", simplify = FALSE)
  vapply(
    parts,
    function(x) {
      if (length(x) < 2L) return(x[1])
      paste0(str_to_title(x[1]), " et al. ", x[2])
    },
    character(1)
  )
}

consensus_matrix <- consensus |>
  arrange(study_id) |>
  mutate(Study = study_label(study_id)) |>
  select(Study, study_id, all_of(item_spec$data_column))

plot_dat <- consensus_matrix |>
  pivot_longer(
    cols = all_of(item_spec$data_column),
    names_to = "data_column",
    values_to = "Rating"
  ) |>
  left_join(item_spec, by = "data_column") |>
  mutate(
    Item = factor(Item, levels = item_spec$Item),
    Study = factor(Study, levels = rev(unique(consensus_matrix$Study))),
    Rating = factor(Rating, levels = rating_levels)
  )

to_long <- function(dat, reviewer_name) {
  dat |>
    select(study_id, all_of(item_spec$data_column)) |>
    pivot_longer(
      cols = all_of(item_spec$data_column),
      names_to = "data_column",
      values_to = reviewer_name
    ) |>
    left_join(item_spec, by = "data_column") |>
    select(study_id, Item, all_of(reviewer_name))
}

agreement_dat <- inner_join(
  to_long(reviewer_1, "Reviewer_1"),
  to_long(consensus, "Reviewer_2"),
  by = c("study_id", "Item")
)

expected_comparisons <- nrow(consensus) * nrow(item_spec)
if (nrow(agreement_dat) != expected_comparisons) {
  stop("The reviewer comparison does not contain the expected study-item pairs.")
}

cohen_kappa <- function(x, y, levels) {
  tab <- table(
    factor(x, levels = levels),
    factor(y, levels = levels)
  )
  n <- sum(tab)
  observed <- sum(diag(tab)) / n
  expected <- sum(rowSums(tab) * colSums(tab)) / n^2
  kappa <- if (isTRUE(all.equal(expected, 1))) NA_real_ else (observed - expected) / (1 - expected)

  tibble(
    comparisons = n,
    observed_agreement = observed,
    expected_agreement = expected,
    kappa = kappa
  )
}

agreement_overall <- cohen_kappa(
  agreement_dat$Reviewer_1,
  agreement_dat$Reviewer_2,
  rating_levels
)

item_disagreement <- agreement_dat |>
  mutate(disagrees = Reviewer_1 != Reviewer_2) |>
  group_by(Item) |>
  summarise(
    Comparisons = n(),
    Disagreements = sum(disagrees),
    `Disagreement rate` = mean(disagrees),
    .groups = "drop"
  ) |>
  mutate(Item = factor(Item, levels = item_spec$Item)) |>
  arrange(Item) |>
  mutate(Item = as.character(Item))

rating_colors <- c(
  "Yes" = "#3A7D44",
  "Partly" = "#D89A24",
  "Unclear" = "#7C8793",
  "No" = "#B24745",
  "NA" = "#477AA8"
)
```

This page reports the adapted CRIME-Q critical-appraisal items used in this review, the final reconciled study-level ratings, and agreement between the two independent appraisals. The assessment covers quality of reporting (QoR), methodological quality (MQ), and risk of bias (RoB), together with three review-specific transparency items.

## CRIME-Q item definitions

The compact table is generated directly from the **Codebook** sheet of `CRIME-Q_Appraisal_Template.xlsx`. It therefore reflects this review's adapted item set rather than a generic CRIME-Q codebook.

```{r}
#| label: crimeq-definitions
#| echo: true
#| results: asis

codebook |>
  kable(
    format = "html",
    col.names = c("Item", "Domain", "Definition", "Allowed ratings"),
    align = c("l", "l", "l", "l"),
    escape = TRUE,
    table.attr = 'class="table table-striped table-hover crimeq-codebook"'
  )
```

## Study-level rating figure

The final workbook contains `r nrow(consensus_matrix)` studies. Each tile represents the reconciled rating for one study and one adapted CRIME-Q item. `Partly` and `Unclear` are displayed separately because they have different meanings: `Partly` indicates partial fulfilment of an item, whereas `Unclear` indicates insufficient information for a definite judgment.

```{r}
#| label: crimeq-heatmap
#| echo: true
#| fig-width: 15
#| fig-height: 8
#| out-width: "100%"
#| fig-cap: "Consensus CRIME-Q ratings for each included study. Green = Yes, gold = Partly, grey = Unclear, red = No, and blue = NA."

crimeq_plot <- ggplot(plot_dat, aes(x = Item, y = Study, fill = Rating)) +
  geom_tile(color = "#FFFDF8", linewidth = 0.65) +
  scale_fill_manual(
    values = rating_colors,
    breaks = rating_levels,
    drop = FALSE,
    name = NULL
  ) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    plot.background = element_rect(fill = "transparent", color = NA),
    panel.background = element_rect(fill = "transparent", color = NA),
    legend.background = element_rect(fill = "transparent", color = NA),
    legend.key = element_rect(fill = "transparent", color = NA),
    panel.grid = element_blank(),
    axis.text.x = element_text(
      angle = 0,
      hjust = 0.5,
      vjust = 0.5,
      face = "bold",
      color = "#242321",
      size = 10
    ),
    axis.text.y = element_text(face = "bold", color = "#242321", size = 10),
    axis.ticks = element_blank(),
    legend.position = "bottom",
    legend.text = element_text(size = 11, color = "#242321"),
    legend.key.width = grid::unit(1.5, "lines"),
    plot.margin = margin(10, 10, 10, 10)
  )

crimeq_plot
```

## Inter-reviewer agreement

Reviewer 1 ratings are read from `CRIME-Q_Appraisal_Template_Anna.xlsx`. Reviewer 2 ratings are represented by `CRIME-Q_Appraisal_Template.xlsx`, because the final workbook retains the second reviewer's original ratings and Reviewer 1 resolved the disagreements by changing 11 entries to match them. The consensus workbook is also the source of the study-level figure above.

The headline coefficient is an unweighted Cohen's kappa pooled across all `r agreement_overall$comparisons` study-item rating pairs. This pooling provides an overall descriptive measure, but the items have different permitted response categories and the observations within studies are not independent; the coefficient should therefore be interpreted alongside percent agreement and the item-level disagreement table.

::: {.agreement-summary}
::: {.agreement-card}
### Overall Cohen's kappa

`r sprintf("%.3f", agreement_overall$kappa)`
:::

::: {.agreement-card}
### Observed agreement

`r scales::percent(agreement_overall$observed_agreement, accuracy = 0.1)`
:::

::: {.agreement-card}
### Disagreements

`r sum(item_disagreement$Disagreements)` of `r agreement_overall$comparisons`
:::
:::

### Disagreement by item

```{r}
#| label: crimeq-item-disagreement
#| echo: true
#| results: asis

item_disagreement |>
  mutate(`Disagreement rate` = scales::percent(`Disagreement rate`, accuracy = 0.1)) |>
  kable(
    format = "html",
    align = c("l", "r", "r", "r"),
    escape = TRUE,
    table.attr = 'class="table table-striped table-hover crimeq-agreement"'
  )
```

```{css}
.crimeq-codebook,
.crimeq-agreement {
  width: 100%;
  font-size: 0.93rem;
}

.crimeq-codebook th,
.crimeq-agreement th {
  background: #eef2f4;
  color: #24313a;
  vertical-align: bottom;
}

.crimeq-codebook td,
.crimeq-agreement td {
  vertical-align: top;
}

.crimeq-codebook th:first-child,
.crimeq-codebook td:first-child,
.crimeq-agreement th:first-child,
.crimeq-agreement td:first-child {
  white-space: nowrap;
  font-weight: 600;
}

.agreement-summary {
  display: grid;
  grid-template-columns: repeat(3, minmax(0, 1fr));
  gap: 1rem;
  margin: 1.25rem 0 1.5rem;
}

.agreement-card {
  border: 1px solid #d9e0e4;
  border-left: 5px solid #477aa8;
  border-radius: 0.35rem;
  background: #f8fafb;
  padding: 1rem 1.1rem;
}

.agreement-card h3 {
  margin-top: 0;
  margin-bottom: 0.45rem;
  font-size: 1rem;
  color: #4d5961;
}

.agreement-card p {
  margin: 0;
  font-size: 1.65rem;
  font-weight: 700;
  color: #24313a;
}

@media (max-width: 800px) {
  .agreement-summary {
    grid-template-columns: 1fr;
  }
}
```
