::: {.callout-warning}
## Superseded analysis file

This exploratory file is retained for provenance but is no longer part of the active workflow. The reviewed implementation is being migrated to `pcne_book/02_effect_size_lnRR.qmd`, which reads the audited output from `pcne_book/01_data_wrangling.qmd`.

Do not use this file to regenerate the analysis dataset.
:::

# Estimating effect sizes (legacy)

```{r}
#| label: setup
#| include: true
pacman::p_load(
               DT,
               dtplyr,
               here, 
               knitr,
               tidyverse,
               patchwork,
               metafor
               )
```

## lnRR Effect Size Calculation Function

This function, `calculate_lnRR()`, is designed to calculate the Log Response Ratio ($\text{lnRR}$) and its variance ($\text{lnRR_var}$) for both continuous and percentage data, accounting for dependent (correlated) and independent comparison structures.

::: callout-note
## Variable definitions

| Variable | Definition |
|------------------------------------|------------------------------------|
| `es_id` | Unique identifier for each effect size |
| `study_id` | Identifier for the study from which the data is drawn |
| `c_n`, `ex_n` | Sample sizes for control and experimental groups |
| `c_mean`, `ex_mean` | Group means for control and experimental groups |
| `c_se`, `ex_se` | Group standard errors for control and experimental groups |
| `c_sd`, `ex_sd` | Group standard deviations for control and experimental groups |
| `data_type` | Outcome type (e.g., time, count, latency, percentage, index) |
| `comparison_structure` | "Independent" (different animals) vs "Dependent" (same animals) |
| `higher_better` | Whether higher values indicate a better outcome (Yes/No) |
:::

```{r}

#| label: calculate_lnRR_final

# =========================
# (1) Robust calculate_lnRR() with dependent-n audit + safer guards
# (2) Safe re-load from CSV without lnRR/lnRR_var becoming character/factor
# =========================

calculate_lnRR <- function(dt) {

  # ---- 1) Guards / cleaning (ONLY where relevant) ----
  dt <- dt %>%
    mutate(
      data_type = trimws(tolower(as.character(data_type))),

      # Guard against log(0) for ALL data types (applies only if mean==0)
      c_mean  = ifelse(!is.na(c_mean)  & c_mean  == 0, 0.005, c_mean),
      ex_mean = ifelse(!is.na(ex_mean) & ex_mean == 0, 0.005, ex_mean),

      # Guard 100% ONLY for percentage outcomes
      c_mean  = ifelse(data_type == "percentage" & !is.na(c_mean)  & c_mean  == 100, 99.5, c_mean),
      ex_mean = ifelse(data_type == "percentage" & !is.na(ex_mean) & ex_mean == 100, 99.5, ex_mean),

      # Symmetric zero-SD guard
      c_sd  = ifelse(!is.na(c_sd)  & c_sd  == 0, 0.01, c_sd),
      ex_sd = ifelse(!is.na(ex_sd) & ex_sd == 0, 0.01, ex_sd)
    )

  # ---- 2) Setup ----
  dt1 <- dt %>%
    mutate(
      lnRR = NA_real_,
      lnRR_var = NA_real_
    )

  asin_trans <- function(p) asin(sqrt(p))

  # ---- 3) Loop ----
  for (i in seq_len(nrow(dt1))) {

    ex_n  <- dt1$ex_n[i]
    c_n   <- dt1$c_n[i]
    ex_m  <- dt1$ex_mean[i]
    c_m   <- dt1$c_mean[i]
    ex_sd <- dt1$ex_sd[i]
    c_sd  <- dt1$c_sd[i]

    data_type <- trimws(tolower(as.character(dt1$data_type[i])))
    Comp      <- trimws(tolower(as.character(dt1$comparison_structure[i])))
    HB        <- trimws(tolower(as.character(dt1$higher_better[i])))

    # ---- A) Continuous data ----
    if (data_type %in% c("time", "count", "latency", "index")) {

      if (Comp == "independent") {

        effect <- metafor::escalc(
          measure = "ROM",
          n1i = ex_n, n2i = c_n,
          m1i = ex_m, m2i = c_m,
          sd1i = ex_sd, sd2i = c_sd,
          var.names = c("lnRR", "lnRR_var")
        )

      } else if (Comp == "dependent") {

        # Audit: dependent design but unequal n
        if (!is.na(ex_n) && !is.na(c_n) && ex_n != c_n) {
          warning(
            "Dependent design but ex_n != c_n for es_id = ", dt1$es_id[i],
            " (ex_n=", ex_n, ", c_n=", c_n, "). Using ni=(ex_n+c_n)/2."
          )
        }

        effect <- metafor::escalc(
          measure = "ROMC",
          ni  = (ex_n + c_n) / 2,
          m1i = ex_m, m2i = c_m,
          sd1i = ex_sd, sd2i = c_sd,
          ri  = 0.5,
          var.names = c("lnRR", "lnRR_var")
        )

      } else {
        warning("Unhandled comparison_structure for es_id = ", dt1$es_id[i], ": ", Comp)
        next
      }

      dt1$lnRR[i]     <- effect$lnRR
      dt1$lnRR_var[i] <- effect$lnRR_var

      # Direction standardization: flip sign only
      if (HB == "yes") dt1$lnRR[i] <- -dt1$lnRR[i]
    }

    # ---- B) Percentage data ----
    else if (data_type == "percentage") {

      # Rescale if values look like 0–100
      if (!is.na(ex_m) && !is.na(c_m) && (ex_m > 1 || c_m > 1)) {
        ex_m  <- ex_m / 100
        c_m   <- c_m  / 100
        ex_sd <- ex_sd / 100
        c_sd  <- c_sd  / 100
      }

      # Clamp proportions away from 0/1
      ex_m <- pmin(pmax(ex_m, 0.005), 0.995)
      c_m  <- pmin(pmax(c_m,  0.005), 0.995)

      # Delta-method approx for SD on arcsin-sqrt scale (as you used)
      ex_sd_t <- sqrt(ex_sd^2 / (4 * ex_m * (1 - ex_m)))
      c_sd_t  <- sqrt(c_sd^2  / (4 * c_m  * (1 - c_m)))

      ex_m_t <- asin_trans(ex_m)
      c_m_t  <- asin_trans(c_m)

      lnRR_calc <- log(ex_m_t / c_m_t)

      v_ex <- (ex_sd_t)^2 * (1 / (ex_m_t^2 * ex_n))
      v_c  <- (c_sd_t )^2 * (1 / (c_m_t^2  * c_n))

      if (Comp == "independent") {
        lnRR_var_calc <- v_ex + v_c
      } else if (Comp == "dependent") {
        lnRR_var_calc <- v_ex + v_c - 2 * 0.5 * sqrt(v_ex) * sqrt(v_c)
      } else {
        warning("Unhandled comparison_structure for es_id = ", dt1$es_id[i], ": ", Comp)
        next
      }

      dt1$lnRR[i]     <- lnRR_calc
      dt1$lnRR_var[i] <- lnRR_var_calc

      # Direction standardization: flip sign only
      if (HB == "yes") dt1$lnRR[i] <- -dt1$lnRR[i]
    }

    # ---- C) Anything else ----
    else {
      warning("Unhandled data_type for es_id = ", dt1$es_id[i], ": ", data_type)
      next
    }
  }

  dt1
}


```

# =========================

::: callout-note
## Minimum-mean rule for small-unit outcomes (seconds, counts)

For outcomes recorded in **seconds** or **counts**, some extracted means were \< 1. This can generate extreme lnRR values and inflated sampling variances because ratio-of-means estimators use (\log(\bar{x}*{Ex}/*\bar{x}{C})) and the sampling variance scales approximately with (1/\bar{x}\^2).

To stabilise computation and reflect plausible measurement resolution, we set a **minimum mean of 1 (in original units)** for these outcomes prior to lnRR calculation.

Standard deviations were **not** floored; we applied only a **zero-SD guard** (replacing SD = 0 with a small positive value) to prevent undefined variances.
:::

::: callout-note
## Converting median and IQR to mean and SD (Wan et al. 2014)

A small subset of studies reported outcomes as the **median** and **interquartile range** ((q_1)–(q_3)) rather than mean ± SD (e.g., Abramova). To harmonize summary statistics for effect-size computation, we approximated the **mean** and **standard deviation** from (n), (q_1), (m), and (q_3) using the quantile-based estimators of **Wan et al. (2014)**.

**Given:** sample size (n), first quartile (q_1), median (m), third quartile (q_3)

\[ \hat{\mu} = \frac{q_1 + m + q_3}{3} \]

\[ \hat{\sigma} = \frac{q_3 - q_1}{2\,\Phi^{-1}\!\left(\frac{0.75n - 0.125}{n + 0.25}\right)} \]

where (\Phi\^{-1}(\cdot)) is the standard normal quantile function (R: `qnorm()`).
:::

```{r}
#| label: wan_helpers
#| echo: false

# Wan et al. (2014) Scenario C3: estimate mean/sd from n, q1, median, q3
wan_c3_mean <- function(q1, med, q3) (q1 + med + q3) / 3

wan_c3_sd <- function(q1, q3, n) {
  p <- (0.75 * n - 0.125) / (n + 0.25)
  (q3 - q1) / (2 * stats::qnorm(p))
}

# Optional: CI half-width -> SE (assumes 95% CI half-width)
ci_halfwidth_to_se <- function(ci_halfwidth) ci_halfwidth / 1.96
```

```{r}
#| label: load_and_clean_data

db <- readr::read_csv(here("Data", "PCN_data_ext_checking_v.3.csv")) %>%
  
  # 0) Make sure numeric fields are numeric
  mutate(across(
    c(c_n, ex_n, c_mean, ex_mean, c_se, ex_se, c_sd, ex_sd,
      c_ci, ex_ci, c_median, c_q1, c_q3, ex_median, ex_q1, ex_q3),
    as.numeric
  )) %>%
  
  # 1) Audit what was missing BEFORE imputation
  mutate(
    miss_c_mean_raw  = is.na(c_mean),
    miss_ex_mean_raw = is.na(ex_mean),
    miss_c_sd_raw    = is.na(c_sd),
    miss_ex_sd_raw   = is.na(ex_sd),
    miss_c_se_raw    = is.na(c_se),
    miss_ex_se_raw   = is.na(ex_se)
  ) %>%
  
  # 2) Wan et al. (2014): median+IQR -> mean/SD where mean/SD are missing
  mutate(
    c_mean = ifelse(
      is.na(c_mean) & !is.na(c_q1) & !is.na(c_median) & !is.na(c_q3),
      wan_c3_mean(c_q1, c_median, c_q3),
      c_mean
    ),
    ex_mean = ifelse(
      is.na(ex_mean) & !is.na(ex_q1) & !is.na(ex_median) & !is.na(ex_q3),
      wan_c3_mean(ex_q1, ex_median, ex_q3),
      ex_mean
    ),
    c_sd = ifelse(
      is.na(c_sd) & !is.na(c_q1) & !is.na(c_q3) & !is.na(c_n) & c_n > 1,
      wan_c3_sd(c_q1, c_q3, c_n),
      c_sd
    ),
    ex_sd = ifelse(
      is.na(ex_sd) & !is.na(ex_q1) & !is.na(ex_q3) & !is.na(ex_n) & ex_n > 1,
      wan_c3_sd(ex_q1, ex_q3, ex_n),
      ex_sd
    )
  ) %>%
  
  # 3) CI -> SE (ONLY if se is missing and you have ci columns)
  # IMPORTANT: assumes c_ci/ex_ci are 95% CI *half-widths*.
  mutate(
    c_se  = ifelse(is.na(c_se)  & !is.na(c_ci),  ci_halfwidth_to_se(c_ci),  c_se),
    ex_se = ifelse(is.na(ex_se) & !is.na(ex_ci), ci_halfwidth_to_se(ex_ci), ex_se)
  ) %>%
  
  # 4) SE -> SD (ONLY where SD is still missing and n exists)
  mutate(
    c_sd  = ifelse(is.na(c_sd)  & !is.na(c_se)  & !is.na(c_n)  & c_n  > 0, c_se  * sqrt(c_n),  c_sd),
    ex_sd = ifelse(is.na(ex_sd) & !is.na(ex_se) & !is.na(ex_n) & ex_n > 0, ex_se * sqrt(ex_n), ex_sd)
  ) %>%
  
  # 4b) SD -> SE (AFTER SD imputation; fills SE where still missing)
  mutate(
    c_se  = ifelse(is.na(c_se)  & !is.na(c_sd)  & !is.na(c_n)  & c_n  > 0, c_sd  / sqrt(c_n),  c_se),
    ex_se = ifelse(is.na(ex_se) & !is.na(ex_sd) & !is.na(ex_n) & ex_n > 0, ex_sd / sqrt(ex_n), ex_se)
  ) %>%
  
  # 5) Minimum-mean rule for seconds/counts (means only; SDs not floored)
  mutate(
    data_units_clean = trimws(tolower(as.character(data_units))),
    mean_floored_to1 = data_units_clean %in% c("s","sec","second","seconds","count","counts") &
      (
        (!is.na(c_mean)  & c_mean  < 1) |
        (!is.na(ex_mean) & ex_mean < 1)
      ),
    c_mean  = ifelse(data_units_clean %in% c("s","sec","second","seconds","count","counts") &
                       !is.na(c_mean)  & c_mean  < 1, 1, c_mean),
    ex_mean = ifelse(data_units_clean %in% c("s","sec","second","seconds","count","counts") &
                       !is.na(ex_mean) & ex_mean < 1, 1, ex_mean)
  ) %>%
  select(-data_units_clean) %>%
  
  # 6) cohort_id
  mutate(
    cohort_id = paste(study_id, ex_id, sep = "_")
  ) %>%
  
  # 7) Post-imputation audit flags
  mutate(
    c_mean_imputed  = miss_c_mean_raw  & !is.na(c_mean),
    ex_mean_imputed = miss_ex_mean_raw & !is.na(ex_mean),
    c_sd_imputed    = miss_c_sd_raw    & !is.na(c_sd),
    ex_sd_imputed   = miss_ex_sd_raw   & !is.na(ex_sd),
    c_se_imputed    = miss_c_se_raw    & !is.na(c_se),
    ex_se_imputed   = miss_ex_se_raw   & !is.na(ex_se)
  ) %>%
  
  # 8) Factors last
  mutate(across(where(is.character), as.factor))


# ---- audits ----
db %>%
  summarise(
    n_rows = n(),
    miss_c_mean = sum(is.na(c_mean)),
    miss_ex_mean = sum(is.na(ex_mean)),
    miss_c_sd = sum(is.na(c_sd)),
    miss_ex_sd = sum(is.na(ex_sd)),
    miss_c_se = sum(is.na(c_se)),
    miss_ex_se = sum(is.na(ex_se)),
    miss_c_n = sum(is.na(c_n)),
    miss_ex_n = sum(is.na(ex_n)),
    n_c_mean_imputed = sum(c_mean_imputed, na.rm = TRUE),
    n_ex_mean_imputed = sum(ex_mean_imputed, na.rm = TRUE),
    n_c_sd_imputed = sum(c_sd_imputed, na.rm = TRUE),
    n_ex_sd_imputed = sum(ex_sd_imputed, na.rm = TRUE),
    n_c_se_imputed = sum(c_se_imputed, na.rm = TRUE),
    n_ex_se_imputed = sum(ex_se_imputed, na.rm = TRUE)
  )
```




# =========================

## Calculate lnRR Effect Sizes

Now, we apply the `calculate_lnRR()` function to the cleaned dataset to compute the effect sizes and their variances.

The resulting dataset, `db_effect_sizes`, is then saved as a CSV file for further analysis.

```{r}
#| label: calculate_effect_sizes
#| eval: true
db_effect_sizes <- calculate_lnRR(db)

db_effect_sizes %>%
  summarise(
    n_total = n(),
    n_lnRR_missing = sum(is.na(lnRR)),
    n_var_missing  = sum(is.na(lnRR_var))
  )

db_effect_sizes %>%
  count(data_type, comparison_structure, higher_better)

db_effect_sizes %>%
  filter(is.na(lnRR) | is.na(lnRR_var)) %>%
  dplyr::select(
    es_id, study_id, data_type, comparison_structure, higher_better,
    c_n, ex_n, c_mean, ex_mean, c_sd, ex_sd
  ) %>%
  head(20)

readr::write_csv(
  db_effect_sizes,
  file = here("Data", "db_lnRR.csv"),
  na = ""
)

```

# =========================

## Calculate lnRR Effect Sizes

We apply the `calculate_lnRR()` function to the cleaned dataset to generate effect sizes and sampling variances. The resulting dataset is saved for reproducibility and downstream modeling.


```{r}
#| label: calculate_effect_sizes
#| eval: false
db_effect_sizes<-calculate_lnRR(db)
readr::write_csv(
  db_effect_sizes, 
  file = here("Data", "db_lnRR.csv"),
  na = "" 
)

```

::: panel-tabset
## Effect Size Table

```{r}
#| label: effect_size_table
#| echo: false

db_effect_sizes <- readr::read_csv(here("Data","db_lnRR.csv")) %>%
mutate(across(c(c_n, ex_n, c_mean, ex_mean, c_se, ex_se, c_sd, ex_sd), as.numeric))%>%
  mutate(across(where(is.character), as.factor))

# Define all columns to display (IDs first, then numerical data)
display_cols <- c("es_id", "study_id",  
                  "c_n", "c_mean", "c_se", "c_sd", 
                  "ex_n", "ex_mean", "ex_se", "ex_sd", 
                  "lnRR", "lnRR_var", "cohort_id")

# Define columns that require numerical rounding (excluding the ID columns)
round_cols <- c("c_n", "c_mean", "c_se", "c_sd", 
                "ex_n", "ex_mean", "ex_se", "ex_sd", 
                "lnRR", "lnRR_var")

DT::datatable(
  db_effect_sizes %>% 
    # Select only the desired columns for display
    dplyr::select(all_of(display_cols)),
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv')
  ),
  extensions = 'Buttons',
  rownames = FALSE
) %>%
  # Apply rounding ONLY to the numerical statistical columns
  formatRound(columns = round_cols, digits = 3) %>%
  # Apply general styling to all displayed columns
  formatStyle(columns = display_cols, `text-align` = 'center')
```

## Effect Size Distributions

The histograms below show the distribution of calculated lnRR values and their sampling variances. These plots allow visual inspection of:

- Directional bias (skew toward negative values)
- Extreme outliers
- Variance heterogeneity across studies


```{r}
#| label: lnRR_histogram

hist(db_effect_sizes$lnRR,
     xlim=c(-9,6.5),
     breaks = seq(-9,6.5,0.5),
     main = "Distribution of lnRR",
     xlab = "lnRR")

hist(db_effect_sizes$lnRR_var,
     xlim=c(0,13),
     breaks = seq(0,13,0.1),
     main = "Distribution of lnRR Variance",
     xlab = "Sampling Variance")
```
:::


The full dataset of computed effect sizes is available in the project repository to facilitate independent replication of all meta-analytic results.



# =========================

# Diagnostics: lnRR & variance plots 

# =========================

```{r}


# 0) (Optional but useful) Create SE from variance for standard diagnostics
db_effect_sizes <- db_effect_sizes %>%
  mutate(
    SE_lnRR = sqrt(lnRR_var)  # standard error of lnRR
  )

# 1) Scatter: lnRR vs variance (raw scale)
# Purpose: check if extreme effect sizes coincide with huge uncertainty
plot(
  x = db_effect_sizes$lnRR,
  y = db_effect_sizes$lnRR_var,
  xlab = "lnRR (log response ratio)",
  ylab = "Variance of lnRR",
  pch  = 16
)
abline(h = 0, lty = 2)

# 2) Scatter: lnRR vs variance (log y-scale)  <-- recommended given your right-skewed variance histogram
# Purpose: makes the low-variance cloud visible when a few points have very large variances
plot(
  x = db_effect_sizes$lnRR,
  y = db_effect_sizes$lnRR_var,
  xlab = "lnRR (log response ratio)",
  ylab = "Variance of lnRR (log scale)",
  pch  = 16,
  log  = "y"
)
abline(h = min(db_effect_sizes$lnRR_var[db_effect_sizes$lnRR_var > 0], na.rm = TRUE), lty = 2)

# 3) Scatter: lnRR vs SE (often easier to interpret than variance)
# Purpose: identifies imprecise effects (high SE) and helps spot influential points
plot(
  x = db_effect_sizes$lnRR,
  y = db_effect_sizes$SE_lnRR,
  xlab = "lnRR (log response ratio)",
  ylab = "SE of lnRR",
  pch  = 16
)
abline(h = 0, lty = 2)

# 4) Optional: label the highest-variance points (top N) on the log-variance plot
# Purpose: quickly identify which es_ids are driving the far-right tail
N <- 10
top_idx <- order(db_effect_sizes$lnRR_var, decreasing = TRUE)[seq_len(min(N, nrow(db_effect_sizes)))]
plot(
  x = db_effect_sizes$lnRR,
  y = db_effect_sizes$lnRR_var,
  xlab = "lnRR (log response ratio)",
  ylab = "Variance of lnRR (log scale)",
  pch  = 16,
  log  = "y"
)
text(
  x = db_effect_sizes$lnRR[top_idx],
  y = db_effect_sizes$lnRR_var[top_idx],
  labels = db_effect_sizes$es_id[top_idx],
  pos = 3, cex = 0.7
)

# 5) Optional: print a quick table of the worst offenders (highest variance)
# Purpose: inspect whether huge variances come from tiny n, huge SD, near-zero means, % transforms, etc.
db_effect_sizes %>%
  arrange(desc(lnRR_var)) %>%
  dplyr::select(
    es_id, study_id, data_type, comparison_structure, higher_better,
    lnRR, lnRR_var, SE_lnRR,
    c_n, ex_n, c_mean, ex_mean, c_sd, ex_sd
  ) %>%
  head(15)
```

# =========================================

# SANITY CHECKS + MODERATOR OVERLAP TABLES

# =========================================

```{r}

dat <- db_effect_sizes

# ---- 0) Define moderator sets (edit if you add/remove columns) ----
mods_exposure_cat <- c("gest_stage_ex", "exposure_type", "noise_type", "control_conditions")
mods_exposure_num <- c("exposure_span_d", "exposure_session_duration_h", "loudness_dB", "frequency_Hz", "control_loudness_dB")

mods_bio_cat      <- c("sp_latin", "sp_common", "strain", "offspring_sex")
mods_bio_num      <- c("offspring_age_d")

mods_outcome_cat  <- c("experimental_procedures", "assay_type", "measurement_variable")

# Keep only moderators that actually exist in the dataset (prevents errors)
mods_exposure_cat <- intersect(mods_exposure_cat, names(dat))
mods_exposure_num <- intersect(mods_exposure_num, names(dat))
mods_bio_cat      <- intersect(mods_bio_cat, names(dat))
mods_bio_num      <- intersect(mods_bio_num, names(dat))
mods_outcome_cat  <- intersect(mods_outcome_cat, names(dat))

mods_cat <- c(mods_exposure_cat, mods_bio_cat, mods_outcome_cat)
mods_num <- c(mods_exposure_num, mods_bio_num)

# ---- 1) Type hygiene (important if some fields became factor/character) ----
as_num_safe <- function(x) {
  if (is.factor(x)) x <- as.character(x)
  suppressWarnings(as.numeric(x))
}

dat <- dat %>%
  mutate(across(all_of(mods_cat), ~ {
    # normalize categorical
    x <- .
    if (is.factor(x)) x <- as.character(x)
    x <- trimws(tolower(x))
    na_if(x, "")
  })) %>%
  mutate(across(all_of(mods_num), as_num_safe))

# ---- 2) Coverage table: missingness + spread ----
coverage_table <- function(d, v) {
  x <- d[[v]]
  tibble(
    moderator   = v,
    n_total     = nrow(d),
    n_missing   = sum(is.na(x)),
    n_present   = sum(!is.na(x)),
    n_unique    = n_distinct(x, na.rm = TRUE),
    # For categorical: number of levels observed (same as unique); for numeric still useful
    type        = ifelse(is.numeric(x), "numeric", "categorical")
  )
}

mod_coverage <- map_dfr(c(mods_cat, mods_num), ~coverage_table(dat, .x)) %>%
  arrange(desc(n_missing))

mod_coverage

# ---- 3) Top levels table for categorical moderators (spot weird labels / rare levels) ----
top_levels <- function(d, v, top_n = 12) {
  d %>%
    count(.data[[v]], sort = TRUE, name = "n") %>%
    mutate(
      moderator = v,
      prop = n / sum(n)
    ) %>%
    slice_head(n = top_n) %>%
    rename(level = 1) %>%
    select(moderator, level, n, prop)
}

mod_top_levels <- map_dfr(mods_cat, ~top_levels(dat, .x, top_n = 12))
mod_top_levels

# ---- 4) Numeric spread summaries (outliers, clumping, impossible values) ----
num_spread <- function(d, v) {
  x <- d[[v]]
  tibble(
    moderator = v,
    n_present = sum(!is.na(x)),
    min = suppressWarnings(min(x, na.rm = TRUE)),
    q25 = suppressWarnings(quantile(x, 0.25, na.rm = TRUE)),
    median = suppressWarnings(median(x, na.rm = TRUE)),
    mean = suppressWarnings(mean(x, na.rm = TRUE)),
    q75 = suppressWarnings(quantile(x, 0.75, na.rm = TRUE)),
    max = suppressWarnings(max(x, na.rm = TRUE))
  )
}

mod_num_spread <- map_dfr(mods_num, ~num_spread(dat, .x))
mod_num_spread

# ---- 5) Study / cohort clustering (non-independence sanity check) ----
# If you have cohort_id already, this will work; otherwise create it like you did:
if (!"cohort_id" %in% names(dat) && all(c("study_id","ex_id") %in% names(dat))) {
  dat <- dat %>% mutate(cohort_id = paste(study_id, ex_id, sep = "_"))
}

cluster_by_study <- dat %>% count(study_id, sort = TRUE, name = "n_ES")
cluster_by_cohort <- if ("cohort_id" %in% names(dat)) dat %>% count(cohort_id, sort = TRUE, name = "n_ES") else NULL

cluster_by_study
cluster_by_cohort

# ---- 6) “Moderator profile” clusters (are the same moderator combos repeating?) ----
# Use a subset of key moderators to define a profile (edit as needed)
profile_vars <- intersect(
  c("sp_latin","strain","gest_stage_ex","exposure_type","noise_type","offspring_sex","assay_type"),
  names(dat)
)

profiles <- dat %>%
  mutate(across(all_of(profile_vars), ~replace_na(as.character(.), "NA"))) %>%
  unite("profile", all_of(profile_vars), sep = " | ", remove = FALSE)

profile_counts <- profiles %>% count(profile, sort = TRUE, name = "n_ES")
profile_counts %>% slice_head(n = 30)

# ---- 7) Pairwise overlap tables (exposure vs bio vs outcome) ----
# Helper: crosstab as a count table (wide)
crosstab_wide <- function(d, a, b, top_b = NULL) {
  dd <- d %>% filter(!is.na(.data[[a]]), !is.na(.data[[b]]))

  # Optional: restrict b to top levels if it’s huge (e.g., measurement_variable)
  if (!is.null(top_b)) {
    keep_b <- dd %>% count(.data[[b]], sort = TRUE) %>% slice_head(n = top_b) %>% pull(1)
    dd <- dd %>% filter(.data[[b]] %in% keep_b)
  }

  dd %>%
    count(.data[[a]], .data[[b]], name = "n") %>%
    pivot_wider(names_from = all_of(b), values_from = n, values_fill = 0) %>%
    arrange(desc(rowSums(across(-all_of(a)))))
}

# Selected high-value overlap checks (edit/add)
# Exposure vs outcome
overlap_exposure_assay <- crosstab_wide(dat, "noise_type", "assay_type")
overlap_exposure_measure <- crosstab_wide(dat, "noise_type", "measurement_variable", top_b = 15)

# Bio vs exposure (confounding checks)
overlap_strain_noise <- crosstab_wide(dat, "strain", "noise_type")
overlap_species_noise <- crosstab_wide(dat, "sp_latin", "noise_type")

# Outcome structure checks
overlap_assay_measure <- crosstab_wide(dat, "assay_type", "measurement_variable", top_b = 20)

overlap_exposure_assay
overlap_exposure_measure
overlap_strain_noise
overlap_species_noise
overlap_assay_measure

# ---- 8) “Exclusivity” summary: are levels of A only observed with 1 level of B? ----
# This flags overlap/collinearity problems (e.g., one strain only ever gets one noise type)
exclusivity_summary <- function(d, a, b) {
  dd <- d %>% filter(!is.na(.data[[a]]), !is.na(.data[[b]]))
  by_a <- dd %>% distinct(.data[[a]], .data[[b]]) %>% count(.data[[a]], name = "n_levels_of_B")
  tibble(
    A = a,
    B = b,
    prop_A_exclusive_to_one_B = mean(by_a$n_levels_of_B == 1),
    median_B_per_A = median(by_a$n_levels_of_B),
    max_B_per_A = max(by_a$n_levels_of_B)
  )
}

pairs_to_check <- tribble(
  ~a, ~b,
  "strain", "noise_type",
  "sp_latin", "noise_type",
  "noise_type", "assay_type",
  "exposure_type", "noise_type",
  "gest_stage_ex", "exposure_type",
  "offspring_sex", "assay_type"
) %>%
  filter(a %in% names(dat), b %in% names(dat))

exclusivity <- pmap_dfr(pairs_to_check, ~exclusivity_summary(dat, ..1, ..2))
exclusivity

```

```{r}
# =========================
# Simple moderator overlap tables
# =========================



```



:::
