---
title: "Data preparation"
---

The original extraction database is retained unchanged. Two derived databases are created for the analysis.

```{r}
#| label: setup

pacman::p_load(
  dplyr,
  knitr,
  readr,
  stringr,
  tibble
)

project_root <- normalizePath("..", mustWork = TRUE)
data_dir <- file.path(project_root, "Data")
derived_dir <- file.path(data_dir, "derived")
dir.create(derived_dir, recursive = TRUE, showWarnings = FALSE)

original_path <- file.path(data_dir, "PCN_data_ext_checking_v.3.csv")
statistics_path <- file.path(
  derived_dir,
  "PCN_data_ext_checking_v.3_with_derived_statistics.csv"
)
analysis_path <- file.path(derived_dir, "PCN_analysis_ready.csv")
```

# Source data

The source file is `Data/PCN_data_ext_checking_v.3.csv`. Its checksum is recorded before the derived files are created.

```{r}
#| label: read-source-data

original_checksum <- unname(tools::md5sum(original_path))

original <- readr::read_csv(
  original_path,
  na = c("", "NA"),
  name_repair = "unique_quiet",
  show_col_types = FALSE
)

required_columns <- c(
  "study_id", "ex_id", "con_id", "out_id", "es_id",
  "comparison_structure", "higher_better",
  "c_mean", "c_sd", "c_n", "c_se", "c_median", "c_q1", "c_q3",
  "ex_mean", "ex_sd", "ex_n", "ex_se", "ex_median", "ex_q1", "ex_q3"
)

if (length(setdiff(required_columns, names(original))) > 0) {
  stop("Required columns are missing from the original extraction database.")
}

empty_unnamed_columns <- names(original)[
  str_detect(names(original), "^\\.\\.\\.[0-9]+$") &
    vapply(original, function(x) all(is.na(x)), logical(1))
]

source_for_derivation <- original |>
  select(-any_of(empty_unnamed_columns))

tibble(
  file = "Data/PCN_data_ext_checking_v.3.csv",
  rows = nrow(original),
  columns = ncol(original),
  studies = n_distinct(original$study_id),
  effect_sizes = n_distinct(original$es_id),
  checksum = original_checksum
) |>
  knitr::kable(caption = "Original extraction database.")
```

# Derivation of missing summary statistics

## Calculation methods

- A missing mean is estimated from Q1, the median, Q3, and sample size with the Luo et al. method [@luo2018].
- A missing SD is estimated from Q1, Q3, and sample size with the Wan et al. method [@wan2014].
- If an SD is missing but an SE is available, `SD = SE × sqrt(n)`.
- If an SE is missing, `SE = SD / sqrt(n)`.
- Existing extracted values are retained unchanged.

$$
\widehat{M} =
\left(0.7 + \frac{0.39}{n}\right)\frac{Q_1 + Q_3}{2}
+ \left(0.3 - \frac{0.39}{n}\right)M_d
$$

$$
\widehat{SD} =
\frac{Q_3-Q_1}
{2\Phi^{-1}\left(\frac{0.75n-0.125}{n+0.25}\right)}
$$

```{r}
#| label: summary-statistic-functions

estimate_mean_luo <- function(q1, median, q3, n) {
  weight <- 0.7 + 0.39 / n
  weight * ((q1 + q3) / 2) + (1 - weight) * median
}

estimate_sd_wan <- function(q1, q3, n) {
  denominator <- 2 * stats::qnorm((0.75 * n - 0.125) / (n + 0.25))
  (q3 - q1) / denominator
}

complete_arm_statistics <- function(data, arm) {
  n <- data[[paste0(arm, "_n")]]
  mean_source <- data[[paste0(arm, "_mean")]]
  sd_source <- data[[paste0(arm, "_sd")]]
  se_source <- data[[paste0(arm, "_se")]]
  median <- data[[paste0(arm, "_median")]]
  q1 <- data[[paste0(arm, "_q1")]]
  q3 <- data[[paste0(arm, "_q3")]]

  mean_from_iqr <- is.na(mean_source) & complete.cases(q1, median, q3, n)
  sd_from_iqr <- is.na(sd_source) & complete.cases(q1, q3, n)
  sd_from_se <- is.na(sd_source) & !sd_from_iqr & !is.na(se_source) & !is.na(n)

  if (any(mean_from_iqr & !(q1 <= median & median <= q3))) {
    stop("Invalid Q1 <= median <= Q3 ordering in the ", arm, " arm.")
  }

  mean_analysis <- mean_source
  mean_analysis[mean_from_iqr] <- estimate_mean_luo(
    q1[mean_from_iqr], median[mean_from_iqr], q3[mean_from_iqr], n[mean_from_iqr]
  )

  sd_analysis <- sd_source
  sd_analysis[sd_from_iqr] <- estimate_sd_wan(
    q1[sd_from_iqr], q3[sd_from_iqr], n[sd_from_iqr]
  )
  sd_analysis[sd_from_se] <- se_source[sd_from_se] * sqrt(n[sd_from_se])

  se_analysis <- se_source
  se_from_sd <- is.na(se_analysis) & !is.na(sd_analysis) & !is.na(n)
  se_analysis[se_from_sd] <- sd_analysis[se_from_sd] / sqrt(n[se_from_sd])

  data[[paste0(arm, "_mean_analysis")]] <- mean_analysis
  data[[paste0(arm, "_mean_method")]] <- if_else(
    mean_from_iqr,
    "Luo et al. (2018): Q1, median, Q3, and n",
    "Extracted mean used unchanged"
  )
  data[[paste0(arm, "_sd_analysis")]] <- sd_analysis
  data[[paste0(arm, "_sd_method")]] <- case_when(
    sd_from_iqr ~ "Wan et al. (2014): Q1, Q3, and n",
    sd_from_se ~ "SD = SE * sqrt(n)",
    TRUE ~ "Extracted SD used unchanged"
  )
  data[[paste0(arm, "_se_analysis")]] <- se_analysis
  data[[paste0(arm, "_se_method")]] <- if_else(
    se_from_sd,
    "SE = SD / sqrt(n)",
    "Extracted SE used unchanged"
  )

  if (anyNA(mean_analysis) || anyNA(sd_analysis) || anyNA(se_analysis)) {
    stop("Unresolved summary statistic in the ", arm, " arm.")
  }

  data
}
```

```{r}
#| label: calculate-summary-statistics

with_statistics <- source_for_derivation |>
  complete_arm_statistics("c") |>
  complete_arm_statistics("ex")

method_counts <- bind_rows(
  with_statistics |> count(arm = "Control", statistic = "Mean", method = c_mean_method, name = "rows"),
  with_statistics |> count(arm = "Control", statistic = "SD", method = c_sd_method, name = "rows"),
  with_statistics |> count(arm = "Control", statistic = "SE", method = c_se_method, name = "rows"),
  with_statistics |> count(arm = "Exposed", statistic = "Mean", method = ex_mean_method, name = "rows"),
  with_statistics |> count(arm = "Exposed", statistic = "SD", method = ex_sd_method, name = "rows"),
  with_statistics |> count(arm = "Exposed", statistic = "SE", method = ex_se_method, name = "rows")
)

method_counts |>
  arrange(arm, statistic, desc(rows)) |>
  knitr::kable(caption = "Sources of the analysis statistics.")
```

# Analysis-ready data

The analysis-ready database contains one mean, SD, SE, and sample size for each arm. The identifiers `ex_id` and `con_id` are retained for the sampling dependency structure, and `comparison_structure` is retained for the lnRR calculation.

```{r}
#| label: create-analysis-ready-data

analysis_ready <- with_statistics |>
  mutate(
    across(where(is.character), str_squish),
    shared_control = str_to_lower(shared_control),
    comparison_structure = str_to_lower(comparison_structure),
    higher_better = str_to_lower(higher_better),
    gest_stage_ex = str_to_lower(gest_stage_ex),
    exposure_type = str_to_lower(exposure_type),
    exposure_type = recode(exposure_type, "intermitent" = "intermittent"),
    noise_type = str_to_lower(noise_type),
    noise_type_2 = str_to_lower(noise_type_2),
    control_conditions = str_to_lower(control_conditions),
    offspring_sex = str_to_lower(offspring_sex),
    outcome_type = str_to_lower(outcome_type),
    data_type = str_to_lower(data_type),
    measurement_timing = str_to_lower(measurement_timing),
    strain = recode(strain, "Wistar albino rat" = "Wistar rat")
  ) |>
  transmute(
    study_id, title, doi, publish_in,
    ex_id, con_id, out_id, es_id, shared_control,
    sp_latin, sp_common, strain, dam_age, gest_stage_ex,
    exposure_span_d, exposure_type, exposure_session_duration_h,
    noise_type, noise_type_2, loudness_dB, frequency_Hz,
    control_conditions, offspring_sex, offspring_age_d,
    experimental_procedures, outcome_type, assay_type,
    measurement_variable, data_type, data_units, measurement_timing,
    data_source, data_file, c_a_id, ex_a_id, comparison,
    comparison_structure, higher_better,
    c_mean = c_mean_analysis,
    c_sd = c_sd_analysis,
    c_n,
    c_se = c_se_analysis,
    c_mean_method, c_sd_method, c_se_method,
    ex_mean = ex_mean_analysis,
    ex_sd = ex_sd_analysis,
    ex_n,
    ex_se = ex_se_analysis,
    ex_mean_method, ex_sd_method, ex_se_method
  )
```

## Data checks

```{r}
#| label: validate-analysis-data

validation <- tibble(
  check = c(
    "One row per es_id",
    "Study identifiers present",
    "Positive sample sizes",
    "Complete means",
    "Complete non-negative SDs",
    "Complete non-negative SEs",
    "Valid higher_better values",
    "Valid comparison_structure values"
  ),
  failures = c(
    sum(is.na(analysis_ready$es_id)) + sum(duplicated(analysis_ready$es_id)),
    sum(is.na(analysis_ready$study_id) | analysis_ready$study_id == ""),
    sum(is.na(analysis_ready$c_n) | is.na(analysis_ready$ex_n) |
          analysis_ready$c_n <= 0 | analysis_ready$ex_n <= 0),
    sum(is.na(analysis_ready$c_mean) | is.na(analysis_ready$ex_mean)),
    sum(is.na(analysis_ready$c_sd) | is.na(analysis_ready$ex_sd) |
          analysis_ready$c_sd < 0 | analysis_ready$ex_sd < 0),
    sum(is.na(analysis_ready$c_se) | is.na(analysis_ready$ex_se) |
          analysis_ready$c_se < 0 | analysis_ready$ex_se < 0),
    sum(!analysis_ready$higher_better %in% c("yes", "no")),
    sum(!analysis_ready$comparison_structure %in% c("independent", "dependent"))
  )
) |>
  mutate(status = if_else(failures == 0, "Pass", "Fail"))

validation |>
  knitr::kable(caption = "Analysis-ready data checks.")

if (any(validation$status == "Fail")) {
  stop("The analysis-ready database failed validation.")
}
```

# Data files

The completed extraction database and the analysis-ready database are saved as comma-separated files. The original extraction file is not overwritten.

```{r}
#| label: export-data-files

readr::write_csv(with_statistics, statistics_path, na = "")
readr::write_csv(analysis_ready, analysis_path, na = "")

if (!identical(original_checksum, unname(tools::md5sum(original_path)))) {
  stop("The original extraction file changed during data preparation.")
}

tibble(
  database = c(
    "Original extraction plus calculated statistics",
    "Analysis-ready data"
  ),
  file = c(
    "Data/derived/PCN_data_ext_checking_v.3_with_derived_statistics.csv",
    "Data/derived/PCN_analysis_ready.csv"
  ),
  rows = c(nrow(with_statistics), nrow(analysis_ready)),
  columns = c(ncol(with_statistics), ncol(analysis_ready))
) |>
  knitr::kable(caption = "Derived data files.")
```

::: callout-note
## Session information

```{r}
#| label: session-information

sessionInfo()
```
:::
