---
title: "Data wrangling"
---

This chapter reads the original extraction file, calculates only the missing summary statistics, and creates the two derived databases used by the analysis. The original CSV is not changed.

```{r}
#| label: setup
#| message: false
#| warning: false

pacman::p_load(
  dplyr,
  knitr,
  readr,
  stringr,
  tibble
)

project_root <- normalizePath("..", mustWork = TRUE)
data_dir <- file.path(project_root, "Data")
derived_dir <- file.path(data_dir, "derived")
dir.create(derived_dir, recursive = TRUE, showWarnings = FALSE)

original_path <- file.path(data_dir, "PCN_data_ext_checking_v.3.csv")
statistics_path <- file.path(
  derived_dir,
  "PCN_data_ext_checking_v.3_with_derived_statistics.csv"
)
analysis_path <- file.path(derived_dir, "PCN_analysis_ready.csv")
```

# Original extraction database

```{r}
#| label: read-original-data

original <- readr::read_csv(
  original_path,
  na = c("", "NA"),
  name_repair = "unique_quiet",
  show_col_types = FALSE
)

# Remove the unnamed source column only when it is completely empty.
empty_unnamed <- names(original)[
  stringr::str_detect(names(original), "^\\.\\.\\.[0-9]+$") &
    vapply(original, function(x) all(is.na(x)), logical(1))
]

original <- original |>
  select(-any_of(empty_unnamed))

tibble(
  rows = nrow(original),
  columns = ncol(original),
  effect_sizes = n_distinct(original$es_id)
) |>
  knitr::kable()
```

# Missing summary statistics

The same formulas are applied separately to the control and exposed arms.

- If a mean is missing and the median, Q1, and Q3 are available, the mean is estimated with the Luo et al. method [@luo2018].
- If an SD is missing and Q1 and Q3 are available, the SD is estimated with the Wan et al. method [@wan2014].
- If an SD is missing and an SE is available, `SD = SE * sqrt(n)`.
- If an SE is missing, `SE = SD / sqrt(n)`.
- Reported or extracted values are otherwise kept unchanged.

$$
\widehat{M} = \left(0.7 + \frac{0.39}{n}\right)\frac{Q_1 + Q_3}{2}
+ \left(0.3 - \frac{0.39}{n}\right)M_d
$$

$$
\widehat{SD} =
\frac{Q_3-Q_1}
{2\Phi^{-1}\left(\frac{0.75n-0.125}{n+0.25}\right)}
$$

```{r}
#| label: summary-statistic-functions

luo_iqr_mean <- function(q1, median, q3, n) {
  weight <- 0.7 + 0.39 / n
  weight * ((q1 + q3) / 2) + (1 - weight) * median
}

wan_iqr_sd <- function(q1, q3, n) {
  denominator <- 2 * stats::qnorm((0.75 * n - 0.125) / (n + 0.25))
  (q3 - q1) / denominator
}

derive_arm_statistics <- function(data, arm) {
  n <- data[[paste0(arm, "_n")]]
  mean_reported <- data[[paste0(arm, "_mean")]]
  sd_reported <- data[[paste0(arm, "_sd")]]
  se_reported <- data[[paste0(arm, "_se")]]
  median <- data[[paste0(arm, "_median")]]
  q1 <- data[[paste0(arm, "_q1")]]
  q3 <- data[[paste0(arm, "_q3")]]

  use_iqr_mean <- is.na(mean_reported) & complete.cases(q1, median, q3, n)
  use_iqr_sd <- is.na(sd_reported) & complete.cases(q1, q3, n)
  use_se_for_sd <- is.na(sd_reported) & !is.na(se_reported) & !is.na(n)

  if (any(use_iqr_mean & !(q1 <= median & median <= q3))) {
    stop("Invalid Q1 <= median <= Q3 ordering in the ", arm, " arm.")
  }

  mean_analysis <- mean_reported
  mean_analysis[use_iqr_mean] <- luo_iqr_mean(
    q1[use_iqr_mean], median[use_iqr_mean], q3[use_iqr_mean], n[use_iqr_mean]
  )

  sd_analysis <- sd_reported
  sd_analysis[use_iqr_sd] <- wan_iqr_sd(
    q1[use_iqr_sd], q3[use_iqr_sd], n[use_iqr_sd]
  )
  sd_analysis[use_se_for_sd] <- se_reported[use_se_for_sd] * sqrt(n[use_se_for_sd])

  se_analysis <- se_reported
  use_sd_for_se <- is.na(se_analysis) & !is.na(sd_analysis) & !is.na(n)
  se_analysis[use_sd_for_se] <- sd_analysis[use_sd_for_se] / sqrt(n[use_sd_for_se])

  mean_note <- ifelse(
    use_iqr_mean,
    "Luo et al. (2018): mean estimated from Q1, median, Q3, and n",
    "Existing mean in original extraction used unchanged"
  )

  sd_note <- case_when(
    use_iqr_sd ~ "Wan et al. (2014): SD estimated from Q1, Q3, and n",
    use_se_for_sd ~ "SD = SE * sqrt(n)",
    TRUE ~ "Existing SD in original extraction used unchanged"
  )

  se_note <- ifelse(
    use_sd_for_se,
    "SE = SD / sqrt(n)",
    "Existing SE in original extraction used unchanged"
  )

  if (anyNA(mean_analysis) || anyNA(sd_analysis) || anyNA(se_analysis)) {
    stop("Unresolved summary statistic in the ", arm, " arm.")
  }

  data[[paste0(arm, "_mean_analysis")]] <- mean_analysis
  data[[paste0(arm, "_mean_note")]] <- mean_note
  data[[paste0(arm, "_sd_analysis")]] <- sd_analysis
  data[[paste0(arm, "_sd_note")]] <- sd_note
  data[[paste0(arm, "_se_analysis")]] <- se_analysis
  data[[paste0(arm, "_se_note")]] <- se_note
  data
}
```

```{r}
#| label: calculate-missing-statistics

with_statistics <- original |>
  derive_arm_statistics("c") |>
  derive_arm_statistics("ex")

bind_rows(
  with_statistics |>
    count(
      arm = "control",
      mean_method = c_mean_note,
      sd_method = c_sd_note,
      se_method = c_se_note,
      name = "rows"
    ),
  with_statistics |>
    count(
      arm = "exposed",
      mean_method = ex_mean_note,
      sd_method = ex_sd_note,
      se_method = ex_se_note,
      name = "rows"
    )
) |>
  knitr::kable(caption = "Summary-statistic sources used in the derived database.")
```

# Analysis-ready database

The analysis-ready database contains the identifiers, planned moderators, source fields, and one mean, SD, SE, and sample size for each arm. `comparison_structure` is retained because Chapter 2 uses it to select the independent- or dependent-groups effect-size formula. `cohort_id` follows the same construction used in the reference workflow: `study_id` plus `ex_id`.

```{r}
#| label: create-analysis-ready-data

analysis_ready <- with_statistics |>
  mutate(
    data_cleaning_note = if_else(
      study_id == "`" & data_file == "whitlow_1978_thesis_fig2",
      "study_id repaired from ` to whitlow_1978_thesis using the source file and title",
      NA_character_
    ),
    study_id = if_else(
      study_id == "`" & data_file == "whitlow_1978_thesis_fig2",
      "whitlow_1978_thesis",
      study_id
    ),
    across(where(is.character), stringr::str_squish),
    shared_control = str_to_lower(shared_control),
    comparison_structure = str_to_lower(comparison_structure),
    higher_better = str_to_lower(higher_better),
    gest_stage_ex = str_to_lower(gest_stage_ex),
    exposure_type = str_to_lower(exposure_type),
    exposure_type = recode(exposure_type, "intermitent" = "intermittent"),
    noise_type = str_to_lower(noise_type),
    noise_type_2 = str_to_lower(noise_type_2),
    control_conditions = str_to_lower(control_conditions),
    offspring_sex = str_to_lower(offspring_sex),
    outcome_type = str_to_lower(outcome_type),
    data_type = str_to_lower(data_type),
    measurement_timing = str_to_lower(measurement_timing),
    strain = recode(strain, "Wistar albino rat" = "Wistar rat"),
    cohort_id = paste(study_id, ex_id, sep = "_")
  ) |>
  transmute(
    study_id, title, doi, publish_in,
    ex_id, con_id, out_id, es_id, cohort_id, shared_control,
    sp_latin, sp_common, strain, dam_age, gest_stage_ex,
    exposure_span_d, exposure_type, exposure_session_duration_h,
    noise_type, noise_type_2, loudness_dB, frequency_Hz,
    control_conditions, offspring_sex, offspring_age_d,
    experimental_procedures, outcome_type, assay_type,
    measurement_variable, data_type, data_units, measurement_timing,
    data_source, data_file, c_a_id, ex_a_id, comparison,
    comparison_structure, higher_better,
    c_mean = c_mean_analysis,
    c_sd = c_sd_analysis,
    c_n,
    c_se = c_se_analysis,
    c_mean_note, c_sd_note, c_se_note,
    ex_mean = ex_mean_analysis,
    ex_sd = ex_sd_analysis,
    ex_n,
    ex_se = ex_se_analysis,
    ex_mean_note, ex_sd_note, ex_se_note,
    data_cleaning_note
  )
```

# Validation and export

```{r}
#| label: validate-and-export

validation <- tibble(
  check = c(
    "one row per es_id",
    "study_id present",
    "sample sizes positive",
    "means complete",
    "SDs complete and non-negative",
    "SEs complete and non-negative",
    "higher_better is yes or no",
    "comparison_structure is independent or dependent"
  ),
  failures = c(
    sum(is.na(analysis_ready$es_id)) + sum(duplicated(analysis_ready$es_id)),
    sum(is.na(analysis_ready$study_id) | analysis_ready$study_id == ""),
    sum(analysis_ready$c_n <= 0 | analysis_ready$ex_n <= 0 | is.na(analysis_ready$c_n) | is.na(analysis_ready$ex_n)),
    sum(is.na(analysis_ready$c_mean) | is.na(analysis_ready$ex_mean)),
    sum(is.na(analysis_ready$c_sd) | is.na(analysis_ready$ex_sd) | analysis_ready$c_sd < 0 | analysis_ready$ex_sd < 0),
    sum(is.na(analysis_ready$c_se) | is.na(analysis_ready$ex_se) | analysis_ready$c_se < 0 | analysis_ready$ex_se < 0),
    sum(!analysis_ready$higher_better %in% c("yes", "no")),
    sum(!analysis_ready$comparison_structure %in% c("independent", "dependent"))
  )
) |>
  mutate(status = if_else(failures == 0, "pass", "fail"))

validation |>
  knitr::kable()

if (any(validation$status == "fail")) {
  stop("Analysis-ready data failed validation.")
}

readr::write_csv(with_statistics, statistics_path, na = "")
readr::write_csv(analysis_ready, analysis_path, na = "")

tibble(
  output = c(
    "Data/derived/PCN_data_ext_checking_v.3_with_derived_statistics.csv",
    "Data/derived/PCN_analysis_ready.csv"
  ),
  rows = c(nrow(with_statistics), nrow(analysis_ready)),
  columns = c(ncol(with_statistics), ncol(analysis_ready))
) |>
  knitr::kable()
```

# Session information

```{r}
#| label: session-information

sessionInfo()
```
