---
title: "Data preparation"
---

We keep the original extraction database unchanged and create two derived databases for the analysis.

```{r}
#| label: setup

pacman::p_load(
  dplyr,
  knitr,
  readr,
  stringr,
  tibble
)

project_root <- normalizePath("..", mustWork = TRUE)
data_dir <- file.path(project_root, "Data")
derived_dir <- file.path(data_dir, "derived")
dir.create(derived_dir, recursive = TRUE, showWarnings = FALSE)

# Define the input and output file paths.
original_path <- file.path(data_dir, "PCN_data_ext_checking_v.3.csv")
statistics_path <- file.path(
  derived_dir,
  "PCN_data_ext_checking_v.3_with_derived_statistics.csv"
)
analysis_path <- file.path(derived_dir, "PCN_analysis_ready.csv")
```

# Source data

We start with `Data/PCN_data_ext_checking_v.3.csv`. We record its checksum before creating the derived files.

::: callout-note
## Variable definitions

| Variable | Definition |
|:--|:--|
| `study_id` | Identifier for the primary research paper. |
| `es_id` | Unique identifier for each effect size. |
| `ex_id` | Identifier for an exposed group within an experiment and study. |
| `con_id` | Identifier for a control group within an experiment and study. |
| `out_id` | Identifier for a behavioural outcome. |
| `c_` | Prefix for control-group statistics. |
| `ex_` | Prefix for exposed-group statistics. |
| `mean`, `sd`, `se`, `n` | Mean, standard deviation, standard error, and sample size. |
| `median`, `q1`, `q3` | Median, first quartile, and third quartile. |
| `comparison_structure` | Indicates an independent- or dependent-group comparison. |
| `higher_better` | Indicates whether a higher outcome value represents better performance. |
:::

```{r}
#| label: read-source-data

# Record the checksum of the original file.
original_checksum <- unname(tools::md5sum(original_path))

# Read the original extraction database.
original <- readr::read_csv(
  original_path,
  na = c("", "NA"),
  name_repair = "unique_quiet",
  show_col_types = FALSE
)

# List the columns that the derivation requires.
required_columns <- c(
  "study_id", "ex_id", "con_id", "out_id", "es_id",
  "comparison_structure", "higher_better", "gest_stage_ex_standardised",
  "c_mean", "c_sd", "c_n", "c_se", "c_median", "c_q1", "c_q3",
  "ex_mean", "ex_sd", "ex_n", "ex_se", "ex_median", "ex_q1", "ex_q3"
)

if (length(setdiff(required_columns, names(original))) > 0) {
  stop("Required columns are missing from the original extraction database.")
}

# Remove unnamed empty export columns.
empty_unnamed_columns <- names(original)[
  str_detect(names(original), "^\\.\\.\\.[0-9]+$") &
    vapply(original, function(x) all(is.na(x)), logical(1))
]

source_for_derivation <- original |>
  select(-any_of(empty_unnamed_columns))

tibble(
  file = "Data/PCN_data_ext_checking_v.3.csv",
  rows = nrow(original),
  columns = ncol(original),
  studies = n_distinct(original$study_id),
  effect_sizes = n_distinct(original$es_id),
  checksum = original_checksum
) |>
  knitr::kable(caption = "Original extraction database.")
```

# Derivation of missing summary statistics

## Calculation methods

- We estimate a missing mean from Q1, the median, Q3, and sample size with the Luo et al. method [@luo2018].
- We estimate a missing SD from Q1, Q3, and sample size with the Wan et al. method [@wan2014].
- When the source reports an SE but not an SD, we calculate `SD = SE × sqrt(n)`.
- When the source does not report an SE, we calculate `SE = SD / sqrt(n)`.
- We retain existing extracted values unchanged.

$$
\widehat{M} =
\left(0.7 + \frac{0.39}{n}\right)\frac{Q_1 + Q_3}{2}
+ \left(0.3 - \frac{0.39}{n}\right)M_d
$$

$$
\widehat{SD} =
\frac{Q_3-Q_1}
{2\Phi^{-1}\left(\frac{0.75n-0.125}{n+0.25}\right)}
$$

We define one function for each published estimator and one function that completes the mean, SD, and SE for either treatment arm. The function also records the source of every analysis value.

```{r}
#| label: summary-statistic-functions

estimate_mean_luo <- function(q1, median, q3, n) {
  weight <- 0.7 + 0.39 / n
  weight * ((q1 + q3) / 2) + (1 - weight) * median
}

estimate_sd_wan <- function(q1, q3, n) {
  denominator <- 2 * stats::qnorm((0.75 * n - 0.125) / (n + 0.25))
  (q3 - q1) / denominator
}

complete_arm_statistics <- function(data, arm) {
  n <- data[[paste0(arm, "_n")]]
  mean_source <- data[[paste0(arm, "_mean")]]
  sd_source <- data[[paste0(arm, "_sd")]]
  se_source <- data[[paste0(arm, "_se")]]
  median <- data[[paste0(arm, "_median")]]
  q1 <- data[[paste0(arm, "_q1")]]
  q3 <- data[[paste0(arm, "_q3")]]

  mean_from_iqr <- is.na(mean_source) & complete.cases(q1, median, q3, n)
  sd_from_iqr <- is.na(sd_source) & complete.cases(q1, q3, n)
  sd_from_se <- is.na(sd_source) & !sd_from_iqr & !is.na(se_source) & !is.na(n)

  if (any(mean_from_iqr & !(q1 <= median & median <= q3))) {
    stop("Invalid Q1 <= median <= Q3 ordering in the ", arm, " arm.")
  }

  mean_analysis <- mean_source
  mean_analysis[mean_from_iqr] <- estimate_mean_luo(
    q1[mean_from_iqr], median[mean_from_iqr], q3[mean_from_iqr], n[mean_from_iqr]
  )

  sd_analysis <- sd_source
  sd_analysis[sd_from_iqr] <- estimate_sd_wan(
    q1[sd_from_iqr], q3[sd_from_iqr], n[sd_from_iqr]
  )
  sd_analysis[sd_from_se] <- se_source[sd_from_se] * sqrt(n[sd_from_se])

  se_analysis <- se_source
  se_from_sd <- is.na(se_analysis) & !is.na(sd_analysis) & !is.na(n)
  se_analysis[se_from_sd] <- sd_analysis[se_from_sd] / sqrt(n[se_from_sd])

  data[[paste0(arm, "_mean_analysis")]] <- mean_analysis
  data[[paste0(arm, "_mean_method")]] <- if_else(
    mean_from_iqr,
    "Luo et al. (2018): Q1, median, Q3, and n",
    "Extracted mean used unchanged"
  )
  data[[paste0(arm, "_sd_analysis")]] <- sd_analysis
  data[[paste0(arm, "_sd_method")]] <- case_when(
    sd_from_iqr ~ "Wan et al. (2014): Q1, Q3, and n",
    sd_from_se ~ "SD = SE * sqrt(n)",
    TRUE ~ "Extracted SD used unchanged"
  )
  data[[paste0(arm, "_se_analysis")]] <- se_analysis
  data[[paste0(arm, "_se_method")]] <- if_else(
    se_from_sd,
    "SE = SD / sqrt(n)",
    "Extracted SE used unchanged"
  )

  if (anyNA(mean_analysis) || anyNA(sd_analysis) || anyNA(se_analysis)) {
    stop("Unresolved summary statistic in the ", arm, " arm.")
  }

  data
}
```

We apply the functions to the control arm (`c`) and then to the exposed arm (`ex`). We summarise the calculation methods by arm and statistic.

```{r}
#| label: calculate-summary-statistics

with_statistics <- source_for_derivation |>
  complete_arm_statistics("c") |>
  complete_arm_statistics("ex")

method_counts <- bind_rows(
  with_statistics |> count(arm = "Control", statistic = "Mean", method = c_mean_method, name = "rows"),
  with_statistics |> count(arm = "Control", statistic = "SD", method = c_sd_method, name = "rows"),
  with_statistics |> count(arm = "Control", statistic = "SE", method = c_se_method, name = "rows"),
  with_statistics |> count(arm = "Exposed", statistic = "Mean", method = ex_mean_method, name = "rows"),
  with_statistics |> count(arm = "Exposed", statistic = "SD", method = ex_sd_method, name = "rows"),
  with_statistics |> count(arm = "Exposed", statistic = "SE", method = ex_se_method, name = "rows")
)

method_counts |>
  arrange(arm, statistic, desc(rows)) |>
  knitr::kable(caption = "Sources of the analysis statistics.")
```

# Analysis-ready data

We create an analysis-ready database with one mean, SD, SE, and sample size for each arm. We retain `ex_id` and `con_id` for the sampling dependency structure and `comparison_structure` for the lnRR calculation.

We also standardise character spacing and category labels before selecting the columns required for the analyses.

```{r}
#| label: create-analysis-ready-data

analysis_ready <- with_statistics |>
  mutate(
    across(where(is.character), str_squish),
    shared_control = str_to_lower(shared_control),
    comparison_structure = str_to_lower(comparison_structure),
    higher_better = str_to_lower(higher_better),
    gest_stage_ex_standardised = str_to_lower(gest_stage_ex_standardised),
    exposure_type = str_to_lower(exposure_type),
    exposure_type = recode(exposure_type, "intermitent" = "intermittent"),
    noise_type = str_to_lower(noise_type),
    noise_type_2 = str_to_lower(noise_type_2),
    control_conditions = str_to_lower(control_conditions),
    offspring_sex = str_to_lower(offspring_sex),
    outcome_type = str_to_lower(outcome_type),
    data_type = str_to_lower(data_type),
    measurement_timing = str_to_lower(measurement_timing),
    strain = recode(strain, "Wistar albino rat" = "Wistar rat")
  ) |>
  transmute(
    study_id, title, doi, publish_in,
    ex_id, con_id, out_id, es_id, shared_control,
    sp_latin, sp_common, strain, dam_age, gest_stage_ex_standardised,
    exposure_span_d, exposure_type, exposure_session_duration_h,
    noise_type, noise_type_2, loudness_dB, frequency_Hz,
    control_conditions, offspring_sex, offspring_age_d,
    experimental_procedures, outcome_type, assay_type,
    measurement_variable, data_type, data_units, measurement_timing,
    data_source, data_file, c_a_id, ex_a_id, comparison,
    comparison_structure, higher_better,
    c_mean = c_mean_analysis,
    c_sd = c_sd_analysis,
    c_n,
    c_se = c_se_analysis,
    c_mean_method, c_sd_method, c_se_method,
    ex_mean = ex_mean_analysis,
    ex_sd = ex_sd_analysis,
    ex_n,
    ex_se = ex_se_analysis,
    ex_mean_method, ex_sd_method, ex_se_method
  )
```

## Data checks

We check the identifiers, summary statistics, sample sizes, and coded variables before exporting the analysis-ready database. The chapter stops if any check fails.

```{r}
#| label: validate-analysis-data

validation <- tibble(
  check = c(
    "One row per es_id",
    "Study identifiers present",
    "Positive sample sizes",
    "Complete means",
    "Complete non-negative SDs",
    "Complete non-negative SEs",
    "Valid higher_better values",
    "Valid comparison_structure values"
  ),
  failures = c(
    sum(is.na(analysis_ready$es_id)) + sum(duplicated(analysis_ready$es_id)),
    sum(is.na(analysis_ready$study_id) | analysis_ready$study_id == ""),
    sum(is.na(analysis_ready$c_n) | is.na(analysis_ready$ex_n) |
          analysis_ready$c_n <= 0 | analysis_ready$ex_n <= 0),
    sum(is.na(analysis_ready$c_mean) | is.na(analysis_ready$ex_mean)),
    sum(is.na(analysis_ready$c_sd) | is.na(analysis_ready$ex_sd) |
          analysis_ready$c_sd < 0 | analysis_ready$ex_sd < 0),
    sum(is.na(analysis_ready$c_se) | is.na(analysis_ready$ex_se) |
          analysis_ready$c_se < 0 | analysis_ready$ex_se < 0),
    sum(!analysis_ready$higher_better %in% c("yes", "no")),
    sum(!analysis_ready$comparison_structure %in% c("independent", "dependent"))
  )
) |>
  mutate(status = if_else(failures == 0, "Pass", "Fail"))

validation |>
  knitr::kable(caption = "Analysis-ready data checks.")

if (any(validation$status == "Fail")) {
  stop("The analysis-ready database failed validation.")
}
```

# Data files

We save the completed extraction database and the analysis-ready database as comma-separated files. We then compare the original checksum with its starting value to confirm that we did not overwrite the source file.

```{r}
#| label: export-data-files

readr::write_csv(with_statistics, statistics_path, na = "")
readr::write_csv(analysis_ready, analysis_path, na = "")

if (!identical(original_checksum, unname(tools::md5sum(original_path)))) {
  stop("The original extraction file changed during data preparation.")
}

tibble(
  database = c(
    "Original extraction plus R-derived statistics",
    "Analysis-ready data"
  ),
  file = c(
    "Data/derived/PCN_data_ext_checking_v.3_with_derived_statistics.csv",
    "Data/derived/PCN_analysis_ready.csv"
  ),
  rows = c(nrow(with_statistics), nrow(analysis_ready)),
  columns = c(ncol(with_statistics), ncol(analysis_ready))
) |>
  knitr::kable(caption = "Derived data files.")
```

::: callout-note
## Session information

```{r}
#| label: session-information

sessionInfo()
```
:::
