---
title: "Effect-size calculation: lnRR"
---

## The log response ratio

This project uses the log response ratio (lnRR), the natural logarithm of the ratio of the treatment mean to the control mean [@hedges1999; @lajeunesse2015], as agreed in the protocol.

::: {.callout-note}
## Formula

$$
\ln RR = \ln\left(\frac{M_T}{M_C}\right) + \frac{1}{2}\left(\frac{SD_T^2}{N_T M_T^2} - \frac{SD_C^2}{N_C M_C^2}\right)
$$

where $M_T$, $SD_T$, $N_T$ are the mean, standard deviation, and sample size of the noise-exposed group, and $M_C$, $SD_C$, $N_C$ are the same for the control group. This is the bias-corrected estimator implemented by `metafor::escalc(measure = "ROM")`.
:::

::: {.callout-note collapse="true"}
## A note on the "robust estimator" wording in the protocol

The protocol cites Nakagawa et al. [@nakagawa2023c] for a "robust estimator of lnRR". That paper (Ecology Letters, 2023) proposes a method specifically for meta-analyses with a substantial share of *missing* standard deviations, using sample-size-based weighting instead of the usual SD-based sampling variance for the affected rows. In this dataset only 31/207 rows needed a derived SD (Chapter 1), and all 207 have a usable SD after that step, so the standard `escalc(measure = "ROM")` estimator above is used throughout. If a future update to this dataset has many more missing SDs, revisit this decision and implement the Nakagawa et al. (2023) weighting instead.
:::

Every treatment-control pair in this dataset uses independent animals (confirmed in the protocol deviation review: "the control and exposed animals are distinct for every comparison"), so the same independent-arm formula applies to all 207 effect sizes - there is no paired/dependent-design branch to handle.

Positive lnRR is not yet interpretable on its own: it only means "the exposed-group mean was higher than the control-group mean" for a given outcome, and outcomes are scored in both directions (e.g., more time in the open arms of an elevated-plus maze is *less* anxiety-like behaviour, while more immobility in a forced-swim test is *more* depression-like behaviour). The next step harmonizes this so the sign is comparable across outcomes.

```{r}
#| label: setup

if (!requireNamespace("pacman", quietly = TRUE)) {
  stop(
    'Package "pacman" is required. Install it once with ',
    'install.packages("pacman").'
  )
}

analysis_packages <- c("dplyr", "readr", "metafor", "knitr", "ggplot2")
pacman::p_load(char = analysis_packages)

options(dplyr.summarise.inform = FALSE)
knitr::opts_chunk$set(comment = "#>")

project_root <- normalizePath("..", mustWork = TRUE)
derived_dir <- file.path(project_root, "Data", "derived")

db <- readr::read_csv(
  file.path(derived_dir, "pcn_data_analysis_ready.csv"),
  show_col_types = FALSE
)
```

## Calculate lnRR and its sampling variance

```{r}
#| label: calculate-lnrr

es <- metafor::escalc(
  measure = "ROM",
  m1i = ex_mean_final, sd1i = ex_sd_final, n1i = ex_n,
  m2i = c_mean_final,  sd2i = c_sd_final,  n2i = c_n,
  data = db
)

es <- es %>%
  rename(lnrr_raw = yi, lnrr_var = vi) %>%
  mutate(
    lnrr_se = sqrt(lnrr_var),
    higher_better = trimws(higher_better),
    lnrr_harm = ifelse(tolower(higher_better) == "yes", -lnrr_raw, lnrr_raw)
  )

if (any(!es$higher_better %in% c("Yes", "No"))) {
  stop("higher_better must be exactly 'Yes' or 'No' for every row.")
}
if (any(is.na(es$lnrr_raw) | is.na(es$lnrr_var))) {
  stop("escalc produced a missing lnRR or variance; check the input means/SDs/n.")
}

tibble(
  effect_sizes = nrow(es),
  studies = n_distinct(es$study_id),
  flipped_for_direction = sum(tolower(es$higher_better) == "yes")
) %>%
  knitr::kable()
```

`lnrr_raw` is the estimator applied directly to the exposed vs. control means. `lnrr_harm` multiplies `lnrr_raw` by $-1$ whenever `higher_better == "Yes"` (a higher raw value on that assay means *less* anxiety/depression), so that a positive `lnrr_harm` consistently means **more** anxiety- or depression-like behaviour under maternal chronic noise exposure, and a negative `lnrr_harm` means less. `lnrr_harm` is the value used in every model from Chapter 3 onward; `lnrr_raw` is kept only for auditing.

## Effect-size table

```{r}
#| label: effect-size-table

es %>%
  select(
    study_id, es_id, outcome_type, assay_type, measurement_variable,
    higher_better, ex_mean_final, ex_sd_final, ex_n,
    c_mean_final, c_sd_final, c_n,
    lnrr_raw, lnrr_harm, lnrr_var, lnrr_se
  ) %>%
  knitr::kable(digits = 3)
```

## Distribution of effect sizes

```{r}
#| label: lnrr-histogram
#| fig-cap: "Distribution of harmonized lnRR across all 207 effect sizes. Positive values indicate more anxiety- or depression-like behaviour under maternal chronic noise exposure."

ggplot(es, aes(x = lnrr_harm)) +
  geom_histogram(bins = 30, fill = "#477AA8", color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#B24745") +
  labs(x = "Harmonized lnRR", y = "Number of effect sizes") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: lnrr-se-scatter
#| fig-cap: "lnRR against its standard error. Points far to the right with a small effect size are the most influential on the funnel-plot check in Chapter 5."

ggplot(es, aes(x = lnrr_se, y = lnrr_harm)) +
  geom_point(alpha = 0.5, color = "#477AA8") +
  labs(x = "Standard error of lnRR", y = "Harmonized lnRR") +
  theme_minimal(base_size = 12)
```

## Export

```{r}
#| label: export

readr::write_csv(es, file.path(derived_dir, "db_effect_sizes.csv"), na = "")

tibble(output = "Data/derived/db_effect_sizes.csv") %>% knitr::kable()
```

## Session information

```{r}
#| label: session-information

sessionInfo()
```
