---
title: "Effect size calculation"
---

This chapter calculates the log response ratio (lnRR) and its sampling variance from the analysis-ready database.

```{r}
#| label: setup
#| message: false
#| warning: false

pacman::p_load(
  dplyr,
  ggplot2,
  knitr,
  metafor,
  readr,
  tibble
)

project_root <- normalizePath("..", mustWork = TRUE)
derived_dir <- file.path(project_root, "Data", "derived")

db <- readr::read_csv(
  file.path(derived_dir, "PCN_analysis_ready.csv"),
  show_col_types = FALSE
)
```

# lnRR

For continuous outcomes, lnRR is calculated with the same design branches as the reference workflow [@hedges1999; @lajeunesse2015]. `comparison_structure == "independent"` uses `metafor::escalc(measure = "ROM")`. `comparison_structure == "dependent"` uses `metafor::escalc(measure = "ROMC", ri = 0.5)` and the average of the exposed and control sample sizes.

$$
\ln RR =
\ln\left(\frac{M_T}{M_C}\right)
+ \frac{1}{2}
\left(
\frac{SD_T^2}{N_T M_T^2}
- \frac{SD_C^2}{N_C M_C^2}
\right)
$$

`higher_better` is used to place outcomes in one direction. The lnRR value is multiplied by -1 when `higher_better == "yes"`.

# Boundary corrections

Log ratios require positive means. For non-percentage outcomes, an exact-zero mean is replaced with 0.9 times the smallest positive mean in the same outcome stratum. An exact-zero SD is replaced with 0.05. Percentage outcomes are kept inside their 0-100 bounds and are calculated after an arcsine-square-root transformation, following the reference workflow.

```{r}
#| label: boundary-correction-functions

min_positive <- function(x) {
  x <- x[is.finite(x) & x > 0]
  if (length(x) == 0) NA_real_ else min(x)
}

apply_continuity_correction <- function(data) {
  strata <- c("assay_type", "measurement_variable", "data_type", "data_units")

  minima <- data |>
    group_by(across(all_of(strata))) |>
    summarise(
      min_c_mean = min_positive(c_mean),
      min_ex_mean = min_positive(ex_mean),
      .groups = "drop"
    )

  global_c_mean <- min_positive(data$c_mean)
  global_ex_mean <- min_positive(data$ex_mean)

  data |>
    left_join(minima, by = strata) |>
    mutate(
      c_mean_continuity_corrected = !is.na(c_mean) & c_mean == 0,
      ex_mean_continuity_corrected = !is.na(ex_mean) & ex_mean == 0,
      c_sd_continuity_corrected = !is.na(c_sd) & c_sd == 0,
      ex_sd_continuity_corrected = !is.na(ex_sd) & ex_sd == 0,
      c_mean = if_else(
        c_mean_continuity_corrected,
        0.9 * coalesce(min_c_mean, global_c_mean),
        c_mean
      ),
      ex_mean = if_else(
        ex_mean_continuity_corrected,
        0.9 * coalesce(min_ex_mean, global_ex_mean),
        ex_mean
      ),
      c_sd = if_else(c_sd_continuity_corrected, 0.05, c_sd),
      ex_sd = if_else(ex_sd_continuity_corrected, 0.05, ex_sd)
    ) |>
    select(-min_c_mean, -min_ex_mean)
}
```

# Calculation

```{r}
#| label: calculate-lnrr

if (any(!db$comparison_structure %in% c("independent", "dependent"))) {
  stop("comparison_structure must be independent or dependent.")
}

continuous <- db |>
  filter(data_type != "percentage") |>
  apply_continuity_correction()

percentage <- db |>
  filter(data_type == "percentage") |>
  mutate(
    c_mean_continuity_corrected = c_mean %in% c(0, 100),
    ex_mean_continuity_corrected = ex_mean %in% c(0, 100),
    c_sd_continuity_corrected = c_sd == 0,
    ex_sd_continuity_corrected = ex_sd == 0,
    c_mean = case_when(c_mean == 0 ~ 0.5, c_mean == 100 ~ 99.5, TRUE ~ c_mean),
    ex_mean = case_when(ex_mean == 0 ~ 0.5, ex_mean == 100 ~ 99.5, TRUE ~ ex_mean),
    c_sd = if_else(c_sd == 0, 0.5, c_sd),
    ex_sd = if_else(ex_sd == 0, 0.5, ex_sd)
  )

db_effect_sizes <- bind_rows(continuous, percentage) |>
  arrange(match(es_id, db$es_id)) |>
  mutate(lnRR = NA_real_, lnRR_var = NA_real_)

asin_trans <- function(proportion) asin(sqrt(proportion))

for (i in seq_len(nrow(db_effect_sizes))) {
  ex_n <- db_effect_sizes$ex_n[i]
  c_n <- db_effect_sizes$c_n[i]
  ex_mean <- db_effect_sizes$ex_mean[i]
  c_mean <- db_effect_sizes$c_mean[i]
  ex_sd <- db_effect_sizes$ex_sd[i]
  c_sd <- db_effect_sizes$c_sd[i]
  data_type <- db_effect_sizes$data_type[i]
  comparison_structure <- db_effect_sizes$comparison_structure[i]

  if (data_type != "percentage") {
    if (comparison_structure == "independent") {
      effect <- metafor::escalc(
        measure = "ROM",
        n1i = ex_n, n2i = c_n,
        m1i = ex_mean, m2i = c_mean,
        sd1i = ex_sd, sd2i = c_sd,
        var.names = c("lnRR", "lnRR_var")
      )
    } else {
      effect <- metafor::escalc(
        measure = "ROMC",
        ni = (ex_n + c_n) / 2,
        m1i = ex_mean, m2i = c_mean,
        sd1i = ex_sd, sd2i = c_sd,
        ri = 0.5,
        var.names = c("lnRR", "lnRR_var")
      )
    }

    db_effect_sizes$lnRR[i] <- effect$lnRR
    db_effect_sizes$lnRR_var[i] <- effect$lnRR_var
  } else {
    if (max(c(ex_mean, c_mean), na.rm = TRUE) > 1) {
      ex_mean <- ex_mean / 100
      c_mean <- c_mean / 100
      ex_sd <- ex_sd / 100
      c_sd <- c_sd / 100
    }

    ex_sd_transformed <- sqrt(ex_sd^2 / (4 * ex_mean * (1 - ex_mean)))
    c_sd_transformed <- sqrt(c_sd^2 / (4 * c_mean * (1 - c_mean)))
    ex_mean_transformed <- asin_trans(ex_mean)
    c_mean_transformed <- asin_trans(c_mean)

    v_ex <- ex_sd_transformed^2 / (ex_mean_transformed^2 * ex_n)
    v_c <- c_sd_transformed^2 / (c_mean_transformed^2 * c_n)

    db_effect_sizes$lnRR[i] <- log(ex_mean_transformed / c_mean_transformed)
    db_effect_sizes$lnRR_var[i] <- if (comparison_structure == "independent") {
      v_ex + v_c
    } else {
      v_ex + v_c - 2 * 0.5 * sqrt(v_ex) * sqrt(v_c)
    }
  }

  if (db_effect_sizes$higher_better[i] == "yes") {
    db_effect_sizes$lnRR[i] <- -db_effect_sizes$lnRR[i]
  }
}

db_effect_sizes <- db_effect_sizes |>
  mutate(lnRR_se = sqrt(lnRR_var))

if (nrow(db_effect_sizes) != nrow(db) ||
    anyNA(db_effect_sizes$lnRR) ||
    anyNA(db_effect_sizes$lnRR_var) ||
    any(db_effect_sizes$lnRR_var < 0)) {
  stop("Effect-size calculation failed validation.")
}

tibble(
  effect_sizes = nrow(db_effect_sizes),
  studies = n_distinct(db_effect_sizes$study_id),
  mean_corrections = sum(
    db_effect_sizes$c_mean_continuity_corrected |
      db_effect_sizes$ex_mean_continuity_corrected
  ),
  sd_corrections = sum(
    db_effect_sizes$c_sd_continuity_corrected |
      db_effect_sizes$ex_sd_continuity_corrected
  )
) |>
  knitr::kable()

db_effect_sizes |>
  count(comparison_structure, name = "effect_sizes") |>
  knitr::kable(caption = "Effect sizes calculated by comparison structure.")
```

# Effect-size distribution

```{r}
#| label: effect-size-distributions
#| fig-width: 6
#| fig-height: 4

p_effect <- ggplot(db_effect_sizes, aes(x = lnRR)) +
  geom_histogram(bins = 30) +
  labs(x = "lnRR", y = "Count") +
  theme_minimal()

p_effect
```

# Export

```{r}
#| label: export-effect-sizes

readr::write_csv(
  db_effect_sizes,
  file.path(derived_dir, "db_effect_sizes.csv"),
  na = ""
)

tibble(output = "Data/derived/db_effect_sizes.csv") |>
  knitr::kable()
```

# Session information

```{r}
#| label: session-information

sessionInfo()
```
