---
title: "Effect size calculation"
---

```{r}
#| label: setup

pacman::p_load(
  dplyr,
  ggplot2,
  knitr,
  metafor,
  readr,
  tibble,
  tidyr
)

project_root <- normalizePath("..", mustWork = TRUE)
derived_dir <- file.path(project_root, "Data", "derived")

db <- readr::read_csv(
  file.path(derived_dir, "PCN_analysis_ready.csv"),
  show_col_types = FALSE
)
```

# lnRR

The log response ratio (lnRR) compares the mean outcome in the prenatal-noise group with the control group.

The bias-corrected lnRR is:

$$
\ln RR =
\ln\left(\frac{M_T}{M_C}\right)
+ \frac{1}{2}
\left(
\frac{SD_T^2}{N_T M_T^2}
- \frac{SD_C^2}{N_C M_C^2}
\right),
$$

where $M$, $SD$, and $N$ are the mean, standard deviation, and sample size for the exposed ($T$) and control ($C$) groups [@hedges1999; @lajeunesse2015].

# Data preparation for lnRR

## Data types

- Non-percentage outcomes are calculated on their extracted scale.
- Percentage outcomes are calculated after an arcsine-square-root transformation.

## Boundary corrections

- For non-percentage outcomes, a zero mean is replaced by 0.9 times the smallest positive mean in the same outcome stratum.
- For all outcomes, a zero SD is replaced by the smallest positive SD for the same treatment arm and outcome stratum.
- Percentage means of 0 and 100 are replaced by 0.5 and 99.5, respectively.

## Comparison structure

- `comparison_structure == "independent"` uses `metafor::escalc(measure = "ROM")`.
- `comparison_structure == "dependent"` uses `metafor::escalc(measure = "ROMC", ri = 0.5)`.

## Direction harmonisation

The lnRR sign is multiplied by -1 when `higher_better == "yes"` so that outcome direction is consistent across effect sizes.

```{r}
#| label: boundary-correction-functions

minimum_positive <- function(x) {
  x <- x[is.finite(x) & x > 0]
  if (length(x) == 0) NA_real_ else min(x)
}

apply_zero_sd_corrections <- function(data) {
  strata <- c("assay_type", "measurement_variable", "data_type", "data_units")

  local_minima <- data |>
    group_by(across(all_of(strata))) |>
    summarise(
      minimum_control_sd = minimum_positive(c_sd),
      minimum_exposed_sd = minimum_positive(ex_sd),
      .groups = "drop"
    )

  corrected <- data |>
    left_join(local_minima, by = strata) |>
    mutate(
      c_sd_continuity_corrected = !is.na(c_sd) & c_sd == 0,
      ex_sd_continuity_corrected = !is.na(ex_sd) & ex_sd == 0,
      c_sd = if_else(c_sd_continuity_corrected, minimum_control_sd, c_sd),
      ex_sd = if_else(ex_sd_continuity_corrected, minimum_exposed_sd, ex_sd)
    )

  if (any(corrected$c_sd_continuity_corrected & is.na(corrected$minimum_control_sd)) ||
      any(corrected$ex_sd_continuity_corrected & is.na(corrected$minimum_exposed_sd))) {
    stop("A zero SD has no positive SD in the same treatment arm and outcome stratum.")
  }

  corrected |>
    select(-minimum_control_sd, -minimum_exposed_sd)
}

apply_nonpercentage_corrections <- function(data) {
  strata <- c("assay_type", "measurement_variable", "data_type", "data_units")

  local_minima <- data |>
    group_by(across(all_of(strata))) |>
    summarise(
      minimum_control_mean = minimum_positive(c_mean),
      minimum_exposed_mean = minimum_positive(ex_mean),
      .groups = "drop"
    )

  global_control_minimum <- minimum_positive(data$c_mean)
  global_exposed_minimum <- minimum_positive(data$ex_mean)

  data |>
    left_join(local_minima, by = strata) |>
    mutate(
      c_mean_continuity_corrected = !is.na(c_mean) & c_mean == 0,
      ex_mean_continuity_corrected = !is.na(ex_mean) & ex_mean == 0,
      c_mean = if_else(
        c_mean_continuity_corrected,
        0.9 * coalesce(minimum_control_mean, global_control_minimum),
        c_mean
      ),
      ex_mean = if_else(
        ex_mean_continuity_corrected,
        0.9 * coalesce(minimum_exposed_mean, global_exposed_minimum),
        ex_mean
      )
    ) |>
    select(-minimum_control_mean, -minimum_exposed_mean) |>
    apply_zero_sd_corrections()
}

prepare_lnrr_data <- function(data) {
  nonpercentage <- data |>
    filter(data_type != "percentage") |>
    apply_nonpercentage_corrections()

  percentage <- data |>
    filter(data_type == "percentage") |>
    mutate(
      c_mean_continuity_corrected = c_mean %in% c(0, 100),
      ex_mean_continuity_corrected = ex_mean %in% c(0, 100),
      c_mean = case_when(c_mean == 0 ~ 0.5, c_mean == 100 ~ 99.5, TRUE ~ c_mean),
      ex_mean = case_when(ex_mean == 0 ~ 0.5, ex_mean == 100 ~ 99.5, TRUE ~ ex_mean)
    ) |>
    apply_zero_sd_corrections()

  bind_rows(nonpercentage, percentage) |>
    arrange(match(es_id, data$es_id))
}
```

```{r}
#| label: boundary-correction-summary

db_prepared <- prepare_lnrr_data(db)

boundary_corrections <- db |>
  select(
    es_id, study_id,
    c_mean_old = c_mean,
    ex_mean_old = ex_mean,
    c_sd_old = c_sd,
    ex_sd_old = ex_sd
  ) |>
  left_join(
    db_prepared |>
      select(
        es_id,
        c_mean_new = c_mean,
        ex_mean_new = ex_mean,
        c_sd_new = c_sd,
        ex_sd_new = ex_sd,
        c_mean_continuity_corrected,
        ex_mean_continuity_corrected,
        c_sd_continuity_corrected,
        ex_sd_continuity_corrected
      ),
    by = "es_id"
  ) |>
  mutate(
    any_correction = c_mean_continuity_corrected |
      ex_mean_continuity_corrected |
      c_sd_continuity_corrected |
      ex_sd_continuity_corrected
  )

boundary_corrections |>
  summarise(
    `Total effect sizes` = n(),
    `Any correction` = sum(any_correction),
    `Control mean` = sum(c_mean_continuity_corrected),
    `Exposed mean` = sum(ex_mean_continuity_corrected),
    `Control SD` = sum(c_sd_continuity_corrected),
    `Exposed SD` = sum(ex_sd_continuity_corrected)
  ) |>
  knitr::kable(caption = "Boundary corrections by statistic.")

modified_statistics <- bind_rows(
  boundary_corrections |>
    filter(c_mean_continuity_corrected) |>
    transmute(study_id, es_id, statistic = "Control mean", original = c_mean_old, corrected = c_mean_new),
  boundary_corrections |>
    filter(ex_mean_continuity_corrected) |>
    transmute(study_id, es_id, statistic = "Exposed mean", original = ex_mean_old, corrected = ex_mean_new),
  boundary_corrections |>
    filter(c_sd_continuity_corrected) |>
    transmute(study_id, es_id, statistic = "Control SD", original = c_sd_old, corrected = c_sd_new),
  boundary_corrections |>
    filter(ex_sd_continuity_corrected) |>
    transmute(study_id, es_id, statistic = "Exposed SD", original = ex_sd_old, corrected = ex_sd_new)
) |>
  arrange(study_id, es_id, statistic)

modified_statistics |>
  knitr::kable(digits = 3, caption = "Statistics modified by a boundary correction.")
```

```{r}
#| label: calculate-lnrr

calculate_lnrr <- function(data, within_subject_correlation = 0.5) {
  if (any(!data$comparison_structure %in% c("independent", "dependent"))) {
    stop("comparison_structure must be independent or dependent.")
  }

  result <- data |>
    mutate(lnRR = NA_real_, lnRR_var = NA_real_)

  arcsine_sqrt <- function(proportion) asin(sqrt(proportion))

  for (i in seq_len(nrow(result))) {
    ex_n <- result$ex_n[i]
    c_n <- result$c_n[i]
    ex_mean <- result$ex_mean[i]
    c_mean <- result$c_mean[i]
    ex_sd <- result$ex_sd[i]
    c_sd <- result$c_sd[i]
    data_type <- result$data_type[i]
    comparison_structure <- result$comparison_structure[i]

    if (data_type != "percentage") {
      if (comparison_structure == "independent") {
        effect <- metafor::escalc(
          measure = "ROM",
          n1i = ex_n, n2i = c_n,
          m1i = ex_mean, m2i = c_mean,
          sd1i = ex_sd, sd2i = c_sd,
          var.names = c("lnRR", "lnRR_var")
        )
      } else {
        effect <- metafor::escalc(
          measure = "ROMC",
          ni = (ex_n + c_n) / 2,
          m1i = ex_mean, m2i = c_mean,
          sd1i = ex_sd, sd2i = c_sd,
          ri = within_subject_correlation,
          var.names = c("lnRR", "lnRR_var")
        )
      }

      result$lnRR[i] <- effect$lnRR
      result$lnRR_var[i] <- effect$lnRR_var
    } else {
      if (max(c(ex_mean, c_mean), na.rm = TRUE) > 1) {
        ex_mean <- ex_mean / 100
        c_mean <- c_mean / 100
        ex_sd <- ex_sd / 100
        c_sd <- c_sd / 100
      }

      ex_sd_transformed <- sqrt(ex_sd^2 / (4 * ex_mean * (1 - ex_mean)))
      c_sd_transformed <- sqrt(c_sd^2 / (4 * c_mean * (1 - c_mean)))
      ex_mean_transformed <- arcsine_sqrt(ex_mean)
      c_mean_transformed <- arcsine_sqrt(c_mean)

      variance_exposed <- ex_sd_transformed^2 / (ex_mean_transformed^2 * ex_n)
      variance_control <- c_sd_transformed^2 / (c_mean_transformed^2 * c_n)

      result$lnRR[i] <- log(ex_mean_transformed / c_mean_transformed)
      result$lnRR_var[i] <- if (comparison_structure == "independent") {
        variance_exposed + variance_control
      } else {
        variance_exposed + variance_control -
          2 * within_subject_correlation *
          sqrt(variance_exposed) * sqrt(variance_control)
      }
    }

    if (result$higher_better[i] == "yes") {
      result$lnRR[i] <- -result$lnRR[i]
    }
  }

  result
}

db_effect_sizes <- calculate_lnrr(db_prepared)

if (
  nrow(db_effect_sizes) != nrow(db) ||
    anyNA(db_effect_sizes$lnRR) ||
    anyNA(db_effect_sizes$lnRR_var) ||
    any(db_effect_sizes$lnRR_var < 0)
) {
  stop("Effect-size calculation failed validation.")
}

db_effect_sizes |>
  count(comparison_structure, name = "effect_sizes") |>
  knitr::kable(caption = "Effect sizes by comparison structure.")
```

# Effect-size database

```{r}
#| label: export-effect-sizes

readr::write_csv(
  db_effect_sizes,
  file.path(derived_dir, "db_effect_sizes.csv"),
  na = ""
)

tibble(
  file = "Data/derived/db_effect_sizes.csv",
  rows = nrow(db_effect_sizes),
  studies = n_distinct(db_effect_sizes$study_id),
  effect_sizes = n_distinct(db_effect_sizes$es_id)
) |>
  knitr::kable(caption = "Effect-size database.")
```

# lnRR distributions

The histograms show the calculated lnRR values and their sampling variances (`lnRR_var`).

```{r}
#| label: lnrr-histogram
#| fig-cap: "Distribution of lnRR effect sizes."
#| fig-width: 7
#| fig-height: 4

ggplot(db_effect_sizes, aes(x = lnRR)) +
  geom_histogram(
    bins = 30,
    boundary = 0,
    colour = "white",
    linewidth = 0.2
  ) +
  labs(x = "lnRR", y = "Count") +
  theme_minimal()
```

```{r}
#| label: lnrr-var-histogram
#| fig-cap: "Distribution of lnRR sampling variances."
#| fig-width: 7
#| fig-height: 4

ggplot(db_effect_sizes, aes(x = lnRR_var)) +
  geom_histogram(
    bins = 30,
    boundary = 0,
    colour = "white",
    linewidth = 0.2
  ) +
  labs(x = "lnRR_var", y = "Count") +
  theme_minimal()
```

::: callout-note
## Session information

```{r}
#| label: session-information

sessionInfo()
```
:::
