---
title: "Effect size calculation"
---

We load the analysis-ready database created in **Data preparation** and calculate three effect sizes: the log response ratio (lnRR), log variability ratio (lnVR), and log coefficient of variation ratio (lnCVR). We also calculate the sampling variance of each effect size.

```{r}
#| label: setup

pacman::p_load(
  dplyr,
  ggplot2,
  knitr,
  metafor,
  readr,
  tibble,
  tidyr
)

project_root <- normalizePath("..", mustWork = TRUE)
derived_dir <- file.path(project_root, "Data", "derived")

# Load the analysis-ready database.
db <- readr::read_csv(
  file.path(derived_dir, "PCN_analysis_ready.csv"),
  show_col_types = FALSE
)
```

# lnRR

The log response ratio (lnRR) compares the mean outcome in the prenatal-noise group with the control group.

The bias-corrected lnRR is:

$$
\ln RR =
\ln\left(\frac{M_T}{M_C}\right)
+ \frac{1}{2}
\left(
\frac{SD_T^2}{N_T M_T^2}
- \frac{SD_C^2}{N_C M_C^2}
\right),
$$

where $M$, $SD$, and $N$ are the mean, standard deviation, and sample size for the exposed ($T$) and control ($C$) groups [@hedges1999; @lajeunesse2015].

# lnVR

The log variability ratio (lnVR) compares the standard deviation of the prenatal-noise treatment group with the standard deviation of the control group. We use it to measure absolute variability.

The bias-corrected lnVR is:

$$
\ln VR =
\ln\left(\frac{SD_T}{SD_C}\right)
+ \frac{1}{2}
\left(
\frac{1}{N_T-1}
- \frac{1}{N_C-1}
\right).
$$

A positive lnVR indicates greater absolute variability in the treatment group, while a negative lnVR indicates lower absolute variability [@nakagawa2015].

# lnCVR

The log coefficient of variation ratio (lnCVR) compares variability relative to the group mean. We first define the coefficient of variation as $CV = SD/M$.

The bias-corrected lnCVR is:

$$
\ln CVR =
\ln\left(\frac{CV_T}{CV_C}\right)
+ \frac{1}{2}
\left(
\frac{1}{N_T-1}
- \frac{1}{N_C-1}
\right)
- \frac{1}{2}
\left(
\frac{SD_T^2}{N_T M_T^2}
- \frac{SD_C^2}{N_C M_C^2}
\right).
$$

A positive lnCVR indicates greater relative variability in the treatment group, while a negative lnCVR indicates lower relative variability [@senior2020].

::: callout-note
## Variable definitions

| Variable | Definition |
|:--|:--|
| `c_mean`, `c_sd`, `c_n` | Control-group mean, standard deviation, and sample size. |
| `ex_mean`, `ex_sd`, `ex_n` | Exposed-group mean, standard deviation, and sample size. |
| `data_type` | Scale of the extracted outcome, including percentage outcomes. |
| `assay_type` | Behavioural test or assay. |
| `measurement_variable` | Behavioural variable reported for the assay. |
| `data_units` | Units of the reported measurement. |
| `comparison_structure` | Indicates an independent- or dependent-group comparison. |
| `higher_better` | Indicates whether a higher outcome value represents better performance. |
| `lnRR` | Bias-corrected log response ratio. |
| `lnRR_var` | Sampling variance of lnRR. |
| `lnVR` | Bias-corrected log variability ratio. |
| `lnVR_var` | Sampling variance of lnVR. |
| `lnCVR` | Bias-corrected log coefficient of variation ratio. |
| `lnCVR_var` | Sampling variance of lnCVR. |
:::

# Data preparation

## Data types

- We calculate lnRR for every outcome. We apply an arcsine-square-root transformation before calculating lnRR for percentage outcomes.
- We calculate lnVR and lnCVR only for non-percentage outcomes. We exclude percentage outcomes because their bounded scale creates a mechanical relationship between the mean and variance.

## Boundary corrections

- For non-percentage outcomes, we replace a zero mean with 0.9 times the smallest positive mean in the same outcome stratum.
- For all outcomes, we replace a zero SD with the smallest positive SD for the same treatment arm and outcome stratum.
- For percentage outcomes, we replace means of 0 and 100 with 0.5 and 99.5, respectively.

We define an outcome stratum with `assay_type`, `measurement_variable`, `data_type`, and `data_units`.

## Comparison structure

- For independent comparisons, we use `ROM`, `VR`, and `CVR` in `metafor::escalc()`.
- For dependent comparisons, we use `ROMC`, `VRC`, and `CVRC` with a within-subject correlation of 0.5.

## Direction harmonisation

When `higher_better == "yes"`, we multiply lnRR by -1 to keep the behavioural direction consistent across effect sizes. After this harmonisation, a positive lnRR represents more anxiety- or depression-like behaviour in the treatment group, while a negative lnRR represents less anxiety- or depression-like behaviour.

We do not reverse lnVR or lnCVR because standard deviations do not have an anxiety- or depression-like direction. For both variability metrics, positive values represent greater variability in the treatment group and negative values represent lower variability.

The following functions apply these rules while preserving the original row order.

```{r}
#| label: boundary-correction-functions

# Return the smallest positive value in a vector.
minimum_positive <- function(x) {
  x <- x[is.finite(x) & x > 0]
  if (length(x) == 0) NA_real_ else min(x)
}

apply_zero_sd_corrections <- function(data) {
  strata <- c("assay_type", "measurement_variable", "data_type", "data_units")

  # Calculate the smallest positive SD within each outcome stratum and arm.
  local_minima <- data |>
    group_by(across(all_of(strata))) |>
    summarise(
      minimum_control_sd = minimum_positive(c_sd),
      minimum_exposed_sd = minimum_positive(ex_sd),
      .groups = "drop"
    )

  corrected <- data |>
    left_join(local_minima, by = strata) |>
    mutate(
      c_sd_continuity_corrected = !is.na(c_sd) & c_sd == 0,
      ex_sd_continuity_corrected = !is.na(ex_sd) & ex_sd == 0,
      c_sd = if_else(c_sd_continuity_corrected, minimum_control_sd, c_sd),
      ex_sd = if_else(ex_sd_continuity_corrected, minimum_exposed_sd, ex_sd)
    )

  if (any(corrected$c_sd_continuity_corrected & is.na(corrected$minimum_control_sd)) ||
      any(corrected$ex_sd_continuity_corrected & is.na(corrected$minimum_exposed_sd))) {
    stop("A zero SD has no positive SD in the same treatment arm and outcome stratum.")
  }

  corrected |>
    select(-minimum_control_sd, -minimum_exposed_sd)
}

apply_nonpercentage_corrections <- function(data) {
  strata <- c("assay_type", "measurement_variable", "data_type", "data_units")

  # Calculate the smallest positive mean within each outcome stratum and arm.
  local_minima <- data |>
    group_by(across(all_of(strata))) |>
    summarise(
      minimum_control_mean = minimum_positive(c_mean),
      minimum_exposed_mean = minimum_positive(ex_mean),
      .groups = "drop"
    )

  global_control_minimum <- minimum_positive(data$c_mean)
  global_exposed_minimum <- minimum_positive(data$ex_mean)

  data |>
    left_join(local_minima, by = strata) |>
    mutate(
      c_mean_continuity_corrected = !is.na(c_mean) & c_mean == 0,
      ex_mean_continuity_corrected = !is.na(ex_mean) & ex_mean == 0,
      c_mean = if_else(
        c_mean_continuity_corrected,
        0.9 * coalesce(minimum_control_mean, global_control_minimum),
        c_mean
      ),
      ex_mean = if_else(
        ex_mean_continuity_corrected,
        0.9 * coalesce(minimum_exposed_mean, global_exposed_minimum),
        ex_mean
      )
    ) |>
    select(-minimum_control_mean, -minimum_exposed_mean) |>
    apply_zero_sd_corrections()
}

prepare_lnrr_data <- function(data) {
  # Apply the non-percentage rules.
  nonpercentage <- data |>
    filter(data_type != "percentage") |>
    apply_nonpercentage_corrections()

  # Apply the percentage rules.
  percentage <- data |>
    filter(data_type == "percentage") |>
    mutate(
      c_mean_continuity_corrected = c_mean %in% c(0, 100),
      ex_mean_continuity_corrected = ex_mean %in% c(0, 100),
      c_mean = case_when(c_mean == 0 ~ 0.5, c_mean == 100 ~ 99.5, TRUE ~ c_mean),
      ex_mean = case_when(ex_mean == 0 ~ 0.5, ex_mean == 100 ~ 99.5, TRUE ~ ex_mean)
    ) |>
    apply_zero_sd_corrections()

  # Recombine the rows in their original order.
  bind_rows(nonpercentage, percentage) |>
    arrange(match(es_id, data$es_id))
}
```

We apply the boundary corrections and report the number of changes. The second table lists only the statistics that the rules modified.

```{r}
#| label: boundary-correction-summary

db_prepared <- prepare_lnrr_data(db)

boundary_corrections <- db |>
  select(
    es_id, study_id,
    c_mean_old = c_mean,
    ex_mean_old = ex_mean,
    c_sd_old = c_sd,
    ex_sd_old = ex_sd
  ) |>
  left_join(
    db_prepared |>
      select(
        es_id,
        c_mean_new = c_mean,
        ex_mean_new = ex_mean,
        c_sd_new = c_sd,
        ex_sd_new = ex_sd,
        c_mean_continuity_corrected,
        ex_mean_continuity_corrected,
        c_sd_continuity_corrected,
        ex_sd_continuity_corrected
      ),
    by = "es_id"
  ) |>
  mutate(
    any_correction = c_mean_continuity_corrected |
      ex_mean_continuity_corrected |
      c_sd_continuity_corrected |
      ex_sd_continuity_corrected
  )

boundary_corrections |>
  summarise(
    `Total effect sizes` = n(),
    `Any correction` = sum(any_correction),
    `Control mean` = sum(c_mean_continuity_corrected),
    `Exposed mean` = sum(ex_mean_continuity_corrected),
    `Control SD` = sum(c_sd_continuity_corrected),
    `Exposed SD` = sum(ex_sd_continuity_corrected)
  ) |>
  knitr::kable(caption = "Boundary corrections by statistic.")

modified_statistics <- bind_rows(
  boundary_corrections |>
    filter(c_mean_continuity_corrected) |>
    transmute(study_id, es_id, statistic = "Control mean", original = c_mean_old, corrected = c_mean_new),
  boundary_corrections |>
    filter(ex_mean_continuity_corrected) |>
    transmute(study_id, es_id, statistic = "Exposed mean", original = ex_mean_old, corrected = ex_mean_new),
  boundary_corrections |>
    filter(c_sd_continuity_corrected) |>
    transmute(study_id, es_id, statistic = "Control SD", original = c_sd_old, corrected = c_sd_new),
  boundary_corrections |>
    filter(ex_sd_continuity_corrected) |>
    transmute(study_id, es_id, statistic = "Exposed SD", original = ex_sd_old, corrected = ex_sd_new)
) |>
  arrange(study_id, es_id, statistic)

modified_statistics |>
  knitr::kable(digits = 3, caption = "Statistics before and after boundary correction.")
```

We calculate lnRR and `lnRR_var` row by row because the formula depends on `data_type` and `comparison_structure`. We then calculate lnVR and lnCVR for the non-percentage rows. `metafor::escalc()` returns each effect size and its sampling variance.

```{r}
#| label: calculate-lnrr

calculate_lnrr <- function(data, within_subject_correlation = 0.5) {
  if (any(!data$comparison_structure %in% c("independent", "dependent"))) {
    stop("comparison_structure must be independent or dependent.")
  }

  result <- data |>
    mutate(lnRR = NA_real_, lnRR_var = NA_real_)

  arcsine_sqrt <- function(proportion) asin(sqrt(proportion))

  for (i in seq_len(nrow(result))) {
    # Read the statistics and calculation settings for one effect size.
    ex_n <- result$ex_n[i]
    c_n <- result$c_n[i]
    ex_mean <- result$ex_mean[i]
    c_mean <- result$c_mean[i]
    ex_sd <- result$ex_sd[i]
    c_sd <- result$c_sd[i]
    data_type <- result$data_type[i]
    comparison_structure <- result$comparison_structure[i]

    if (data_type != "percentage") {
      # Use metafor for outcomes on their extracted scale.
      if (comparison_structure == "independent") {
        effect <- metafor::escalc(
          measure = "ROM",
          n1i = ex_n, n2i = c_n,
          m1i = ex_mean, m2i = c_mean,
          sd1i = ex_sd, sd2i = c_sd,
          var.names = c("lnRR", "lnRR_var")
        )
      } else {
        effect <- metafor::escalc(
          measure = "ROMC",
          ni = (ex_n + c_n) / 2,
          m1i = ex_mean, m2i = c_mean,
          sd1i = ex_sd, sd2i = c_sd,
          ri = within_subject_correlation,
          var.names = c("lnRR", "lnRR_var")
        )
      }

      result$lnRR[i] <- effect$lnRR
      result$lnRR_var[i] <- effect$lnRR_var
    } else {
      # Convert percentage inputs to proportions and transform them.
      if (max(c(ex_mean, c_mean), na.rm = TRUE) > 1) {
        ex_mean <- ex_mean / 100
        c_mean <- c_mean / 100
        ex_sd <- ex_sd / 100
        c_sd <- c_sd / 100
      }

      ex_sd_transformed <- sqrt(ex_sd^2 / (4 * ex_mean * (1 - ex_mean)))
      c_sd_transformed <- sqrt(c_sd^2 / (4 * c_mean * (1 - c_mean)))
      ex_mean_transformed <- arcsine_sqrt(ex_mean)
      c_mean_transformed <- arcsine_sqrt(c_mean)

      variance_exposed <- ex_sd_transformed^2 / (ex_mean_transformed^2 * ex_n)
      variance_control <- c_sd_transformed^2 / (c_mean_transformed^2 * c_n)

      result$lnRR[i] <- log(ex_mean_transformed / c_mean_transformed)
      result$lnRR_var[i] <- if (comparison_structure == "independent") {
        variance_exposed + variance_control
      } else {
        variance_exposed + variance_control -
          2 * within_subject_correlation *
          sqrt(variance_exposed) * sqrt(variance_control)
      }
    }

    if (result$higher_better[i] == "yes") {
      # Harmonise the direction after calculating the effect size.
      result$lnRR[i] <- -result$lnRR[i]
    }
  }

  result
}

calculate_lnvr <- function(data, within_subject_correlation = 0.5) {
  # Exclude percentage outcomes before calculating variability.
  result <- data |>
    filter(data_type != "percentage") |>
    mutate(lnVR = NA_real_, lnVR_var = NA_real_)

  for (i in seq_len(nrow(result))) {
    # Use the estimator that matches the comparison structure.
    effect <- if (result$comparison_structure[i] == "independent") {
      metafor::escalc(
        measure = "VR",
        n1i = result$ex_n[i], n2i = result$c_n[i],
        sd1i = result$ex_sd[i], sd2i = result$c_sd[i],
        var.names = c("lnVR", "lnVR_var")
      )
    } else {
      metafor::escalc(
        measure = "VRC",
        ni = (result$ex_n[i] + result$c_n[i]) / 2,
        sd1i = result$ex_sd[i], sd2i = result$c_sd[i],
        ri = within_subject_correlation,
        var.names = c("lnVR", "lnVR_var")
      )
    }

    result$lnVR[i] <- effect$lnVR
    result$lnVR_var[i] <- effect$lnVR_var
  }

  result
}

calculate_lncvr <- function(data, within_subject_correlation = 0.5) {
  # Exclude percentage outcomes before calculating relative variability.
  result <- data |>
    filter(data_type != "percentage") |>
    mutate(lnCVR = NA_real_, lnCVR_var = NA_real_)

  for (i in seq_len(nrow(result))) {
    # Use the estimator that matches the comparison structure.
    effect <- if (result$comparison_structure[i] == "independent") {
      metafor::escalc(
        measure = "CVR",
        n1i = result$ex_n[i], n2i = result$c_n[i],
        m1i = result$ex_mean[i], m2i = result$c_mean[i],
        sd1i = result$ex_sd[i], sd2i = result$c_sd[i],
        var.names = c("lnCVR", "lnCVR_var")
      )
    } else {
      metafor::escalc(
        measure = "CVRC",
        ni = (result$ex_n[i] + result$c_n[i]) / 2,
        m1i = result$ex_mean[i], m2i = result$c_mean[i],
        sd1i = result$ex_sd[i], sd2i = result$c_sd[i],
        ri = within_subject_correlation,
        var.names = c("lnCVR", "lnCVR_var")
      )
    }

    result$lnCVR[i] <- effect$lnCVR
    result$lnCVR_var[i] <- effect$lnCVR_var
  }

  result
}

# Calculate the three effect-size types.
db_lnrr <- calculate_lnrr(db_prepared)
db_lnvr <- calculate_lnvr(db_prepared)
db_lncvr <- calculate_lncvr(db_prepared)

# Keep every lnRR row and join variability estimates by effect-size ID.
db_effect_sizes <- db_lnrr |>
  left_join(
    db_lnvr |>
      select(es_id, lnVR, lnVR_var),
    by = "es_id"
  ) |>
  left_join(
    db_lncvr |>
      select(es_id, lnCVR, lnCVR_var),
    by = "es_id"
  )

if (
  nrow(db_effect_sizes) != nrow(db) ||
    anyNA(db_effect_sizes$lnRR) ||
    anyNA(db_effect_sizes$lnRR_var) ||
    any(db_effect_sizes$lnRR_var <= 0)
) {
  stop("The lnRR calculation failed validation.")
}

nonpercentage <- db_effect_sizes$data_type != "percentage"
percentage <- db_effect_sizes$data_type == "percentage"

if (
  any(!is.finite(db_effect_sizes$lnVR[nonpercentage])) ||
    any(!is.finite(db_effect_sizes$lnVR_var[nonpercentage])) ||
    any(db_effect_sizes$lnVR_var[nonpercentage] <= 0) ||
    any(!is.finite(db_effect_sizes$lnCVR[nonpercentage])) ||
    any(!is.finite(db_effect_sizes$lnCVR_var[nonpercentage])) ||
    any(db_effect_sizes$lnCVR_var[nonpercentage] <= 0) ||
    any(!is.na(db_effect_sizes$lnVR[percentage])) ||
    any(!is.na(db_effect_sizes$lnCVR[percentage]))
) {
  stop("The variability effect-size calculations failed validation.")
}

db_effect_sizes |>
  summarise(
    lnRR = sum(is.finite(lnRR)),
    lnVR = sum(is.finite(lnVR)),
    lnCVR = sum(is.finite(lnCVR))
  ) |>
  tidyr::pivot_longer(
    everything(),
    names_to = "effect_size",
    values_to = "k"
  ) |>
  knitr::kable(
    col.names = c("Effect size", "k"),
    caption = "Effect sizes available for analysis."
  )
```

# Effect-size database

We add all three effect sizes and their sampling variances to the analysis data and save the result. Percentage rows retain missing values for lnVR and lnCVR. The following sections use this file.

```{r}
#| label: export-effect-sizes

readr::write_csv(
  db_effect_sizes,
  file.path(derived_dir, "db_effect_sizes.csv"),
  na = ""
)

tibble(
  file = "Data/derived/db_effect_sizes.csv",
  rows = nrow(db_effect_sizes),
  studies = n_distinct(db_effect_sizes$study_id),
  effect_sizes = n_distinct(db_effect_sizes$es_id)
) |>
  knitr::kable(caption = "Effect-size database.")
```

# Effect-size distributions

We plot each effect size and its sampling variance in a separate histogram.

```{r}
#| label: lnrr-histogram
#| fig-cap: "Distribution of lnRR effect sizes."
#| fig-width: 7
#| fig-height: 4

ggplot(db_effect_sizes, aes(x = lnRR)) +
  geom_histogram(
    bins = 30,
    boundary = 0,
    colour = "white",
    linewidth = 0.2
  ) +
  labs(
    x = "lnRR (- less, + more anxiety/depression-like behaviour)",
    y = "Count"
  ) +
  theme_minimal()
```

```{r}
#| label: lnrr-var-histogram
#| fig-cap: "Distribution of lnRR sampling variances."
#| fig-width: 7
#| fig-height: 4

ggplot(db_effect_sizes, aes(x = lnRR_var)) +
  geom_histogram(
    bins = 30,
    boundary = 0,
    colour = "white",
    linewidth = 0.2
  ) +
  labs(x = "lnRR_var", y = "Count") +
  theme_minimal()
```

```{r}
#| label: lnvr-histogram
#| fig-cap: "Distribution of lnVR effect sizes."
#| fig-width: 7
#| fig-height: 4

db_effect_sizes |>
  filter(is.finite(lnVR)) |>
  ggplot(aes(x = lnVR)) +
  geom_histogram(
    bins = 30,
    boundary = 0,
    colour = "white",
    linewidth = 0.2
  ) +
  labs(x = "lnVR (- lower, + greater absolute variability)", y = "Count") +
  theme_minimal()
```

```{r}
#| label: lnvr-var-histogram
#| fig-cap: "Distribution of lnVR sampling variances."
#| fig-width: 7
#| fig-height: 4

db_effect_sizes |>
  filter(is.finite(lnVR_var)) |>
  ggplot(aes(x = lnVR_var)) +
  geom_histogram(
    bins = 30,
    boundary = 0,
    colour = "white",
    linewidth = 0.2
  ) +
  labs(x = "lnVR_var", y = "Count") +
  theme_minimal()
```

```{r}
#| label: lncvr-histogram
#| fig-cap: "Distribution of lnCVR effect sizes."
#| fig-width: 7
#| fig-height: 4

db_effect_sizes |>
  filter(is.finite(lnCVR)) |>
  ggplot(aes(x = lnCVR)) +
  geom_histogram(
    bins = 30,
    boundary = 0,
    colour = "white",
    linewidth = 0.2
  ) +
  labs(x = "lnCVR (- lower, + greater relative variability)", y = "Count") +
  theme_minimal()
```

```{r}
#| label: lncvr-var-histogram
#| fig-cap: "Distribution of lnCVR sampling variances."
#| fig-width: 7
#| fig-height: 4

db_effect_sizes |>
  filter(is.finite(lnCVR_var)) |>
  ggplot(aes(x = lnCVR_var)) +
  geom_histogram(
    bins = 30,
    boundary = 0,
    colour = "white",
    linewidth = 0.2
  ) +
  labs(x = "lnCVR_var", y = "Count") +
  theme_minimal()
```

::: callout-note
## Session information

```{r}
#| label: session-information

sessionInfo()
```
:::
