---
title: "Dependency structure audit"
subtitle: "Step 1 before construction of the sampling VCV matrix"
---

This chapter inspects the experimental dependency information in the original extraction database. It does not construct a sampling variance-covariance matrix and does not fit a meta-analytic model.

```{r}
#| label: setup

pacman::p_load(
  dplyr,
  knitr,
  readr,
  stringr,
  tibble,
  tidyr
)

project_root <- normalizePath("..", mustWork = TRUE)
data_dir <- file.path(project_root, "Data")
derived_dir <- file.path(data_dir, "derived")

original_path <- file.path(data_dir, "PCN_data_ext_checking_v.3.csv")

original <- readr::read_csv(
  original_path,
  na = c("", "NA"),
  name_repair = "unique_quiet",
  show_col_types = FALSE
)

audit_data <- original |>
  mutate(
    study_id_raw = study_id,
    study_id = if_else(
      study_id == "`" & data_file == "whitlow_1978_thesis_fig2",
      "whitlow_1978_thesis",
      study_id
    )
  )
```

# Database structure

The original extraction file has one row per effect size. One row contains a malformed `study_id` value (`\``). Its title, data file, and all other study fields identify it as `whitlow_1978_thesis`. The correction is made only in the in-memory audit copy; the original file is not changed.

```{r}
#| label: database-summary

database_summary <- tibble(
  item = c(
    "Rows",
    "Columns",
    "Raw study_id values",
    "Studies after documented ID repair",
    "Unique effect sizes"
  ),
  value = c(
    nrow(original),
    ncol(original),
    n_distinct(original$study_id),
    n_distinct(audit_data$study_id),
    n_distinct(original$es_id)
  )
)

database_summary |>
  knitr::kable(caption = "Structure of the original extraction database.")
```

## Column inventory

The inventory reports every column, its imported R class, number of non-missing values, and number of unique non-missing values.

```{r}
#| label: column-inventory

column_inventory <- tibble(
  column_number = seq_along(original),
  column_name = names(original),
  r_class = vapply(original, function(x) class(x)[1], character(1)),
  non_missing_rows = vapply(original, function(x) sum(!is.na(x)), integer(1)),
  missing_rows = vapply(original, function(x) sum(is.na(x)), integer(1)),
  unique_non_missing_values = vapply(
    original,
    function(x) dplyr::n_distinct(x, na.rm = TRUE),
    integer(1)
  )
)

column_inventory |>
  knitr::kable(caption = "All columns in the original extraction database.")
```

# Requested-variable map

The table maps each variable required for the dependency assessment to the existing database. The definitions follow Table 5 of the protocol and the clarification that `con_id` identifies a control group within an experiment and study. Dam- and litter-level identifiers were not extraction items and are outside the planned VCV structure.

```{r}
#| label: requested-variable-map

variable_map <- tribble(
  ~requested_variable, ~existing_column, ~status, ~assessment,
  "effect_id", "es_id", "Available", "Unique in all 207 rows.",
  "study_id", "study_id", "Available after repair", "One malformed Whitlow value must be repaired; study_id is a publication identifier, not an experimental unit.",
  "experiment_id", "ex_id", "Available", "The CSV ex_id corresponds to the protocol ex_cohort_id: an independent exposed experimental cohort retained across its assays and outcomes.",
  "cohort_id", "ex_id; con_id", "Available", "ex_id identifies the exposed cohort and con_id identifies the control group within the experiment and study.",
  "dam_group_id", NA_character_, "Out of scope", "Not collected and not used in the planned VCV.",
  "litter_group_id", NA_character_, "Out of scope", "Not collected and not used in the planned VCV.",
  "treatment_group_id", "ex_id", "Available", "Stable identifier for the exposed experimental cohort across assays and outcomes.",
  "control_group_id", "con_id", "Available", "Stable identifier for the control group within an experiment and study.",
  "animal_sample_id", NA_character_, "Not required", "Individual-animal IDs were not extracted; dependence is represented at the recorded cohort/group level.",
  "sex", "offspring_sex", "Available", "Recorded for every effect size, but does not identify whether male and female offspring came from the same litters.",
  "species", "sp_latin", "Available", "Recorded for every effect size.",
  "behavioural_test", "assay_type", "Available", "Recorded for every effect size.",
  "behavioural_outcome", "out_id; measurement_variable", "Available", "out_id is the outcome identifier and measurement_variable is the reported measure.",
  "outcome_domain", "outcome_type", "Available", "Anxiety or depression domain.",
  "time_id", "offspring_age_d; measurement_timing; out_id", "Available descriptors", "Age, test timing, and outcome ID distinguish measurements; repeated outcomes from the same cohort remain linked by ex_id or con_id.",
  "yi", "lnRR in db_effect_sizes.csv", "Derived", "Created in Chapter 2; not present in the original extraction file.",
  "vi", "lnRR_var in db_effect_sizes.csv", "Derived", "Created in Chapter 2; not present in the original extraction file.",
  "treatment mean", "ex_mean", "Available/derived when missing", "Original values are completed in Chapter 1 when sufficient summary statistics are available.",
  "treatment SD", "ex_sd", "Available/derived when missing", "Original values are completed in Chapter 1.",
  "treatment sample size", "ex_n", "Available", "Recorded for every effect size.",
  "control mean", "c_mean", "Available/derived when missing", "Original values are completed in Chapter 1 when sufficient summary statistics are available.",
  "control SD", "c_sd", "Available/derived when missing", "Original values are completed in Chapter 1.",
  "control sample size", "c_n", "Available", "Recorded for every effect size.",
  "treatment information", "ex_id; gest_stage_ex; exposure_span_d; exposure_type; exposure_session_duration_h; noise_type; noise_type_2; loudness_dB; frequency_Hz", "Available", "Describes exposure but does not identify animal overlap.",
  "analysis-specific arm labels", "ex_a_id; c_a_id; comparison", "Proxy only", "Labels commonly change by outcome or time and are not stable animal-sample identifiers.",
  "within-effect comparison type", "comparison_structure", "Available", "Selects ROM or ROMC for one effect size; it does not describe pairwise dependence between different effect sizes.",
  "shared-control identifier", "con_id; shared_control", "Available", "Identical con_id values identify the shared control group; shared_control is retained as an audit flag."
)

variable_map |>
  knitr::kable(caption = "Mapping of required dependency variables to the extraction database.")
```

# Existing identifier structure

```{r}
#| label: identifier-summary

identifier_columns <- c(
  "study_id", "ex_id", "con_id", "out_id", "es_id",
  "c_a_id", "ex_a_id", "comparison", "data_file"
)

identifier_summary <- bind_rows(lapply(identifier_columns, function(id) {
  grouped <- audit_data |>
    filter(!is.na(.data[[id]])) |>
    group_by(.data[[id]]) |>
    summarise(
      rows = n(),
      studies = n_distinct(study_id),
      .groups = "drop"
    )

  tibble(
    identifier = id,
    unique_values = nrow(grouped),
    values_repeated_across_rows = sum(grouped$rows > 1),
    values_spanning_multiple_studies = sum(grouped$studies > 1)
  )
}))

identifier_summary |>
  knitr::kable(caption = "Existing identifier variables.")
```

`es_id` is unique. `ex_id` identifies the exposed experimental cohort and `con_id` identifies the control group within the experiment and study. The `ex_a_id` and `c_a_id` fields are author-provided analysis labels and are not used as cohort identifiers.

## Study-level structure

```{r}
#| label: study-level-audit

study_audit <- audit_data |>
  group_by(study_id) |>
  summarise(
    effect_sizes = n(),
    exposed_group_ids = n_distinct(ex_id),
    control_group_ids = n_distinct(con_id),
    outcome_ids = n_distinct(out_id),
    sexes = str_c(sort(unique(offspring_sex)), collapse = " | "),
    offspring_ages_d = str_c(sort(unique(offspring_age_d)), collapse = " | "),
    behavioural_tests = n_distinct(assay_type),
    behavioural_outcomes = n_distinct(measurement_variable),
    data_files = n_distinct(data_file, na.rm = TRUE),
    comparison_structures = str_c(
      sort(unique(comparison_structure)),
      collapse = " | "
    ),
    id_notes = str_c(unique(id_notes), collapse = " || "),
    .groups = "drop"
  ) |>
  arrange(study_id)

study_audit |>
  knitr::kable(caption = "Study-level identifier and outcome structure.")
```

## Arm-ID resolution

The following audit describes how each cohort or control-group ID is reused across outcomes. Different sample sizes within one ID are retained as outcome-specific availability or missingness; they do not create a new cohort unless the extraction assigns a new `ex_id` or `con_id`.

```{r}
#| label: arm-id-audit

audit_arm_ids <- function(data, arm) {
  id_column <- paste0(arm, "_id")
  n_column <- if (arm == "ex") "ex_n" else "c_n"

  data |>
    group_by(study_id, group_id = .data[[id_column]]) |>
    summarise(
      arm = if_else(arm == "ex", "Treatment", "Control"),
      effect_sizes = n(),
      distinct_sample_sizes = n_distinct(.data[[n_column]]),
      sample_sizes = str_c(sort(unique(.data[[n_column]])), collapse = " | "),
      sexes = n_distinct(offspring_sex),
      ages = n_distinct(offspring_age_d),
      behavioural_tests = n_distinct(assay_type),
      behavioural_outcomes = n_distinct(measurement_variable),
      data_files = n_distinct(data_file, na.rm = TRUE),
      cohort_or_control_group_identified = TRUE,
      sample_size_varies = distinct_sample_sizes > 1,
      .groups = "drop"
    )
}

group_id_audit <- bind_rows(
  audit_arm_ids(audit_data, "ex"),
  audit_arm_ids(audit_data, "con")
) |>
  arrange(study_id, arm, group_id)

group_id_audit |>
  filter(
    distinct_sample_sizes > 1 |
      ages > 1 |
      behavioural_tests > 1 |
      behavioural_outcomes > 1 |
      data_files > 1
  ) |>
  knitr::kable(caption = "Broad group IDs spanning more than one analysis context.")
```

# Pairwise dependency audit

The pairwise audit uses the existing IDs only to measure how much dependency is identifiable and how much remains ambiguous. It does not assign covariance values.

```{r}
#| label: pairwise-identifier-audit

pair_index <- t(utils::combn(seq_len(nrow(audit_data)), 2))

pairwise_ids <- tibble(
  row_1 = pair_index[, 1],
  row_2 = pair_index[, 2]
) |>
  mutate(
    same_study =
      audit_data$study_id[row_1] == audit_data$study_id[row_2],
    same_exposed_group =
      same_study & audit_data$ex_id[row_1] == audit_data$ex_id[row_2],
    same_control_group =
      same_study & audit_data$con_id[row_1] == audit_data$con_id[row_2],
    same_broad_arm = same_exposed_group | same_control_group,
    different_treatment_n =
      same_exposed_group & audit_data$ex_n[row_1] != audit_data$ex_n[row_2],
    different_control_n =
      same_control_group & audit_data$c_n[row_1] != audit_data$c_n[row_2],
    different_sample_size = different_treatment_n | different_control_n,
    different_age =
      same_broad_arm &
        audit_data$offspring_age_d[row_1] != audit_data$offspring_age_d[row_2],
    different_test =
      same_broad_arm & audit_data$assay_type[row_1] != audit_data$assay_type[row_2],
    different_outcome =
      same_broad_arm &
        audit_data$measurement_variable[row_1] !=
          audit_data$measurement_variable[row_2]
  )

pairwise_summary <- tibble(
  check = c(
    "All effect-size pairs",
    "Pairs from the same publication",
    "Pairs sharing an exposed cohort or control-group ID",
    "Same-publication pairs without a shared cohort/group ID",
    "Shared-group pairs with any different arm sample size",
    "Shared-group pairs with different treatment sample sizes",
    "Shared-group pairs with different control sample sizes",
    "Shared-group pairs spanning offspring ages",
    "Shared-group pairs spanning behavioural tests",
    "Shared-group pairs spanning behavioural outcomes"
  ),
  effect_size_pairs = c(
    nrow(pairwise_ids),
    sum(pairwise_ids$same_study),
    sum(pairwise_ids$same_broad_arm),
    sum(pairwise_ids$same_study & !pairwise_ids$same_broad_arm),
    sum(pairwise_ids$different_sample_size),
    sum(pairwise_ids$different_treatment_n),
    sum(pairwise_ids$different_control_n),
    sum(pairwise_ids$different_age),
    sum(pairwise_ids$different_test),
    sum(pairwise_ids$different_outcome)
  )
)

pairwise_summary |>
  knitr::kable(caption = "Pairwise resolution available from the current identifiers.")
```

# Step 1 decision

::: callout-important
## The database is sufficient for the planned cohort-level VCV matrix

The protocol defines `ex_id` as the exposed experimental cohort and `con_id` identifies the control group within the experiment and study. These identifiers provide the dependency structure required by the planned analysis:

1. Effect sizes with the same `study_id` and `ex_id` share an exposed cohort.
2. Effect sizes with the same `study_id` and `con_id` share a control group.
3. Effect sizes from the same publication that share neither `ex_id` nor `con_id` are treated as independent experimental cohorts for the sampling VCV.
4. Multiple outcomes, assays, or ages from the same cohort/group are assigned the fixed working outcome correlation `rho = 0.5` because their correlations were not reported.
5. Different sample sizes under one cohort/group ID are treated as outcome-specific availability or missingness; the recorded cohort/group ID is retained.
6. `ex_a_id` and `c_a_id` are author-provided labels used for auditing, not for defining the covariance clusters.
7. `comparison_structure` determines the within-effect calculation in Chapter 2 and is retained in the database, but it does not replace `ex_id` or `con_id` in the VCV.

Dam- and litter-level identifiers are not used. No additional sign reversal is performed during VCV construction.
:::

# Audit exports

```{r}
#| label: export-step-1-audit

readr::write_csv(
  column_inventory,
  file.path(derived_dir, "vcv_step1_column_inventory.csv"),
  na = ""
)
readr::write_csv(
  variable_map,
  file.path(derived_dir, "vcv_step1_variable_map.csv"),
  na = ""
)
readr::write_csv(
  study_audit,
  file.path(derived_dir, "vcv_step1_study_audit.csv"),
  na = ""
)
readr::write_csv(
  group_id_audit,
  file.path(derived_dir, "vcv_step1_group_id_audit.csv"),
  na = ""
)
readr::write_csv(
  pairwise_summary,
  file.path(derived_dir, "vcv_step1_pairwise_summary.csv"),
  na = ""
)
```

::: callout-note
## Session information

```{r}
#| label: session-information

sessionInfo()
```
:::
