---
title: "Sampling variance-covariance matrix"
---

We construct the sampling variance-covariance matrix from the lnRR sampling variances and the exposed- and control-group identifiers in the extraction database.

We first load the effect-size database and check the columns required to construct `V`.

::: callout-note
## Variable definitions

| Variable | Definition |
|:--|:--|
| `study_id` | Identifier for the primary research paper and the cluster argument in `vcalc()`. |
| `es_id` | Unique identifier for each effect size. |
| `ex_id` | Identifier for an exposed group within an experiment and study. |
| `con_id` | Identifier for a control group within an experiment and study. |
| `out_id` | Identifier for a behavioural outcome. |
| `ex_n`, `c_n` | Exposed- and control-group sample sizes. |
| `lnRR_var` | Sampling variance of lnRR and the diagonal value of `V`. |
| `V` | Sampling variance-covariance matrix for the lnRR estimates. |
| $\rho$ | Working correlation for multiple or repeated outcomes that share a group. |
:::

```{r}
#| label: setup

pacman::p_load(
  dplyr,
  knitr,
  metafor,
  readr,
  tibble
)

project_root <- normalizePath("..", mustWork = TRUE)
derived_dir <- file.path(project_root, "Data", "derived")

# Load the effect-size database from Chapter 2.
vcv_data <- readr::read_csv(
  file.path(derived_dir, "db_effect_sizes.csv"),
  show_col_types = FALSE
)

# List the variables required to construct V and its index.
required_columns <- c(
  "study_id", "es_id", "ex_id", "con_id", "out_id",
  "ex_n", "c_n", "lnRR", "lnRR_var"
)

if (length(setdiff(required_columns, names(vcv_data))) > 0) {
  stop("Required columns are missing from the effect-size database.")
}
if (anyDuplicated(vcv_data$es_id)) {
  stop("es_id must be unique before constructing V.")
}
if (anyNA(vcv_data[c("study_id", "es_id", "ex_id", "con_id")])) {
  stop("Study and group identifiers must be complete before constructing V.")
}
if (any(!is.finite(vcv_data$lnRR_var)) || any(vcv_data$lnRR_var <= 0)) {
  stop("lnRR_var must contain positive finite sampling variances.")
}
if (any(vcv_data$ex_n <= 0) || any(vcv_data$c_n <= 0)) {
  stop("Group sample sizes must be positive before constructing V.")
}
```

# Matrix construction

We use `metafor::vcalc()` to construct `V` [@viechtbauer2010]. We use `study_id` as the study cluster, `ex_id` and `con_id` as the contrasted-group identifiers, and `ex_n` and `c_n` as the group weights. We supply each `es_id` as a distinct observation. For multiple or repeated outcomes that share an exposed or control group, we use a working correlation of $\rho = 0.5$. We assign zero sampling covariance to effect sizes that share neither group.

```{r}
#| label: construct-vcv

# Set the fixed working correlation.
working_rho <- 0.5

# Construct the sampling variance-covariance matrix.
V <- metafor::vcalc(
  vi = lnRR_var,
  cluster = study_id,
  grp1 = ex_id,
  grp2 = con_id,
  w1 = ex_n,
  w2 = c_n,
  obs = es_id,
  rho = working_rho,
  checkpd = TRUE,
  data = vcv_data
)

# Label the matrix rows and columns with es_id.
dimnames(V) <- list(vcv_data$es_id, vcv_data$es_id)
```

# Matrix checks

We check the diagonal, finite values, symmetry, dependency pattern, and positive definiteness of `V` before exporting it.

```{r}
#| label: check-vcv

# Identify effect-size pairs from the same study that share either group.
same_study <- outer(vcv_data$study_id, vcv_data$study_id, `==`)
shared_exposed_group <- same_study &
  outer(vcv_data$ex_id, vcv_data$ex_id, `==`)
shared_control_group <- same_study &
  outer(vcv_data$con_id, vcv_data$con_id, `==`)
shared_group <- shared_exposed_group | shared_control_group
diag(shared_group) <- FALSE

# Identify effect-size pairs with non-zero covariance.
nonzero_covariance <- abs(V) > sqrt(.Machine$double.eps)
diag(nonzero_covariance) <- FALSE

# Use the smallest eigenvalue to check positive definiteness.
minimum_eigenvalue <- min(eigen(V, symmetric = TRUE, only.values = TRUE)$values)

vcv_checks <- tibble(
  check = c(
    "Matrix rows",
    "Matrix columns",
    "Diagonal equals lnRR_var",
    "Finite matrix entries",
    "Symmetric matrix",
    "Positive-definite matrix",
    "Effect-size pairs with non-zero covariance",
    "Non-zero covariance pairs without a shared group"
  ),
  value = c(
    as.character(nrow(V)),
    as.character(ncol(V)),
    as.character(max(abs(diag(V) - vcv_data$lnRR_var)) < sqrt(.Machine$double.eps)),
    as.character(all(is.finite(V))),
    as.character(isSymmetric(V)),
    as.character(minimum_eigenvalue > 0),
    as.character(sum(upper.tri(V) & nonzero_covariance)),
    as.character(sum(upper.tri(V) & nonzero_covariance & !shared_group))
  )
)

vcv_checks |>
  knitr::kable(caption = "Checks of the sampling variance-covariance matrix.")
```

We display the first eight rows and columns to show the matrix structure.

```{r}
#| label: display-vcv

V[seq_len(8), seq_len(8)] |>
  round(6) |>
  knitr::kable(caption = "First eight rows and columns of V.")
```

# Output files

We save the matrix as CSV and RDS files. We also save an index that records the effect-size order in the rows and columns of `V`.

```{r}
#| label: export-vcv

# Create the index for the rows and columns of V.
vcv_index <- vcv_data |>
  transmute(
    matrix_row = row_number(),
    study_id,
    es_id,
    ex_id,
    con_id,
    out_id,
    ex_n,
    c_n,
    lnRR,
    lnRR_var
  )

vcv_matrix_export <- as.data.frame(V, check.names = FALSE) |>
  tibble::rownames_to_column("es_id")

# Export the index and both matrix formats.
readr::write_csv(
  vcv_index,
  file.path(derived_dir, "PCN_vcv_index.csv"),
  na = ""
)
readr::write_csv(
  vcv_matrix_export,
  file.path(derived_dir, "PCN_vcv_matrix.csv"),
  na = ""
)
saveRDS(V, file.path(derived_dir, "PCN_vcv_matrix.rds"))

tibble(
  file = c(
    "Data/derived/PCN_vcv_index.csv",
    "Data/derived/PCN_vcv_matrix.csv",
    "Data/derived/PCN_vcv_matrix.rds"
  ),
  contents = c(
    "Effect-size order and identifiers",
    "Sampling variance-covariance matrix",
    "R matrix for modelling"
  )
) |>
  knitr::kable(caption = "Sampling variance-covariance matrix files.")
```

::: callout-note
## Session information

```{r}
#| label: session-information

sessionInfo()
```
:::
