---
title: "Sampling variance-covariance matrices"
---

We construct one sampling variance-covariance matrix for each effect-size type. Each matrix uses the corresponding sampling variances and the exposed- and control-group identifiers in the extraction database.

We first load the effect-size database and check the columns required to construct the matrices.

::: callout-note
## Variable definitions

| Variable | Definition |
|:--|:--|
| `study_id` | Identifier for the primary research paper and the cluster argument in `vcalc()`. |
| `es_id` | Unique identifier for each effect size. |
| `ex_id` | Identifier for an exposed group within an experiment and study. |
| `con_id` | Identifier for a control group within an experiment and study. |
| `out_id` | Identifier for a behavioural outcome. |
| `ex_n`, `c_n` | Exposed- and control-group sample sizes. |
| `lnRR_var`, `lnVR_var`, `lnCVR_var` | Sampling variances and the diagonal values of the corresponding matrices. |
| `V` | Sampling variance-covariance matrix for one effect-size type. |
| $\rho$ | Working correlation for multiple or repeated outcomes that share a group. |
:::

```{r}
#| label: setup

pacman::p_load(
  dplyr,
  knitr,
  metafor,
  purrr,
  readr,
  tibble,
  tidyr
)

project_root <- normalizePath("..", mustWork = TRUE)
derived_dir <- file.path(project_root, "Data", "derived")

# Load the effect-size database from Chapter 2.
vcv_data <- readr::read_csv(
  file.path(derived_dir, "db_effect_sizes.csv"),
  show_col_types = FALSE
)

# List the variables required to construct V and its index.
required_columns <- c(
  "study_id", "es_id", "ex_id", "con_id", "out_id",
  "ex_n", "c_n",
  "lnRR", "lnRR_var", "lnVR", "lnVR_var", "lnCVR", "lnCVR_var"
)

if (length(setdiff(required_columns, names(vcv_data))) > 0) {
  stop("Required columns are missing from the effect-size database.")
}
if (anyDuplicated(vcv_data$es_id)) {
  stop("es_id must be unique before constructing V.")
}
if (anyNA(vcv_data[c("study_id", "es_id", "ex_id", "con_id")])) {
  stop("Study and group identifiers must be complete before constructing V.")
}
if (any(vcv_data$ex_n <= 0) || any(vcv_data$c_n <= 0)) {
  stop("Group sample sizes must be positive before constructing V.")
}

# Link each effect size to its sampling-variance column.
effect_specs <- tibble::tribble(
  ~effect, ~variance,
  "lnRR",  "lnRR_var",
  "lnVR",  "lnVR_var",
  "lnCVR", "lnCVR_var"
)
```

# Matrix construction

We use `metafor::vcalc()` to construct each matrix [@viechtbauer2010]. We use `study_id` as the study cluster, `ex_id` and `con_id` as the contrasted-group identifiers, and `ex_n` and `c_n` as the group weights. We supply each `es_id` as a distinct observation. For multiple or repeated outcomes that share an exposed or control group, we use a working correlation of $\rho = 0.5$. We assign zero sampling covariance to effect sizes that share neither group.

The lnRR matrix contains every effect-size row. The lnVR and lnCVR matrices contain only non-percentage outcomes because Chapter 2 does not calculate variability metrics for percentage outcomes.

```{r}
#| label: construct-vcv

# Set the fixed working correlation.
working_rho <- 0.5

# Construct one matrix and retain the matching data rows.
construct_vcv <- function(data, effect, variance, rho = 0.5) {
  # Retain rows with a complete effect size and sampling variance.
  idx <- which(is.finite(data[[effect]]) & is.finite(data[[variance]]))
  metric_data <- data[idx, , drop = FALSE]

  if (any(metric_data[[variance]] <= 0)) {
    stop(variance, " must contain positive sampling variances.")
  }

  # Construct V from the group identifiers and sampling variances.
  V <- metafor::vcalc(
    vi = metric_data[[variance]],
    cluster = metric_data$study_id,
    grp1 = metric_data$ex_id,
    grp2 = metric_data$con_id,
    w1 = metric_data$ex_n,
    w2 = metric_data$c_n,
    obs = metric_data$es_id,
    rho = rho,
    checkpd = TRUE
  )

  # Label the rows and columns with the retained effect-size IDs.
  dimnames(V) <- list(metric_data$es_id, metric_data$es_id)

  list(effect = effect, variance = variance, data = metric_data, V = V)
}

# Apply the same matrix construction to lnRR, lnVR, and lnCVR.
vcv_results <- purrr::map2(
  effect_specs$effect,
  effect_specs$variance,
  ~ construct_vcv(vcv_data, .x, .y, rho = working_rho)
)
names(vcv_results) <- effect_specs$effect
```

# Matrix checks

We check the diagonal, finite values, symmetry, dependency pattern, and positive definiteness of each matrix before exporting it.

```{r}
#| label: check-vcv

check_vcv <- function(result) {
  data <- result$data
  V <- result$V

  # Identify pairs from the same study that share either group.
  same_study <- outer(data$study_id, data$study_id, `==`)
  shared_exposed_group <- same_study & outer(data$ex_id, data$ex_id, `==`)
  shared_control_group <- same_study & outer(data$con_id, data$con_id, `==`)
  shared_group <- shared_exposed_group | shared_control_group
  diag(shared_group) <- FALSE

  # Identify off-diagonal non-zero sampling covariances.
  nonzero_covariance <- abs(V) > sqrt(.Machine$double.eps)
  diag(nonzero_covariance) <- FALSE

  tibble(
    effect_size = result$effect,
    k = nrow(V),
    diagonal_matches = max(abs(diag(V) - data[[result$variance]])) <
      sqrt(.Machine$double.eps),
    finite = all(is.finite(V)),
    symmetric = isSymmetric(V),
    positive_definite = min(eigen(V, symmetric = TRUE, only.values = TRUE)$values) > 0,
    nonzero_covariance_pairs = sum(upper.tri(V) & nonzero_covariance),
    nonzero_pairs_without_shared_group = sum(
      upper.tri(V) & nonzero_covariance & !shared_group
    )
  )
}

vcv_checks <- purrr::map_dfr(vcv_results, check_vcv)

vcv_checks |>
  knitr::kable(caption = "Checks of the sampling variance-covariance matrix.")
```

We display the first eight rows and columns of each matrix to show its structure.

```{r}
#| label: display-vcv
#| results: asis

for (result in vcv_results) {
  n_display <- min(8, nrow(result$V))
  cat("\n\n## ", result$effect, "\n\n", sep = "")
  cat(
    knitr::kable(
      round(result$V[seq_len(n_display), seq_len(n_display)], 6),
      caption = paste("First rows and columns of the", result$effect, "matrix.")
    ),
    sep = "\n"
  )
  cat("\n\n")
}
```

# Output files

We save each matrix as CSV and RDS files. We also save an index that records the effect-size order in the rows and columns of each matrix.

```{r}
#| label: export-vcv

# Define file names for the three sets of outputs. We retain the existing
# lnRR names so the overall-effect chapter remains compatible.
vcv_files <- tribble(
  ~effect, ~index_file, ~csv_file, ~rds_file,
  "lnRR",  "PCN_vcv_index.csv",       "PCN_vcv_matrix.csv",       "PCN_vcv_matrix.rds",
  "lnVR",  "PCN_lnVR_vcv_index.csv",  "PCN_lnVR_vcv_matrix.csv",  "PCN_lnVR_vcv_matrix.rds",
  "lnCVR", "PCN_lnCVR_vcv_index.csv", "PCN_lnCVR_vcv_matrix.csv", "PCN_lnCVR_vcv_matrix.rds"
)

for (i in seq_len(nrow(vcv_files))) {
  result <- vcv_results[[vcv_files$effect[i]]]

  # Record the row order and the effect-size values used in this matrix.
  vcv_index <- result$data |>
    transmute(
      matrix_row = row_number(),
      study_id,
      es_id,
      ex_id,
      con_id,
      out_id,
      ex_n,
      c_n,
      effect_size = .data[[result$effect]],
      sampling_variance = .data[[result$variance]]
    )

  vcv_matrix_export <- as.data.frame(result$V, check.names = FALSE) |>
    tibble::rownames_to_column("es_id")

  readr::write_csv(
    vcv_index,
    file.path(derived_dir, vcv_files$index_file[i]),
    na = ""
  )
  readr::write_csv(
    vcv_matrix_export,
    file.path(derived_dir, vcv_files$csv_file[i]),
    na = ""
  )
  saveRDS(result$V, file.path(derived_dir, vcv_files$rds_file[i]))
}

vcv_files |>
  tidyr::pivot_longer(
    cols = c(index_file, csv_file, rds_file),
    names_to = "format",
    values_to = "file"
  ) |>
  mutate(file = file.path("Data", "derived", file)) |>
  select(effect, format, file) |>
  knitr::kable(
    col.names = c("Effect size", "File type", "File"),
    caption = "Sampling variance-covariance matrix files."
  )
```

::: callout-note
## Session information

```{r}
#| label: session-information

sessionInfo()
```
:::
