---
title: "Sampling variance-covariance matrix"
---

The sampling variance-covariance matrix was constructed from the lnRR sampling variances and the recorded exposed- and control-group identifiers.

```{r}
#| label: setup

pacman::p_load(
  dplyr,
  knitr,
  metafor,
  readr,
  tibble
)

project_root <- normalizePath("..", mustWork = TRUE)
derived_dir <- file.path(project_root, "Data", "derived")

vcv_data <- readr::read_csv(
  file.path(derived_dir, "db_effect_sizes.csv"),
  show_col_types = FALSE
)

required_columns <- c(
  "study_id", "es_id", "ex_id", "con_id", "out_id",
  "ex_n", "c_n", "lnRR", "lnRR_var"
)

if (length(setdiff(required_columns, names(vcv_data))) > 0) {
  stop("Required columns are missing from the effect-size database.")
}
if (anyDuplicated(vcv_data$es_id)) {
  stop("es_id must be unique before constructing V.")
}
if (anyNA(vcv_data[c("study_id", "es_id", "ex_id", "con_id")])) {
  stop("Study and group identifiers must be complete before constructing V.")
}
if (any(!is.finite(vcv_data$lnRR_var)) || any(vcv_data$lnRR_var <= 0)) {
  stop("lnRR_var must contain positive finite sampling variances.")
}
if (any(vcv_data$ex_n <= 0) || any(vcv_data$c_n <= 0)) {
  stop("Group sample sizes must be positive before constructing V.")
}
```

# Matrix construction

`metafor::vcalc()` was used to construct `V` [@viechtbauer2010]. `study_id` defines the study cluster, `ex_id` and `con_id` identify the contrasted groups, and `ex_n` and `c_n` provide the group weights. Each `es_id` is supplied as a distinct observation. A working correlation of $\rho = 0.5$ is used for multiple or repeated outcomes that share an exposed or control group. Effect sizes that share neither group have sampling covariance equal to zero.

```{r}
#| label: construct-vcv

working_rho <- 0.5

V <- metafor::vcalc(
  vi = lnRR_var,
  cluster = study_id,
  grp1 = ex_id,
  grp2 = con_id,
  w1 = ex_n,
  w2 = c_n,
  obs = es_id,
  rho = working_rho,
  checkpd = TRUE,
  data = vcv_data
)

dimnames(V) <- list(vcv_data$es_id, vcv_data$es_id)
```

# Matrix checks

The diagonal, symmetry, dependency pattern, and positive definiteness of `V` were checked before export.

```{r}
#| label: check-vcv

same_study <- outer(vcv_data$study_id, vcv_data$study_id, `==`)
shared_exposed_group <- same_study &
  outer(vcv_data$ex_id, vcv_data$ex_id, `==`)
shared_control_group <- same_study &
  outer(vcv_data$con_id, vcv_data$con_id, `==`)
shared_group <- shared_exposed_group | shared_control_group
diag(shared_group) <- FALSE

nonzero_covariance <- abs(V) > sqrt(.Machine$double.eps)
diag(nonzero_covariance) <- FALSE

minimum_eigenvalue <- min(eigen(V, symmetric = TRUE, only.values = TRUE)$values)

vcv_checks <- tibble(
  check = c(
    "Matrix rows",
    "Matrix columns",
    "Diagonal equals lnRR_var",
    "Finite matrix entries",
    "Symmetric matrix",
    "Positive-definite matrix",
    "Effect-size pairs with non-zero covariance",
    "Non-zero covariance pairs without a shared group"
  ),
  value = c(
    as.character(nrow(V)),
    as.character(ncol(V)),
    as.character(max(abs(diag(V) - vcv_data$lnRR_var)) < sqrt(.Machine$double.eps)),
    as.character(all(is.finite(V))),
    as.character(isSymmetric(V)),
    as.character(minimum_eigenvalue > 0),
    as.character(sum(upper.tri(V) & nonzero_covariance)),
    as.character(sum(upper.tri(V) & nonzero_covariance & !shared_group))
  )
)

vcv_checks |>
  knitr::kable(caption = "Checks of the sampling variance-covariance matrix.")
```

```{r}
#| label: display-vcv

V[seq_len(8), seq_len(8)] |>
  round(6) |>
  knitr::kable(caption = "First eight rows and columns of V.")
```

# Output files

The matrix is saved as CSV and RDS files. The index file records the effect-size order used for the rows and columns of `V`.

```{r}
#| label: export-vcv

vcv_index <- vcv_data |>
  transmute(
    matrix_row = row_number(),
    study_id,
    es_id,
    ex_id,
    con_id,
    out_id,
    ex_n,
    c_n,
    lnRR,
    lnRR_var
  )

vcv_matrix_export <- as.data.frame(V, check.names = FALSE) |>
  tibble::rownames_to_column("es_id")

readr::write_csv(
  vcv_index,
  file.path(derived_dir, "PCN_vcv_index.csv"),
  na = ""
)
readr::write_csv(
  vcv_matrix_export,
  file.path(derived_dir, "PCN_vcv_matrix.csv"),
  na = ""
)
saveRDS(V, file.path(derived_dir, "PCN_vcv_matrix.rds"))

tibble(
  file = c(
    "Data/derived/PCN_vcv_index.csv",
    "Data/derived/PCN_vcv_matrix.csv",
    "Data/derived/PCN_vcv_matrix.rds"
  ),
  contents = c(
    "Effect-size order and identifiers",
    "Sampling variance-covariance matrix",
    "R matrix used for modelling"
  )
) |>
  knitr::kable(caption = "Files created for the sampling variance-covariance matrix.")
```

::: callout-note
## Session information

```{r}
#| label: session-information

sessionInfo()
```
:::
