---
title: "Sampling dependency structure"
---

Effect sizes within a publication may share an exposed cohort, a control group, or both. The identifiers used to represent these sampling dependencies are described below.

```{r}
#| label: setup

pacman::p_load(
  dplyr,
  knitr,
  readr,
  stringr,
  tibble,
  tidyr
)

project_root <- normalizePath("..", mustWork = TRUE)
data_dir <- file.path(project_root, "Data")
derived_dir <- file.path(data_dir, "derived")

original_path <- file.path(data_dir, "PCN_data_ext_checking_v.3.csv")

original <- readr::read_csv(
  original_path,
  na = c("", "NA"),
  name_repair = "unique_quiet",
  show_col_types = FALSE
)

dependency_data <- original
```

# Data structure

The extraction database contains one row per effect size.

```{r}
#| label: database-summary

database_summary <- tibble(
  item = c(
    "Rows",
    "Columns",
    "Studies",
    "Unique effect sizes"
  ),
  value = c(
    nrow(original),
    ncol(original),
    n_distinct(original$study_id),
    n_distinct(original$es_id)
  )
)

database_summary |>
  knitr::kable(caption = "Structure of the original extraction database.")
```

## Column inventory

The table lists the imported class, completeness, and number of unique values for each column in the original extraction file.

```{r}
#| label: column-inventory

column_inventory <- tibble(
  column_number = seq_along(original),
  column_name = names(original),
  r_class = vapply(original, function(x) class(x)[1], character(1)),
  non_missing_rows = vapply(original, function(x) sum(!is.na(x)), integer(1)),
  missing_rows = vapply(original, function(x) sum(is.na(x)), integer(1)),
  unique_non_missing_values = vapply(
    original,
    function(x) dplyr::n_distinct(x, na.rm = TRUE),
    integer(1)
  )
)

column_inventory |>
  knitr::kable(caption = "Columns in the original extraction database.")
```

# Variable definitions

The variables used to identify effect sizes, experimental groups, outcomes, and sampling statistics are defined below.

```{r}
#| label: variable-definitions

variable_definitions <- tribble(
  ~analysis_variable, ~database_column, ~definition,
  "Effect size", "es_id", "Unique identifier for each effect size.",
  "Study", "study_id", "Identifier for the primary publication.",
  "Exposed cohort", "ex_id", "Identifier for an exposed experimental cohort retained across its assays and outcomes.",
  "Control group", "con_id", "Identifier for a control group within an experiment and study.",
  "Outcome", "out_id; measurement_variable", "Outcome identifier and reported behavioural measure.",
  "Outcome domain", "outcome_type", "Anxiety- or depression-like behaviour.",
  "Behavioural test", "assay_type", "Behavioural test used to measure the outcome.",
  "Measurement time", "offspring_age_d; measurement_timing", "Offspring age and timing of behavioural measurement.",
  "Sex", "offspring_sex", "Sex of the assessed offspring.",
  "Species", "sp_latin", "Species of the assessed offspring.",
  "Treatment statistics", "ex_mean; ex_sd; ex_n", "Mean, standard deviation, and sample size for the exposed group.",
  "Control statistics", "c_mean; c_sd; c_n", "Mean, standard deviation, and sample size for the control group.",
  "Effect-size estimate", "lnRR", "Log response ratio calculated in Chapter 2.",
  "Sampling variance", "lnRR_var", "Sampling variance of the log response ratio calculated in Chapter 2.",
  "Comparison structure", "comparison_structure", "Specifies an independent or correlated comparison within one effect size.",
  "Shared-control flag", "shared_control", "Indicates comparisons that use a shared control group.",
  "Reported group labels", "ex_a_id; c_a_id; comparison", "Labels for reported comparisons; not used as cohort identifiers."
)

variable_definitions |>
  knitr::kable(caption = "Variables used for the sampling dependency structure.")
```

# Experimental identifiers

The code below summarizes the number and reuse of the identifiers present in the extraction database.

```{r}
#| label: identifier-summary

identifier_columns <- c(
  "study_id", "ex_id", "con_id", "out_id", "es_id",
  "c_a_id", "ex_a_id", "comparison", "data_file"
)

identifier_summary <- bind_rows(lapply(identifier_columns, function(id) {
  grouped <- dependency_data |>
    filter(!is.na(.data[[id]])) |>
    group_by(.data[[id]]) |>
    summarise(
      rows = n(),
      studies = n_distinct(study_id),
      .groups = "drop"
    )

  tibble(
    identifier = id,
    unique_values = nrow(grouped),
    values_repeated_across_rows = sum(grouped$rows > 1),
    values_spanning_multiple_studies = sum(grouped$studies > 1)
  )
}))

identifier_summary |>
  knitr::kable(caption = "Identifier variables in the extraction database.")
```

`es_id` identifies one effect size. Within each study, `ex_id` identifies the exposed cohort and `con_id` identifies the control group. The fields `ex_a_id` and `c_a_id` are retained as reported group labels.

## Study structure

The number of exposed cohorts, control groups, outcomes, tests, and measurement ages is summarized for each study.

```{r}
#| label: study-structure

study_structure <- dependency_data |>
  group_by(study_id) |>
  summarise(
    effect_sizes = n(),
    exposed_group_ids = n_distinct(ex_id),
    control_group_ids = n_distinct(con_id),
    outcome_ids = n_distinct(out_id),
    sexes = str_c(sort(unique(offspring_sex)), collapse = " | "),
    offspring_ages_d = str_c(sort(unique(offspring_age_d)), collapse = " | "),
    behavioural_tests = n_distinct(assay_type),
    behavioural_outcomes = n_distinct(measurement_variable),
    data_files = n_distinct(data_file, na.rm = TRUE),
    comparison_structures = str_c(
      sort(unique(comparison_structure)),
      collapse = " | "
    ),
    .groups = "drop"
  ) |>
  arrange(study_id)

study_structure |>
  knitr::kable(caption = "Experimental and outcome structure by study.")
```

## Exposed cohorts and control groups

Group identifiers were retained across outcomes, behavioural tests, and measurement ages. Differences in sample size within a group identifier were retained as recorded for each outcome.

```{r}
#| label: group-structure

summarize_group_ids <- function(data, arm) {
  id_column <- paste0(arm, "_id")
  n_column <- if (arm == "ex") "ex_n" else "c_n"

  data |>
    group_by(study_id, group_id = .data[[id_column]]) |>
    summarise(
      group = if_else(arm == "ex", "Exposed", "Control"),
      effect_sizes = n(),
      distinct_sample_sizes = n_distinct(.data[[n_column]]),
      sample_sizes = str_c(sort(unique(.data[[n_column]])), collapse = " | "),
      sexes = n_distinct(offspring_sex),
      ages = n_distinct(offspring_age_d),
      behavioural_tests = n_distinct(assay_type),
      behavioural_outcomes = n_distinct(measurement_variable),
      data_files = n_distinct(data_file, na.rm = TRUE),
      sample_size_varies = distinct_sample_sizes > 1,
      .groups = "drop"
    )
}

group_structure <- bind_rows(
  summarize_group_ids(dependency_data, "ex"),
  summarize_group_ids(dependency_data, "con")
) |>
  arrange(study_id, group, group_id)

group_structure |>
  filter(
    distinct_sample_sizes > 1 |
      ages > 1 |
      behavioural_tests > 1 |
      behavioural_outcomes > 1 |
      data_files > 1
  ) |>
  knitr::kable(
    caption = "Group identifiers used across more than one analysis context."
  )
```

# Pairwise dependency structure

Pairs of effect sizes were classified according to shared study, exposed-cohort, and control-group identifiers. This step does not assign covariance values.

```{r}
#| label: pairwise-dependency-structure

pair_index <- t(utils::combn(seq_len(nrow(dependency_data)), 2))

pairwise_ids <- tibble(
  row_1 = pair_index[, 1],
  row_2 = pair_index[, 2]
) |>
  mutate(
    same_study =
      dependency_data$study_id[row_1] == dependency_data$study_id[row_2],
    same_exposed_group =
      same_study & dependency_data$ex_id[row_1] == dependency_data$ex_id[row_2],
    same_control_group =
      same_study & dependency_data$con_id[row_1] == dependency_data$con_id[row_2],
    same_group = same_exposed_group | same_control_group,
    different_treatment_n =
      same_exposed_group & dependency_data$ex_n[row_1] != dependency_data$ex_n[row_2],
    different_control_n =
      same_control_group & dependency_data$c_n[row_1] != dependency_data$c_n[row_2],
    different_sample_size = different_treatment_n | different_control_n,
    different_age =
      same_group &
        dependency_data$offspring_age_d[row_1] != dependency_data$offspring_age_d[row_2],
    different_test =
      same_group & dependency_data$assay_type[row_1] != dependency_data$assay_type[row_2],
    different_outcome =
      same_group &
        dependency_data$measurement_variable[row_1] !=
          dependency_data$measurement_variable[row_2]
  )

pairwise_summary <- tibble(
  pair_classification = c(
    "All effect-size pairs",
    "Same publication",
    "Shared exposed cohort or control group",
    "Same publication without a shared cohort or control group",
    "Shared group with different sample sizes",
    "Shared exposed cohort with different treatment sample sizes",
    "Shared control group with different control sample sizes",
    "Shared group spanning offspring ages",
    "Shared group spanning behavioural tests",
    "Shared group spanning behavioural outcomes"
  ),
  effect_size_pairs = c(
    nrow(pairwise_ids),
    sum(pairwise_ids$same_study),
    sum(pairwise_ids$same_group),
    sum(pairwise_ids$same_study & !pairwise_ids$same_group),
    sum(pairwise_ids$different_sample_size),
    sum(pairwise_ids$different_treatment_n),
    sum(pairwise_ids$different_control_n),
    sum(pairwise_ids$different_age),
    sum(pairwise_ids$different_test),
    sum(pairwise_ids$different_outcome)
  )
)

pairwise_summary |>
  knitr::kable(caption = "Pairwise sampling dependency structure.")
```

# Dependency rules for VCV construction

The sampling variance-covariance matrix will use the following rules:

1. Effect sizes with the same `study_id` and `ex_id` share an exposed cohort.
2. Effect sizes with the same `study_id` and `con_id` share a control group.
3. Effect sizes within a publication that share neither identifier have sampling covariance equal to zero.
4. For multiple outcomes, tests, or ages measured in the same group, an unreported outcome correlation of $\rho = 0.5$ is used.
5. The recorded group identifier is retained when outcome-specific sample sizes differ.
6. `comparison_structure` specifies the calculation of an individual effect size; it is not used as a dependency identifier.
7. No additional direction reversal is applied during VCV construction.

# Supplementary data files

The tables generated in this chapter are saved as comma-separated files.

```{r}
#| label: export-supplementary-tables

readr::write_csv(
  column_inventory,
  file.path(derived_dir, "dependency_column_inventory.csv"),
  na = ""
)
readr::write_csv(
  variable_definitions,
  file.path(derived_dir, "dependency_variable_definitions.csv"),
  na = ""
)
readr::write_csv(
  study_structure,
  file.path(derived_dir, "dependency_study_structure.csv"),
  na = ""
)
readr::write_csv(
  group_structure,
  file.path(derived_dir, "dependency_group_structure.csv"),
  na = ""
)
readr::write_csv(
  pairwise_summary,
  file.path(derived_dir, "dependency_pairwise_summary.csv"),
  na = ""
)
```

::: callout-note
## Session information

```{r}
#| label: session-information

sessionInfo()
```
:::
