---
title: "Overall effect"
---

## Model

This chapter fits one multilevel, intercept-only meta-analytic model to estimate the overall effect of maternal chronic noise exposure on offspring anxiety- and depression-like behaviour, using the harmonized lnRR (`lnrr_harm`) from Chapter 2. A positive overall effect means more anxiety-/depression-like behaviour under exposure; a negative effect means less.

Several effect sizes are not independent: some share a control group, and several studies contribute more than one outcome per study. Two things are used together to account for this, following standard practice for multilevel meta-analyses [@nakagawa2023b]:

- a variance-covariance (VCV) matrix built with `metafor::vcalc()`, assuming a correlation of $\rho = 0.5$ between sampling errors of effect sizes that share a control group (`dependency_cluster_id`, defined in Chapter 1) - 0.5 is a conventional working value used when the true within-cluster correlation is not estimable from the data, not an estimated parameter;
- random intercepts for `study_id` (between-study heterogeneity), `es_id` (residual/within-study heterogeneity), and `strain` (heterogeneity attributable to rodent strain).

```{r}
#| label: setup

if (!requireNamespace("pacman", quietly = TRUE)) {
  stop(
    'Package "pacman" is required. Install it once with ',
    'install.packages("pacman").'
  )
}

analysis_packages <- c("dplyr", "readr", "metafor", "knitr")
pacman::p_load(char = analysis_packages)

if (!requireNamespace("orchaRd", quietly = TRUE)) {
  if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
  remotes::install_github("daniel1noble/orchaRd", force = FALSE)
}
library(orchaRd)

options(dplyr.summarise.inform = FALSE)
knitr::opts_chunk$set(comment = "#>")

project_root <- normalizePath("..", mustWork = TRUE)
derived_dir <- file.path(project_root, "Data", "derived")

es <- readr::read_csv(file.path(derived_dir, "db_effect_sizes.csv"), show_col_types = FALSE) %>%
  mutate(across(where(is.character), as.factor)) %>%
  mutate(across(c(lnrr_harm, lnrr_var), as.numeric))
```

```{r}
#| label: vcv

VCV <- metafor::vcalc(
  vi = lnrr_var,
  cluster = dependency_cluster_id,
  obs = es_id,
  rho = 0.5,
  data = es
)
```

```{r}
#| label: model-overall

overall_model <- metafor::rma.mv(
  yi = lnrr_harm,
  V = VCV,
  random = list(~1 | study_id, ~1 | es_id, ~1 | strain),
  test = "t",
  method = "REML",
  sparse = TRUE,
  data = es
)

summary(overall_model)
```

## Heterogeneity

```{r}
#| label: heterogeneity

orchaRd::i2_ml(overall_model)
```

$I^2$ is partitioned across the three random-effects levels (study, effect size, strain); the total is the proportion of variance in observed effect sizes not attributable to sampling error.

## Orchard plot

```{r}
#| label: orchard-overall
#| fig-width: 6
#| fig-height: 4

orchard_overall <- orchaRd::orchard_plot(
  overall_model,
  group = "study_id",
  xlab = "lnRR (harmonized)",
  flip = TRUE
) +
  ggplot2::scale_y_discrete(labels = c("1" = "Overall effect", "All" = "Overall effect"))

orchard_overall
```

Each small point is one effect size, sized by precision (inverse sampling variance); the large point and its interval are the model estimate and its 95% confidence and prediction intervals.

## Export

```{r}
#| label: export

readr::write_rds(overall_model, file.path(derived_dir, "overall_model_lnRR.rds"))
tibble::tibble(output = "Data/derived/overall_model_lnRR.rds") %>% knitr::kable()
```

## Session information

```{r}
#| label: session-information

sessionInfo()
```
