---
title: "Overall effect"
---

This chapter fits a multilevel, intercept-only meta-analytic model for lnRR.

`comparison_structure` was already used to calculate each sampling variance in Chapter 2. It is not used again in the variance-covariance matrix. The matrix instead uses `cohort_id`, `es_id`, and an assumed sampling-error correlation of `rho = 0.5`.

```{r}
#| label: setup
#| message: false
#| warning: false

pacman::p_load(
  dplyr,
  ggplot2,
  knitr,
  metafor,
  orchaRd,
  readr,
  tibble
)

project_root <- normalizePath("..", mustWork = TRUE)
derived_dir <- file.path(project_root, "Data", "derived")

db <- readr::read_csv(
  file.path(derived_dir, "db_effect_sizes.csv"),
  show_col_types = FALSE
) |>
  mutate(across(where(is.character), as.factor)) |>
  mutate(across(c(lnRR, lnRR_var), as.numeric))
```

# Variance-covariance matrix

Multiple outcomes from the same cohort are assigned an assumed sampling-error correlation of `rho = 0.5`. The cohort identifier is `study_id` plus `ex_id`, created in Chapter 1.

```{r}
#| label: variance-covariance-matrix

VCV <- metafor::vcalc(
  vi = lnRR_var,
  cluster = cohort_id,
  obs = es_id,
  rho = 0.5,
  data = db
)
```

# Multilevel model

The model uses random intercepts for study, effect size, and strain. Estimation uses REML and `test = "t"`.

```{r}
#| label: overall-model

overall_model <- metafor::rma.mv(
  yi = lnRR,
  V = VCV,
  random = list(
    ~ 1 | study_id,
    ~ 1 | es_id,
    ~ 1 | strain
  ),
  test = "t",
  method = "REML",
  sparse = TRUE,
  data = db
)

summary(overall_model)
```

# Heterogeneity

```{r}
#| label: heterogeneity

orchaRd::i2_ml(overall_model)
```

# Orchard plot

```{r}
#| label: orchard-plot
#| fig-width: 6
#| fig-height: 4

overall_plot <- orchaRd::orchard_plot(
  overall_model,
  group = "study_id",
  xlab = "lnRR",
  flip = TRUE
) +
  scale_x_discrete(labels = "Overall effect")

overall_plot
```

# Export

```{r}
#| label: export-model

readr::write_rds(
  overall_model,
  file.path(derived_dir, "overall_model_lnRR.rds")
)

tibble(output = "Data/derived/overall_model_lnRR.rds") |>
  knitr::kable()
```

# Session information

```{r}
#| label: session-information

sessionInfo()
```
