---
title: "Overall effect"
---

After estimating the effect sizes and constructing their sampling variance-covariance matrix, we fit one multilevel, intercept-only meta-analytic model to estimate the overall mean effect of prenatal noise exposure on lnRR.

```{r}
#| label: setup
#| include: false

pacman::p_load(
  dplyr,
  ggplot2,
  knitr,
  metafor,
  orchaRd,
  readr,
  tibble
)

project_root <- normalizePath("..", mustWork = TRUE)
derived_dir <- file.path(project_root, "Data", "derived")
```

# Analysis data

We first load the effect-size database, the sampling variance-covariance matrix (`V`), and the file that records the row order of `V`.

::: callout-note
## Variable definitions

| Variable | Definition |
|:--|:--|
| `study_id` | Identifier for the primary research paper. |
| `es_id` | Unique identifier for each effect size. |
| `strain` | Rodent strain in the experiment. |
| `lnRR` | Bias-corrected log response ratio. |
| `lnRR_var` | Sampling variance of lnRR. |
| `V` | Sampling variance-covariance matrix for the lnRR estimates. |
:::

```{r}
#| label: load-model-data

# Load the effect-size database.
db <- readr::read_csv(
  file.path(derived_dir, "db_effect_sizes.csv"),
  show_col_types = FALSE
)

# Load V and its effect-size index.
V <- readRDS(file.path(derived_dir, "PCN_vcv_matrix.rds"))
vcv_index <- readr::read_csv(
  file.path(derived_dir, "PCN_vcv_index.csv"),
  show_col_types = FALSE
)

# Check that the database and V contain the same effect sizes in the same order.
if (
  nrow(db) != nrow(V) ||
    ncol(V) != nrow(V) ||
    !identical(as.character(db$es_id), as.character(vcv_index$es_id))
) {
  stop("The effect-size database and VCV matrix are not aligned.")
}

# Check that the diagonal of V contains the lnRR sampling variances.
if (max(abs(diag(V) - db$lnRR_var)) >= sqrt(.Machine$double.eps)) {
  stop("The diagonal of V does not match lnRR_var.")
}
```

# Multilevel intercept-only meta-analytic model

We use `metafor::rma.mv()` to fit the model [@viechtbauer2010]. We provide the lnRR estimates as the response and the matrix from the preceding chapter as `V`. We include random intercepts for study (`study_id`), effect size (`es_id`), and strain (`strain`). We use restricted maximum likelihood estimation and `test = "t"`.

```{r}
#| label: fit-overall-lnrr-model

# Fit the intercept-only multilevel model.
overall_lnRR_model <- metafor::rma.mv(
  yi = lnRR,
  V = V,
  random = list(
    ~ 1 | study_id,
    ~ 1 | es_id,
    ~ 1 | strain
  ),
  test = "t",
  method = "REML",
  sparse = TRUE,
  data = db
)

# Save the fitted model for the following analyses.
saveRDS(
  overall_lnRR_model,
  file.path(derived_dir, "PCN_overall_lnRR_model.rds")
)
```

## lnRR

We print the complete model summary, including the overall lnRR estimate, its standard error, confidence interval, heterogeneity test, and variance components.

```{r}
#| label: summary-overall-lnrr

summary(overall_lnRR_model)
```

We calculate multilevel $I^2$ for the total heterogeneity and for each random-effect level.

```{r}
#| label: i2-overall-lnrr

orchaRd::i2_ml(overall_lnRR_model)
```

We use an orchard plot to display the effect sizes and the overall model estimate. Point size represents precision, and the horizontal interval shows the confidence interval around the overall estimate.

```{r}
#| label: orchard-overall-lnrr
#| fig-cap: "Overall lnRR from the multilevel intercept-only meta-analysis."
#| fig-width: 6
#| fig-height: 4

p_lnRR <- orchaRd::orchard_plot(
  overall_lnRR_model,
  group = "study_id",
  xlab = "lnRR",
  flip = TRUE
) +
  ggplot2::scale_x_discrete(labels = "Overall effect")

p_lnRR
```

# Output file

We save the fitted model as an R object so that the following chapters can use the same model specification.

```{r}
#| label: overall-model-output

tibble(
  file = "Data/derived/PCN_overall_lnRR_model.rds",
  contents = "Intercept-only multilevel lnRR model"
) |>
  knitr::kable(caption = "Model object for the following chapters.")
```

::: callout-note
## Session information

```{r}
#| label: session-information

sessionInfo()
```
:::
