---
title: "Overall effects"
---

After estimating effect sizes, we fit three multilevel, intercept-only meta-analytic models to estimate the overall effect of prenatal noise exposure on:

- the mean outcome (lnRR),
- the absolute variability (lnVR),
- the relative variability (lnCVR).

Because studies can contribute multiple dependent outcomes, we use `metafor::rma.mv()` with the sampling variance-covariance matrices constructed in **Sampling variance-covariance matrices**.

```{r}
#| label: setup
#| include: false

pacman::p_load(
  dplyr,
  ggplot2,
  knitr,
  metafor,
  orchaRd,
  purrr,
  readr,
  tibble
)

# Keep the rendered output concise and use a consistent code-output prefix.
knitr::opts_chunk$set(comment = "#>")

# Define the project folders used in this section.
project_root <- normalizePath("..", mustWork = TRUE)
derived_dir <- file.path(project_root, "Data", "derived")
```

# Variance-covariance matrices (shared-group dependence)

We load the effect-size database and the matching sampling variance-covariance matrix (`V`) for each effect-size type. **Sampling variance-covariance matrices** uses `metafor::vcalc()` to represent dependence from shared prenatal-noise and control groups. It uses a working correlation of $\rho = 0.5$ for multiple or repeated outcomes that share a group. Each matrix index records the effect-size order in the rows and columns of `V`.

::: callout-note
## Variable definitions

| Variable | Definition |
|:--|:--|
| `study_id` | Identifier for the primary research paper. |
| `es_id` | Unique identifier for each effect size. |
| `strain` | Rodent strain in the experiment. |
| `lnRR` | Log response ratio, which measures differences in group means. |
| `lnVR` | Log variability ratio, which measures absolute variability. |
| `lnCVR` | Log coefficient of variation ratio, which measures variability relative to the group mean. |
| `lnRR_var`, `lnVR_var`, `lnCVR_var` | Sampling variance of the corresponding effect size. |
| `V` | Sampling variance-covariance matrix for one effect-size type. |
| $k$ | Number of effect sizes included in a model. |
:::

**Direction conventions:** lnRR values below zero represent less anxiety- or depression-like behaviour, while values above zero represent more. For lnVR and lnCVR, values below zero represent lower variability and values above zero represent greater variability in the prenatal-noise group.

```{r}
#| label: load-model-data

# Load the database created in Effect size calculation.
db <- readr::read_csv(
  file.path(derived_dir, "db_effect_sizes.csv"),
  show_col_types = FALSE
)

# Link each effect-size type to its sampling variance, VCV matrix, and index.
metric_specs <- tribble(
  ~effect, ~variance, ~matrix_file, ~index_file, ~axis_label,
  "lnRR",  "lnRR_var",  "PCN_vcv_matrix.rds",       "PCN_vcv_index.csv",       "lnRR",
  "lnVR",  "lnVR_var",  "PCN_lnVR_vcv_matrix.rds",  "PCN_lnVR_vcv_index.csv",  "lnVR",
  "lnCVR", "lnCVR_var", "PCN_lnCVR_vcv_matrix.rds", "PCN_lnCVR_vcv_index.csv", "lnCVR"
)

load_metric_input <- function(effect, variance, matrix_file, index_file) {
  # Load the matrix and its row-order index from Sampling variance-covariance matrices.
  V <- readRDS(file.path(derived_dir, matrix_file))
  index <- readr::read_csv(
    file.path(derived_dir, index_file),
    show_col_types = FALSE
  )

  # Arrange the effect-size database in the order recorded by the index.
  positions <- match(index$es_id, db$es_id)
  if (anyNA(positions)) {
    stop("The ", effect, " index contains an unknown es_id.")
  }
  metric_data <- db[positions, , drop = FALSE]

  # Confirm that the data, index, and matrix describe the same effect sizes.
  if (
    nrow(metric_data) != nrow(V) ||
      ncol(V) != nrow(V) ||
      !identical(as.character(metric_data$es_id), as.character(index$es_id))
  ) {
    stop("The ", effect, " data and VCV matrix are not aligned.")
  }

  # Confirm that every model row has a valid effect size and sampling variance.
  if (
    any(!is.finite(metric_data[[effect]])) ||
      any(!is.finite(metric_data[[variance]])) ||
      any(metric_data[[variance]] <= 0)
  ) {
    stop("The ", effect, " model data contain invalid values.")
  }

  # Confirm that the matrix diagonal contains the corresponding variances.
  if (
    max(abs(diag(V) - metric_data[[variance]])) >=
      sqrt(.Machine$double.eps)
  ) {
    stop("The diagonal of the ", effect, " matrix does not match ", variance, ".")
  }

  list(data = metric_data, V = V)
}

# Load and validate the three model datasets and matrices.
metric_inputs <- purrr::pmap(
  metric_specs[, c("effect", "variance", "matrix_file", "index_file")],
  load_metric_input
)
names(metric_inputs) <- metric_specs$effect

# Report the number of effect sizes available for each overall model.
metric_specs |>
  transmute(
    `Effect size` = effect,
    k = purrr::map_int(effect, ~ nrow(metric_inputs[[.x]]$data))
  ) |>
  knitr::kable(caption = "Effect sizes included in the overall models.")
```

The lnRR model includes every effect size. The lnVR and lnCVR models exclude percentage outcomes because the mean and variance of a proportion are mathematically dependent. They do not provide independent information, so we do not calculate separate variability effect sizes for these outcomes.

# Multilevel intercept-only meta-analytic models

We use `metafor::rma.mv()` to fit each model [@viechtbauer2010]. We provide one effect-size column as the response and its matching matrix as `V`. We use the same specification for all three models:

- random intercepts for study (`study_id`),
- random intercepts for effect size (`es_id`),
- random intercepts for strain (`strain`),
- REML estimation and `test = "t"`.

```{r}
#| label: fit-overall-models

fit_overall_model <- function(input, effect) {
  # Fit one intercept-only model with the matching data and VCV matrix.
  metafor::rma.mv(
    yi = input$data[[effect]],
    V = input$V,
    random = list(
      ~ 1 | study_id,
      ~ 1 | es_id,
      ~ 1 | strain
    ),
    test = "t",
    method = "REML",
    sparse = TRUE,
    data = input$data
  )
}

# Apply the same model specification to lnRR, lnVR, and lnCVR.
overall_models <- purrr::map2(
  metric_inputs,
  metric_specs$effect,
  fit_overall_model
)
names(overall_models) <- metric_specs$effect

overall_lnRR_model <- overall_models$lnRR
overall_lnVR_model <- overall_models$lnVR
overall_lnCVR_model <- overall_models$lnCVR

# Save each fitted model for the following analyses.
saveRDS(
  overall_lnRR_model,
  file.path(derived_dir, "PCN_overall_lnRR_model.rds")
)
saveRDS(
  overall_lnVR_model,
  file.path(derived_dir, "PCN_overall_lnVR_model.rds")
)
saveRDS(
  overall_lnCVR_model,
  file.path(derived_dir, "PCN_overall_lnCVR_model.rds")
)
```

For each effect-size type, we print the complete model summary and calculate multilevel $I^2$. We then use an orchard plot to display the effect sizes and the overall model estimate. Point size represents precision, and the horizontal interval shows the confidence interval around the overall estimate.

```{r}
#| label: overall-plot-function
#| include: false

plot_overall_model <- function(model, axis_label) {
  # Draw one orchard plot with the direction stated on the effect-size axis.
  orchaRd::orchard_plot(
    model,
    group = "study_id",
    xlab = axis_label,
    flip = TRUE
  ) +
    ggplot2::scale_x_discrete(labels = "Overall effect")
}
```

:::::::::: columns
::::: {.column width="49%"}
::: panel-text
## lnRR
:::

```{r}
#| label: summary-overall-lnrr

# Print the complete lnRR model results.
summary(overall_lnRR_model)
```

```{r}
#| label: i2-overall-lnrr

# Calculate total and level-specific multilevel I-squared.
orchaRd::i2_ml(overall_lnRR_model)
```

::: card
```{r}
#| label: orchard-overall-lnrr
#| fig-cap: "Overall lnRR from the multilevel intercept-only meta-analysis."
#| fig-width: 6
#| fig-height: 4

plot_overall_model(
  overall_lnRR_model,
  metric_specs$axis_label[metric_specs$effect == "lnRR"]
)
```
:::
:::::

::: {.column width="2%"}
:::

::::: {.column width="49%"}
::: panel-text
## lnVR
:::

```{r}
#| label: summary-overall-lnvr

# Print the complete lnVR model results.
summary(overall_lnVR_model)
```

```{r}
#| label: i2-overall-lnvr

# Calculate total and level-specific multilevel I-squared.
orchaRd::i2_ml(overall_lnVR_model)
```

::: card
```{r}
#| label: orchard-overall-lnvr
#| fig-cap: "Overall lnVR from the multilevel intercept-only meta-analysis."
#| fig-width: 6
#| fig-height: 4

plot_overall_model(
  overall_lnVR_model,
  metric_specs$axis_label[metric_specs$effect == "lnVR"]
)
```
:::
:::::
::::::::::

:::::::: columns
::: {.column width="25%"}
:::

::::: {.column width="50%"}
::: panel-text
## lnCVR
:::

```{r}
#| label: summary-overall-lncvr

# Print the complete lnCVR model results.
summary(overall_lnCVR_model)
```

```{r}
#| label: i2-overall-lncvr

# Calculate total and level-specific multilevel I-squared.
orchaRd::i2_ml(overall_lnCVR_model)
```

::: card
```{r}
#| label: orchard-overall-lncvr
#| fig-cap: "Overall lnCVR from the multilevel intercept-only meta-analysis."
#| fig-width: 6
#| fig-height: 4

plot_overall_model(
  overall_lnCVR_model,
  metric_specs$axis_label[metric_specs$effect == "lnCVR"]
)
```
:::
:::::

::: {.column width="25%"}
:::
::::::::

::: callout-note
## Session information

```{r}
#| label: session-information

sessionInfo()
```
:::
