---
title: "Publication bias"
---

## Approach

Following the protocol, publication bias is assessed with a funnel plot and an Egger-style regression test [@nakagawa2022]. Because lnRR and its sampling variance are mathematically linked (the variance depends on the same means used in the numerator), regressing directly on the standard error can itself create a spurious asymmetry. Following the recommendation in Nakagawa et al. [@nakagawa2022], the Egger-type test below uses a sample-size-based precision proxy, $\sqrt{1/n_{ex} + 1/n_c}$, as the moderator in the same multilevel model used for the overall effect, rather than the standard error itself.

Both diagnostics are exploratory: heterogeneity is expected to be substantial and many effect sizes are not independent, both of which make classic publication-bias tests less reliable than in a simple two-group design. They are reported for transparency and to flag whether a sensitivity analysis is warranted, not as a definitive test.

```{r}
#| label: setup

if (!requireNamespace("pacman", quietly = TRUE)) {
  stop(
    'Package "pacman" is required. Install it once with ',
    'install.packages("pacman").'
  )
}

analysis_packages <- c("dplyr", "readr", "metafor", "knitr", "ggplot2")
pacman::p_load(char = analysis_packages)

if (!requireNamespace("orchaRd", quietly = TRUE)) {
  if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
  remotes::install_github("daniel1noble/orchaRd", force = FALSE)
}
library(orchaRd)

options(dplyr.summarise.inform = FALSE)
knitr::opts_chunk$set(comment = "#>")

project_root <- normalizePath("..", mustWork = TRUE)
derived_dir <- file.path(project_root, "Data", "derived")

es <- readr::read_csv(file.path(derived_dir, "db_effect_sizes.csv"), show_col_types = FALSE) %>%
  mutate(across(where(is.character), as.factor)) %>%
  mutate(across(c(lnrr_harm, lnrr_var), as.numeric))

VCV <- metafor::vcalc(
  vi = lnrr_var,
  cluster = dependency_cluster_id,
  obs = es_id,
  rho = 0.5,
  data = es
)

overall_model <- metafor::rma.mv(
  yi = lnrr_harm,
  V = VCV,
  random = list(~1 | study_id, ~1 | es_id, ~1 | strain),
  test = "t",
  method = "REML",
  sparse = TRUE,
  data = es
)
```

## Funnel plot

```{r}
#| label: funnel-plot
#| fig-width: 7
#| fig-height: 5.5

metafor::funnel(
  overall_model,
  yaxis = "seinv",
  las = 1,
  digits = c(2, 1),
  xlab = "lnRR (harmonized)",
  ylab = "Precision (1 / SE)",
  pch = 21,
  bg = scales::alpha("black", 0.25),
  col = "black",
  cex = 1.1
)
```

A funnel plot centred asymmetrically around the overall estimate, with a gap among small/imprecise studies, is the classic visual signature of publication bias - but with only 16 studies and substantial expected heterogeneity, some asymmetry is also consistent with genuine between-study variation. Read this plot alongside the regression test below, not on its own.

## Egger-type regression

```{r}
#| label: egger-regression

es_egger <- es %>%
  mutate(precision_proxy = sqrt((1 / ex_n) + (1 / c_n)))

VCV_egger <- metafor::vcalc(
  vi = lnrr_var,
  cluster = dependency_cluster_id,
  obs = es_id,
  rho = 0.5,
  data = es_egger
)

egger_model <- metafor::rma.mv(
  yi = lnrr_harm,
  V = VCV_egger,
  mods = ~ 1 + precision_proxy,
  random = list(~1 | study_id, ~1 | es_id, ~1 | strain),
  test = "t",
  method = "REML",
  sparse = TRUE,
  data = es_egger
)

summary(egger_model)
```

A significant slope on `precision_proxy` (p < 0.05) is read as evidence of small-study effects: effect sizes from smaller samples (larger `precision_proxy`) departing systematically from those with larger samples.

```{r}
#| label: egger-bubble-plot
#| fig-width: 6
#| fig-height: 4.5

orchaRd::bubble_plot(
  egger_model,
  mod = "precision_proxy",
  group = "study_id",
  xlab = "Square root of (1/n_exposed + 1/n_control)",
  ylab = "lnRR (harmonized)"
)
```

::: {.callout-note}
If this test flags evidence of small-study effects, the next step is a sensitivity analysis (e.g., trim-and-fill is not appropriate for multilevel models; refitting the overall and moderator models excluding the smallest/most imprecise effect sizes, or excluding one study at a time, is more appropriate here and can be added once this diagnostic result is reviewed).
:::

## Session information

```{r}
#| label: session-information

sessionInfo()
```
