---
title: "Uni-moderators"
---

We fit one uni-moderator multilevel meta-regression for each study characteristic. Every model uses lnRR as the response and the same random-effects structure as the overall-effect model: study (`study_id`), effect size (`es_id`), and strain (`strain`). We construct the complete VCV matrix with `rho = 0.5`. For categorical moderators, we remove levels with fewer than five effect sizes ($k < 5$) and subset the corresponding rows and columns of the VCV matrix.

::: callout-note
## Variable definitions

| Variable | Definition |
|:--|:--|
| $k$ | Number of effect sizes in a model. |
| `V` | Sampling variance-covariance matrix for the lnRR estimates. |
| $\rho$ | Working correlation for multiple or repeated outcomes that share a group. |
| `QM` | Omnibus test statistic for the moderator. |
| Marginal $R^2$ | Proportion of model variance associated with the moderator. |
| Conditional $R^2$ | Proportion of model variance associated with the moderator and random effects. |
:::

```{r}
#| label: setup
#| include: false

pacman::p_load(
  dplyr,
  ggplot2,
  knitr,
  metafor,
  multcomp,
  orchaRd,
  purrr,
  readr,
  tibble
)

# Keep the rendered output concise and use a consistent code-output prefix.
options(dplyr.summarise.inform = FALSE)
knitr::opts_chunk$set(comment = "#>")

# Define the project folders used in this chapter.
project_root <- normalizePath("..", mustWork = TRUE)
derived_dir <- file.path(project_root, "Data", "derived")
```

# Analysis data and VCV matrix

We load the effect-size database from Chapter 2 and construct one complete VCV matrix using the same code as Chapter 3.

```{r}
#| label: load-effect-size-data

# Load the effect-size database created in Chapter 2.
db <- readr::read_csv(
  file.path(derived_dir, "db_effect_sizes.csv"),
  show_col_types = FALSE
)

# List the variables required to construct V and fit the models.
required_columns <- c(
  "study_id", "es_id", "ex_id", "con_id", "strain",
  "ex_n", "c_n", "lnRR", "lnRR_var"
)

# Stop if any required variable is missing.
if (length(setdiff(required_columns, names(db))) > 0) {
  stop("The effect-size database does not contain all model variables.")
}

# Confirm that each row has a unique effect-size identifier.
if (anyDuplicated(db$es_id)) {
  stop("es_id must identify one effect-size row.")
}

# Construct the complete sampling VCV matrix once.
VCV_lnRR <- metafor::vcalc(
  vi = lnRR_var,
  cluster = study_id,
  grp1 = ex_id,
  grp2 = con_id,
  w1 = ex_n,
  w2 = c_n,
  obs = es_id,
  rho = 0.5,
  data = db
)
```

# Moderators

The moderator list follows the biological and methodological domains in the protocol. We use the column names from the analysis-ready database.

```{r}
#| label: moderator-list

moderators <- tribble(
  ~variable,                       ~label,                                  ~kind,          ~missing_label,
  "sp_latin",                     "Species",                               "categorical",  NA_character_,
  "gest_stage_ex_standardised",   "Gestational exposure stage",            "categorical",  NA_character_,
  "exposure_type",                "Exposure type",                         "categorical",  NA_character_,
  "noise_type_2",                 "Noise type",                            "categorical",  "unclear",
  "control_conditions",           "Control conditions",                    "categorical",  NA_character_,
  "offspring_sex",                "Offspring sex",                         "categorical",  NA_character_,
  "outcome_type",                 "Outcome type",                          "categorical",  NA_character_,
  "assay_type",                   "Behavioural assay",                     "categorical",  NA_character_,
  "offspring_age_d",              "Offspring age at testing (days)",       "continuous",   NA_character_,
  "loudness_dB",                  "Loudness in dB",                         "continuous",   NA_character_
)

# Stop if the database does not contain one of the planned moderators.
if (length(setdiff(moderators$variable, names(db))) > 0) {
  stop("The effect-size database does not contain all moderators.")
}

# Display the moderator names used in the models.
moderators |>
  dplyr::select(variable, label, kind) |>
  knitr::kable(
    col.names = c("Database variable", "Moderator", "Type"),
    caption = "Uni-moderators in the analysis."
  )
```

For `noise_type_2`, we treat `unclear` as a missing acoustic-band value rather than a moderator level. We retain the recorded categories for the other categorical moderators. For `loudness_dB`, we retain unweighted dB values. Barzegar et al. (2014) reported 95 dBA, so we record its loudness value as missing because dBA and unweighted dB are not directly comparable.

The `gest_stage_ex_standardised` variable groups the recorded exposure windows as `early_mid`, `full_gestation`, `mid_late`, and `late`.

# Model functions

## Prepare the moderator data

For a categorical moderator, we remove levels with $k < 5$ and subset the same rows and columns from `V`. For a continuous moderator, we retain finite values and subset `V` in the same way.

```{r}
#| label: prepare-moderator-data

prepare_categorical <- function(data, V, moderator, missing_label = NA_character_, min_k = 5) {
  # Start with rows that contain a recorded moderator value.
  usable <- !is.na(data[[moderator]])

  # Treat a specified text label as missing when required.
  if (!is.na(missing_label)) {
    usable <- usable & as.character(data[[moderator]]) != missing_label
  }

  # Count the effect sizes in each moderator level.
  level_counts <- data[usable, ] |>
    count(level = .data[[moderator]], name = "effect_sizes") |>
    mutate(
      studies = vapply(
        level,
        function(x) n_distinct(data$study_id[usable & data[[moderator]] == x]),
        integer(1)
      ),
      include = effect_sizes >= min_k
    )

  # Retain levels represented by at least five effect sizes.
  keep_levels <- level_counts |>
    filter(include) |>
    pull(level)

  # Use the same row positions to subset the data and VCV matrix.
  idx <- which(usable & data[[moderator]] %in% keep_levels)
  dat <- data[idx, , drop = FALSE]
  V_subset <- V[idx, idx, drop = FALSE]

  # Remove unused factor levels after filtering.
  dat[[moderator]] <- droplevels(factor(dat[[moderator]]))

  # A categorical moderator requires at least two retained levels.
  if (nlevels(dat[[moderator]]) < 2) {
    stop("Fewer than two levels remain for ", moderator, ".")
  }

  list(data = dat, V = V_subset, level_counts = level_counts)
}

prepare_continuous <- function(data, V, moderator) {
  # Retain rows with finite moderator values.
  idx <- which(is.finite(data[[moderator]]))

  # Subset the data and VCV matrix with the same row positions.
  dat <- data[idx, , drop = FALSE]
  V_subset <- V[idx, idx, drop = FALSE]

  # Store the moderator as a numeric variable.
  dat[[moderator]] <- as.numeric(dat[[moderator]])

  # A continuous moderator requires at least two distinct values.
  if (n_distinct(dat[[moderator]]) < 2) {
    stop("The moderator does not contain at least two values: ", moderator, ".")
  }

  list(data = dat, V = V_subset, level_counts = NULL)
}
```

## Fit one uni-moderator model

The function below prepares one moderator, subsets the complete VCV matrix, and fits its model. For categorical moderators with at least three retained levels, it also calculates unadjusted Tukey contrasts with a separate no-intercept model.

```{r}
#| label: fit-one-unimoderator

fit_unimoderator <- function(
  data,
  V,
  moderator,
  label,
  kind,
  missing_label = NA_character_,
  min_k = 5
) {
  # Apply the appropriate preparation rule for the moderator type.
  prepared <- if (kind == "categorical") {
    prepare_categorical(data, V, moderator, missing_label, min_k)
  } else {
    prepare_continuous(data, V, moderator)
  }

  model_data <- prepared$data
  V_subset <- prepared$V

  # Fit one multilevel meta-regression with the subsetted VCV matrix.
  model <- metafor::rma.mv(
    yi = lnRR,
    V = V_subset,
    mods = reformulate(moderator),
    random = list(
      ~ 1 | study_id,
      ~ 1 | es_id,
      ~ 1 | strain
    ),
    test = "t",
    method = "REML",
    sparse = TRUE,
    data = model_data
  )

  # Draw an orchard plot for a categorical moderator. We draw continuous
  # moderators with metafor::regplot() when we display the model results.
  model_plot <- if (kind == "categorical") {
    orchaRd::orchard_plot(
      model,
      mod = moderator,
      group = "study_id",
      xlab = "lnRR",
      flip = TRUE,
      angle = 0,
      legend.pos = "top.left"
    )
  } else {
    NULL
  }

  contrasts <- NULL

  # Calculate unadjusted Tukey contrasts when at least three levels remain.
  if (kind == "categorical" && nlevels(model_data[[moderator]]) >= 3) {
    # Fit a no-intercept model so that each coefficient represents one level.
    no_intercept_model <- metafor::rma.mv(
      yi = lnRR,
      V = V_subset,
      mods = reformulate(moderator, intercept = FALSE),
      random = list(
        ~ 1 | study_id,
        ~ 1 | es_id,
        ~ 1 | strain
      ),
      test = "t",
      method = "REML",
      sparse = TRUE,
      data = model_data
    )

    # Create all pairwise Tukey contrasts among the retained levels.
    contrast_matrix <- multcomp::contrMat(
      table(model_data[[moderator]]),
      type = "Tukey"
    )
    colnames(contrast_matrix) <- names(coef(no_intercept_model))

    # Report the unadjusted p-value for each contrast.
    contrasts <- summary(
      multcomp::glht(no_intercept_model, linfct = contrast_matrix),
      test = multcomp::adjusted("none")
    )
  }

  list(
    variable = moderator,
    label = label,
    kind = kind,
    data = model_data,
    V = V_subset,
    level_counts = prepared$level_counts,
    model = model,
    r2 = orchaRd::r2_ml(model),
    plot = model_plot,
    contrasts = contrasts
  )
}
```

# Fit all uni-moderator models

We call `fit_unimoderator()` once for each row of the moderator table. The function subsets the complete VCV matrix to match the effect sizes retained for each moderator.

```{r}
#| label: fit-all-unimoderators

# Apply the same function to each moderator in the table.
unimoderator_results <- purrr::pmap(
  moderators,
  function(variable, label, kind, missing_label) {
    fit_unimoderator(
      data = db,
      V = VCV_lnRR,
      moderator = variable,
      label = label,
      kind = kind,
      missing_label = missing_label,
      min_k = 5
    )
  }
)

# Name each result with its database variable.
names(unimoderator_results) <- moderators$variable
```

# Model results

For each moderator, we print the model summary and multilevel $R^2$. We use orchard plots for categorical moderators and `metafor::regplot()` for continuous moderators. For categorical moderators with three or more levels, we also print the unadjusted pairwise contrasts.

```{r}
#| label: display-unimoderator-results
#| results: asis
#| echo: false
#| fig-width: 8
#| fig-height: 5

# Print the same set of outputs for each uni-moderator model.
for (result in unimoderator_results) {
  cat("\n\n## ", result$label, "\n\n", sep = "")

  cat("### Model summary\n\n```\n")
  print(summary(result$model))
  cat("```\n\n")

  cat("### Multilevel R-squared\n\n")
  cat(
    knitr::kable(
      as.data.frame(t(result$r2)),
      digits = 4,
      caption = paste("Multilevel R-squared for", result$label)
    ),
    sep = "\n"
  )
  cat("\n\n")

  cat("### Plot\n\n")

  if (result$kind == "categorical") {
    print(result$plot)
  } else {
    # Plot the marginal regression line and its 95% confidence interval.
    metafor::regplot(
      result$model,
      mod = result$variable,
      pred = TRUE,
      ci = TRUE,
      pi = FALSE,
      xlab = result$label,
      ylab = "lnRR",
      psize = "seinv",
      plim = c(0.7, 2.2),
      refline = 0
    )
  }

  cat("\n\n")

  # Print pairwise contrasts only when the function created them.
  if (!is.null(result$contrasts)) {
    cat("### Pairwise contrasts\n\n")
    cat("We report unadjusted Tukey contrasts between the retained levels.\n\n```\n")
    print(result$contrasts)
    cat("```\n\n")
  }
}
```

::: callout-note
## Session information

```{r}
#| label: session-information

sessionInfo()
```
:::
