---
title: "Uni-moderators"
---

We fit one uni-moderator multilevel meta-regression for each study characteristic and effect-size type. The models use lnRR, lnVR, or lnCVR as the response and the same random-effects structure as the overall-effect model: study (`study_id`), effect size (`es_id`), and strain (`strain`). We construct a complete VCV matrix for each effect-size type with `rho = 0.5`. For categorical moderators, we remove levels with fewer than five effect sizes ($k < 5$) and subset the corresponding rows and columns of the matching VCV matrix.

::: callout-note
## Variable definitions

| Variable | Definition |
|:--|:--|
| $k$ | Number of effect sizes in a model. |
| `lnRR` | Log response ratio, which measures differences in group means. |
| `lnVR` | Log variability ratio, which measures absolute variability. |
| `lnCVR` | Log coefficient of variation ratio, which measures variability relative to the group mean. |
| `V` | Sampling variance-covariance matrix for one effect-size type. |
| $\rho$ | Working correlation for multiple or repeated outcomes that share a group. |
| `QM` | Omnibus test statistic for the moderator. |
| Marginal $R^2$ | Proportion of model variance associated with the moderator. |
| Conditional $R^2$ | Proportion of model variance associated with the moderator and random effects. |
:::

Across every lnRR plot, negative values represent less anxiety- or depression-like behaviour and positive values represent more anxiety- or depression-like behaviour. For lnVR and lnCVR, negative values represent lower variability and positive values represent greater variability in the treatment group.

```{r}
#| label: setup
#| include: false

pacman::p_load(
  dplyr,
  ggplot2,
  knitr,
  metafor,
  multcomp,
  orchaRd,
  purrr,
  readr,
  tibble
)

# Keep the rendered output concise and use a consistent code-output prefix.
options(dplyr.summarise.inform = FALSE)
knitr::opts_chunk$set(comment = "#>")

# Define the project folders used in this section.
project_root <- normalizePath("..", mustWork = TRUE)
derived_dir <- file.path(project_root, "Data", "derived")
```

We load the database created in **Effect size calculation** and construct a complete VCV matrix for lnRR, lnVR, and lnCVR using the same group identifiers as **Sampling variance-covariance matrices**. The lnVR and lnCVR datasets exclude percentage outcomes because their effect-size values are missing by design.

```{r}
#| label: load-effect-size-data

# Load the database created in Effect size calculation.
db <- readr::read_csv(
  file.path(derived_dir, "db_effect_sizes.csv"),
  show_col_types = FALSE
)

# List the variables required to construct V and fit the models.
required_columns <- c(
  "study_id", "es_id", "ex_id", "con_id", "strain",
  "ex_n", "c_n",
  "lnRR", "lnRR_var", "lnVR", "lnVR_var", "lnCVR", "lnCVR_var"
)

# Stop if any required variable is missing.
if (length(setdiff(required_columns, names(db))) > 0) {
  stop("The effect-size database does not contain all model variables.")
}

# Confirm that each row has a unique effect-size identifier.
if (anyDuplicated(db$es_id)) {
  stop("es_id must identify one effect-size row.")
}

# Link each effect-size column to its sampling variance and display label.
effect_specs <- tribble(
  ~effect, ~variance, ~label, ~axis_label,
  "lnRR",  "lnRR_var",  "lnRR",  "lnRR",
  "lnVR",  "lnVR_var",  "lnVR",  "lnVR",
  "lnCVR", "lnCVR_var", "lnCVR", "lnCVR"
)

construct_metric_data <- function(data, effect, variance, rho = 0.5) {
  # Retain only rows available for this effect-size type.
  idx <- which(is.finite(data[[effect]]) & is.finite(data[[variance]]))
  metric_data <- data[idx, , drop = FALSE]

  # Construct the matching complete VCV matrix.
  V <- metafor::vcalc(
    vi = metric_data[[variance]],
    cluster = metric_data$study_id,
    grp1 = metric_data$ex_id,
    grp2 = metric_data$con_id,
    w1 = metric_data$ex_n,
    w2 = metric_data$c_n,
    obs = metric_data$es_id,
    rho = rho
  )

  list(data = metric_data, V = V)
}

# Create the three analysis datasets and complete matrices.
metric_inputs <- purrr::map2(
  effect_specs$effect,
  effect_specs$variance,
  ~ construct_metric_data(db, .x, .y, rho = 0.5)
)
names(metric_inputs) <- effect_specs$effect

effect_specs |>
  mutate(k = purrr::map_int(effect, ~ nrow(metric_inputs[[.x]]$data))) |>
  dplyr::select(`Effect size` = label, k) |>
  knitr::kable(caption = "Effect sizes available for the uni-moderator analyses.")
```

::: callout-note
## Moderator definitions

The moderator table lists every variable used in the uni-moderator analyses and its definition from the protocol codebook.

```{r}
#| label: moderator-list

moderators <- tribble(
  ~variable,                     ~label,                             ~definition,                                                                                                                          ~kind,        ~missing_label,
  "gest_stage_ex_standardised", "Timing of prenatal exposure",     "Gestation stage when dams were exposed to chronic noise: early_mid, full_gestation, mid_late, or late.",                                  "categorical", NA_character_,
  "exposure_span_d",            "Exposure duration (days)",        "Total number of days the dam was exposed to chronic noise.",                                                                          "continuous",  NA_character_,
  "exposure_type",              "Exposure type",                   "Intermittent when noise was played for some hours per day; continuous when noise was not turned off during the total exposure duration.", "categorical", NA_character_,
  "noise_frequency_range",      "Noise frequency range",           "Standardized acoustic frequency range of the exposure: audible, ultrasound, or unclear. We treat unclear as missing rather than as a moderator level.", "categorical", "unclear",
  "control_conditions",         "Control conditions",              "Conditions under which control animals were kept during noise exposure, such as silence, ambient sound, white noise, or other.",         "categorical", NA_character_,
  "offspring_sex",              "Offspring sex",                   "Sex of the offspring investigated: male, female, mixed, or not reported.",                                                            "categorical", NA_character_,
  "outcome_type",               "Measured outcome type",           "Primary behavioural construct used to classify whether an assay measures anxiety, depression, or both.",                               "categorical", NA_character_,
  "assay_type",                 "Behavioural assay type",          "Type of assay used to measure anxiety- or depression-like behaviour, as listed in the Protocol, Appendix 2.",                            "categorical", NA_character_,
  "offspring_age_d",            "Offspring age at testing (days)", "Offspring age in days from parturition at the start of a behavioural test. Ranges use the midpoint, and ages reported after conception are converted to postnatal days.", "continuous", NA_character_,
  "loudness_dB",                "Noise intensity: loudness in dB",  "Loudness of the noise in decibels. We retain only unweighted dB values. We excluded values reported in dBA, which applies a human-hearing frequency weighting and is not directly comparable with unweighted dB.", "continuous", NA_character_
)

# Stop if the database does not contain one of the planned moderators.
if (length(setdiff(moderators$variable, names(db))) > 0) {
  stop("The effect-size database does not contain all moderators.")
}

# Display the moderator names used in the models.
moderators |>
  dplyr::select(label, variable, definition) |>
  knitr::kable(
    col.names = c("Moderator", "Variable name", "Definition in codebook"),
    caption = "Moderators used in the uni-moderator analyses."
  )
```

:::

::: callout-note
## Helper functions

### Match category colours across plots

We assign one named colour to each category in the complete database. The named colours keep a category visually consistent when the retained levels differ among lnRR, lnVR, and lnCVR.

```{r}
#| label: moderator-colour-palettes
#| code-fold: true

# Retain the categorical moderators and their missing-value labels.
categorical_moderators <- moderators |>
  filter(kind == "categorical")

# Create one named colour palette for each categorical moderator.
category_palettes <- purrr::map2(
  categorical_moderators$variable,
  categorical_moderators$missing_label,
  function(variable, missing_label) {
    category_levels <- sort(unique(as.character(db[[variable]])))
    category_levels <- category_levels[!is.na(category_levels)]

    # Exclude a category that the analysis treats as missing.
    if (!is.na(missing_label)) {
      category_levels <- category_levels[category_levels != missing_label]
    }

    # orchaRd capitalizes the first letter of plotted moderator levels.
    # Name the palette with those displayed labels so ggplot can match every
    # category instead of falling back to grey. Because this palette is built
    # once per moderator, each level keeps its colour across all three metrics.
    plotted_levels <- category_levels
    substr(plotted_levels, 1, 1) <- toupper(substr(plotted_levels, 1, 1))

    stats::setNames(
      grDevices::hcl.colors(length(category_levels), palette = "Dark 3"),
      plotted_levels
    )
  }
)
names(category_palettes) <- categorical_moderators$variable
```

### Prepare the moderator data

For a categorical moderator, we remove levels with $k < 5$ and subset the same rows and columns from `V`. For a continuous moderator, we retain finite values and subset `V` in the same way.

```{r}
#| label: prepare-moderator-data
#| code-fold: true

prepare_categorical <- function(data, V, moderator, missing_label = NA_character_, min_k = 5) {
  # Start with rows that contain a recorded moderator value.
  usable <- !is.na(data[[moderator]])

  # Treat a specified text label as missing when required.
  if (!is.na(missing_label)) {
    usable <- usable & as.character(data[[moderator]]) != missing_label
  }

  # Count the effect sizes in each moderator level.
  level_counts <- data[usable, ] |>
    count(level = .data[[moderator]], name = "effect_sizes") |>
    mutate(
      studies = vapply(
        level,
        function(x) n_distinct(data$study_id[usable & data[[moderator]] == x]),
        integer(1)
      ),
      include = effect_sizes >= min_k
    )

  # Retain levels represented by at least five effect sizes.
  keep_levels <- level_counts |>
    filter(include) |>
    pull(level)

  # Use the same row positions to subset the data and VCV matrix.
  idx <- which(usable & data[[moderator]] %in% keep_levels)
  dat <- data[idx, , drop = FALSE]
  V_subset <- V[idx, idx, drop = FALSE]

  # Remove unused factor levels after filtering.
  dat[[moderator]] <- droplevels(factor(dat[[moderator]]))

  # A categorical moderator requires at least two retained levels.
  if (nlevels(dat[[moderator]]) < 2) {
    stop("Fewer than two levels remain for ", moderator, ".")
  }

  list(data = dat, V = V_subset, level_counts = level_counts)
}

prepare_continuous <- function(data, V, moderator) {
  # Retain rows with finite moderator values.
  idx <- which(is.finite(data[[moderator]]))

  # Subset the data and VCV matrix with the same row positions.
  dat <- data[idx, , drop = FALSE]
  V_subset <- V[idx, idx, drop = FALSE]

  # Store the moderator as a numeric variable.
  dat[[moderator]] <- as.numeric(dat[[moderator]])

  # A continuous moderator requires at least two distinct values.
  if (n_distinct(dat[[moderator]]) < 2) {
    stop("The moderator does not contain at least two values: ", moderator, ".")
  }

  list(data = dat, V = V_subset, level_counts = NULL)
}
```

### Fit one uni-moderator model

The function below prepares one moderator, subsets the complete VCV matrix, and fits its model. For categorical moderators with at least three retained levels, it also calculates unadjusted Tukey contrasts with a separate no-intercept model.

```{r}
#| label: fit-one-unimoderator
#| code-fold: true

fit_unimoderator <- function(
  data,
  V,
  yi,
  effect_label,
  effect_axis_label,
  moderator,
  label,
  kind,
  missing_label = NA_character_,
  min_k = 5
) {
  # Apply the appropriate preparation rule for the moderator type.
  prepared <- if (kind == "categorical") {
    prepare_categorical(data, V, moderator, missing_label, min_k)
  } else {
    prepare_continuous(data, V, moderator)
  }

  model_data <- prepared$data
  V_subset <- prepared$V

  # Fit one multilevel meta-regression with the subsetted VCV matrix.
  model <- metafor::rma.mv(
    yi = model_data[[yi]],
    V = V_subset,
    mods = reformulate(moderator),
    random = list(
      ~ 1 | study_id,
      ~ 1 | es_id,
      ~ 1 | strain
    ),
    test = "t",
    method = "REML",
    sparse = TRUE,
    data = model_data
  )

  # Draw an orchard plot for a categorical moderator. We draw continuous
  # moderators with metafor::regplot() when we display the model results.
  model_plot <- if (kind == "categorical") {
    orchaRd::orchard_plot(
      model,
      mod = moderator,
      group = "study_id",
      xlab = effect_axis_label,
      flip = TRUE,
      angle = 0,
      legend.pos = "top.left"
    ) +
      # Use the complete named palette so shared categories keep their colour.
      ggplot2::scale_colour_manual(
        values = category_palettes[[moderator]],
        drop = FALSE
      ) +
      ggplot2::scale_fill_manual(
        values = category_palettes[[moderator]],
        drop = FALSE
      )
  } else {
    NULL
  }

  contrasts <- NULL

  # Calculate unadjusted Tukey contrasts when at least three levels remain.
  if (kind == "categorical" && nlevels(model_data[[moderator]]) >= 3) {
    # Fit a no-intercept model so that each coefficient represents one level.
    no_intercept_model <- metafor::rma.mv(
      yi = model_data[[yi]],
      V = V_subset,
      mods = reformulate(moderator, intercept = FALSE),
      random = list(
        ~ 1 | study_id,
        ~ 1 | es_id,
        ~ 1 | strain
      ),
      test = "t",
      method = "REML",
      sparse = TRUE,
      data = model_data
    )

    # Create all pairwise Tukey contrasts among the retained levels.
    contrast_matrix <- multcomp::contrMat(
      table(model_data[[moderator]]),
      type = "Tukey"
    )
    colnames(contrast_matrix) <- names(coef(no_intercept_model))

    # Report the unadjusted p-value for each contrast.
    contrasts <- summary(
      multcomp::glht(no_intercept_model, linfct = contrast_matrix),
      test = multcomp::adjusted("none")
    )
  }

  list(
    effect = yi,
    effect_label = effect_label,
    effect_axis_label = effect_axis_label,
    variable = moderator,
    label = label,
    kind = kind,
    data = model_data,
    V = V_subset,
    level_counts = prepared$level_counts,
    model = model,
    r2 = orchaRd::r2_ml(model),
    plot = model_plot,
    contrasts = contrasts
  )
}
```

### Fit all uni-moderator models

We call `fit_unimoderator()` once for each combination of effect-size type and moderator. The function subsets the corresponding complete VCV matrix to match the effect sizes retained for each model.

```{r}
#| label: fit-all-unimoderators
#| code-fold: true

# Fit every moderator separately for lnRR, lnVR, and lnCVR.
unimoderator_results <- purrr::map(
  effect_specs$effect,
  function(effect) {
    effect_label <- effect_specs$label[effect_specs$effect == effect]
    effect_axis_label <- effect_specs$axis_label[effect_specs$effect == effect]
    input <- metric_inputs[[effect]]

    results <- purrr::pmap(
      moderators |>
        dplyr::select(variable, label, kind, missing_label),
      function(variable, label, kind, missing_label) {
        fit_unimoderator(
          data = input$data,
          V = input$V,
          yi = effect,
          effect_label = effect_label,
          effect_axis_label = effect_axis_label,
          moderator = variable,
          label = label,
          kind = kind,
          missing_label = missing_label,
          min_k = 5
        )
      }
    )

    names(results) <- moderators$variable
    results
  }
)

# Name the outer list by effect-size type.
names(unimoderator_results) <- effect_specs$effect
```
:::

For each moderator, we present lnRR and lnVR side by side, followed by lnCVR below. Each result contains the model summary, multilevel $R^2$, and plot. We use orchard plots for categorical moderators and `metafor::regplot()` for continuous moderators. For categorical moderators with three or more levels, we also print the unadjusted pairwise contrasts.

```{r}
#| label: display-unimoderator-results
#| results: asis
#| echo: false
#| fig-width: 8
#| fig-height: 5

# Print one effect-size result within the current layout column.
print_metric_result <- function(result) {
  cat("### ", result$effect_label, "\n\n", sep = "")

  # Print the model summary.
  cat("```\n")
  print(summary(result$model))
  cat("```\n\n")

  # Print marginal and conditional multilevel R-squared.
  cat("```\n")
  print(round(result$r2, 4))
  cat("```\n\n")

  # Place each plot in a card.
  cat("::: card\n\n")
  if (result$kind == "categorical") {
    print(result$plot)
  } else {
    # Plot the marginal regression line and its 95% confidence interval.
    metafor::regplot(
      result$model,
      mod = result$variable,
      pred = TRUE,
      ci = TRUE,
      pi = FALSE,
      xlab = result$label,
      ylab = result$effect_axis_label,
      psize = "seinv",
      plim = c(0.7, 2.2),
      refline = 0
    )
  }
  cat("\n\n:::\n\n")

  # Print pairwise contrasts only when the function created them.
  if (!is.null(result$contrasts)) {
    cat("We report unadjusted Tukey contrasts between the retained levels.\n\n```\n")
    print(result$contrasts)
    cat("```\n\n")
  }
}

# Use the same three-panel layout for every moderator.
for (moderator in moderators$variable) {
  moderator_label <- moderators$label[moderators$variable == moderator]
  cat("\n\n## ", moderator_label, "\n\n", sep = "")

  # Place lnRR and lnVR in the first row.
  cat(":::: columns\n\n")
  cat(":::: {.column width=\"49%\"}\n\n")
  print_metric_result(unimoderator_results$lnRR[[moderator]])
  cat("::::\n\n")

  cat(":::: {.column width=\"2%\"}\n::::\n\n")

  cat(":::: {.column width=\"49%\"}\n\n")
  print_metric_result(unimoderator_results$lnVR[[moderator]])
  cat("::::\n\n")
  cat("::::\n\n")

  # Centre lnCVR in the second row.
  cat(":::: columns\n\n")
  cat(":::: {.column width=\"25%\"}\n::::\n\n")

  cat(":::: {.column width=\"50%\"}\n\n")
  print_metric_result(unimoderator_results$lnCVR[[moderator]])
  cat("::::\n\n")

  cat(":::: {.column width=\"25%\"}\n::::\n\n")
  cat("::::\n\n")
}
```

::: callout-note
## Session information

```{r}
#| label: session-information

sessionInfo()
```
:::
