---
title: "CRIME-Q appraisal"
---

We summarize the finalized CRIME-Q appraisal for the included studies. We use the adapted items defined in the appraisal workbook to assess quality of reporting, methodological quality, and risk of bias [@andersen2024]. The code below reads the item definitions and ratings from the same workbook so their order remains aligned.

::: callout-note
## Variable definitions

| Variable | Definition |
|:--|:--|
| `study_id` | Unique identifier for each included study. |
| QoR | Quality of reporting. |
| MQ | Methodological quality. |
| RoB | Risk of bias. |
| Rating | Recorded response for an appraisal item: `Yes`, `Partly`, `No`, `Unclear`, or `NA`, where applicable. |
:::

```{r}
#| label: setup
#| include: false

pacman::p_load(
  dplyr,
  ggplot2,
  knitr,
  readxl,
  stringr,
  tibble,
  tidyr
)

knitr::opts_chunk$set(comment = "#>")

# Define the location of the CRIME-Q workbook.
project_root <- normalizePath("..", mustWork = TRUE)
input_file <- file.path(
  project_root,
  "Data",
  "CRIME-Q_Appraisal_Template.xlsx"
)
```

We load the finalized study ratings and adapted item definitions from the CRIME-Q workbook.

```{r}
#| label: load-crime-q
#| code-fold: true
#| code-summary: "Show data-loading code"

# Read the finalized study-level ratings.
appraisal <- readxl::read_excel(
  path = input_file,
  sheet = "CRIME-Q_Appraisal",
  col_types = "text",
  na = ""
) |>
  dplyr::mutate(
    dplyr::across(
      dplyr::everything(),
      ~ dplyr::na_if(stringr::str_squish(.x), "")
    )
  )

# Read the codebook table below the rating legend.
codebook_raw <- readxl::read_excel(
  path = input_file,
  sheet = "Codebook",
  skip = 7,
  col_types = "text",
  na = ""
)

# Assign concise names to the five codebook columns.
names(codebook_raw)[1:5] <- c(
  "item",
  "item_name",
  "domain",
  "question",
  "criteria"
)

# Match the rating options recorded at the end of each question.
rating_pattern <- "(Yes|Partly|No|Unclear|NA)(/(Yes|Partly|No|Unclear|NA))+$"

# Keep the 21 scored items and omit their paired notes fields.
codebook <- codebook_raw |>
  dplyr::mutate(
    dplyr::across(
      dplyr::everything(),
      ~ dplyr::na_if(stringr::str_squish(.x), "")
    )
  ) |>
  dplyr::filter(
    domain %in% c("QoR", "MQ", "RoB"),
    !stringr::str_detect(item, "n$|_n$")
  ) |>
  dplyr::mutate(
    item_key = stringr::str_remove_all(item, "[^[:alnum:]]"),
    ratings = stringr::str_replace_all(
      stringr::str_extract(question, rating_pattern),
      "/",
      "; "
    ),
    definition = stringr::str_squish(
      stringr::str_remove(
        question,
        paste0("\\s*", rating_pattern)
      )
    )
  )
```

```{r}
#| label: validate-crime-q
#| code-fold: true
#| code-summary: "Show validation code"
#| results: hide

# Select rating columns while excluding notes and appraisal time.
score_columns <- names(appraisal)[
  stringr::str_detect(names(appraisal), "^[0-9]+[A-Z]") &
    !stringr::str_detect(names(appraisal), "notes$")
]

# Match workbook column prefixes to the item codes in the codebook.
score_keys <- stringr::str_extract(score_columns, "^[^_]+")

if (!identical(score_keys, codebook$item_key)) {
  stop("The CRIME-Q rating columns do not match the codebook item order.")
}

if (anyDuplicated(appraisal$study_id) > 0L || anyNA(appraisal$study_id)) {
  stop("Each CRIME-Q row must have one unique study_id.")
}

# Convert the rating matrix to long format for validation and plotting.
item_map <- tibble::tibble(
  score_column = score_columns,
  item = codebook$item,
  domain = codebook$domain
)

plot_data <- appraisal |>
  dplyr::select(study_id, dplyr::all_of(score_columns)) |>
  tidyr::pivot_longer(
    cols = dplyr::all_of(score_columns),
    names_to = "score_column",
    values_to = "rating"
  ) |>
  dplyr::left_join(item_map, by = "score_column")

valid_ratings <- c("Yes", "Partly", "No", "Unclear", "NA")

if (anyNA(plot_data$rating)) {
  stop("The CRIME-Q appraisal contains a missing rating.")
}

unexpected_ratings <- setdiff(unique(plot_data$rating), valid_ratings)

if (length(unexpected_ratings) > 0L) {
  stop(
    "Unexpected CRIME-Q rating: ",
    paste(unexpected_ratings, collapse = ", ")
  )
}

if (anyNA(codebook$ratings)) {
  stop("A CRIME-Q item does not contain a recorded rating list.")
}
```

## CRIME-Q item definitions

The table defines each adapted CRIME-Q item used in the figure and lists its allowed ratings.

```{r}
#| label: crime-q-definitions
#| results: asis
#| code-fold: true
#| code-summary: "Show item-table code"

codebook |>
  dplyr::transmute(
    Item = item,
    Domain = domain,
    Definition = definition,
    Ratings = ratings
  ) |>
  knitr::kable(
    format = "html",
    align = c("l", "l", "l", "l"),
    escape = TRUE,
    table.attr = 'class="table table-striped table-hover crime-q-codebook"'
  )
```

## Study-level ratings

The workbook contains `r dplyr::n_distinct(appraisal$study_id)` studies and `r length(score_columns)` scored CRIME-Q items. We plot the finalized ratings below. Notes and appraisal time remain in the workbook but do not enter the figure.

```{r}
#| label: crime-q-heatmap
#| fig-width: 13
#| fig-height: 8
#| out-width: "100%"
#| fig-cap: "Final CRIME-Q ratings for each included study and adapted appraisal item."
#| code-fold: true
#| code-summary: "Show figure code"

# Create a concise study label from the study identifier.
study_labels <- appraisal |>
  dplyr::transmute(
    study_id,
    study = paste(
      stringr::str_to_title(stringr::str_extract(study_id, "^[^_]+")),
      "et al.",
      stringr::str_extract(study_id, "(19|20)[0-9]{2}")
    )
  ) |>
  dplyr::arrange(study_id)

# Set the item and study order used in the heatmap.
plot_data <- plot_data |>
  dplyr::left_join(study_labels, by = "study_id") |>
  dplyr::mutate(
    item = factor(item, levels = codebook$item),
    study = factor(study, levels = rev(study_labels$study)),
    rating = factor(rating, levels = valid_ratings)
  )

# Use one colour for each recorded rating.
rating_colours <- c(
  "Yes" = "#009E60",
  "Partly" = "#EFBF04",
  "No" = "#DC143C",
  "Unclear" = "#F4B183",
  "NA" = "#D9D9D9"
)

# Plot one tile for every study-item rating.
ggplot2::ggplot(
  plot_data,
  ggplot2::aes(x = item, y = study, fill = rating)
) +
  ggplot2::geom_tile(color = "#FFFDF8", linewidth = 0.65) +
  ggplot2::scale_x_discrete(position = "top") +
  ggplot2::scale_fill_manual(
    values = rating_colours,
    breaks = valid_ratings,
    drop = FALSE,
    name = NULL
  ) +
  ggplot2::labs(x = NULL, y = NULL) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_text(
      face = "bold",
      colour = "#242321",
      size = 10
    ),
    axis.text.y = ggplot2::element_text(
      face = "bold",
      colour = "#242321",
      size = 10
    ),
    axis.ticks = ggplot2::element_blank(),
    legend.position = "bottom",
    legend.text = ggplot2::element_text(size = 11),
    legend.key.width = grid::unit(1.5, "lines"),
    plot.margin = ggplot2::margin(10, 10, 10, 10)
  )
```

```{css}
.crime-q-codebook {
  width: 100%;
  font-size: 0.93rem;
}

.crime-q-codebook th {
  background: #eef2f4;
  color: #24313a;
  vertical-align: bottom;
}

.crime-q-codebook td {
  vertical-align: top;
}

.crime-q-codebook th:first-child,
.crime-q-codebook td:first-child,
.crime-q-codebook th:nth-child(3),
.crime-q-codebook td:nth-child(3) {
  white-space: nowrap;
}
```

::: {.callout-note collapse="true"}
## Session information

```{r}
#| label: session-information

sessionInfo()
```
:::
