---
title: "Leave-one-out analysis"
---

We use a leave-one-out analysis to examine whether the overall lnRR estimate depends on any single study. We remove one study at a time, reconstruct the sampling variance-covariance matrix (`V`), and refit the multilevel intercept-only model.

::: callout-note
## Variable definitions

| Variable | Definition |
|:--|:--|
| `study_id` | Identifier for the primary research paper and the unit removed in each leave-one-out iteration. |
| `es_id` | Unique identifier for each effect size. |
| `ex_id`, `con_id` | Identifiers for the exposed and control groups used to construct `V`. |
| `lnRR` | Log response ratio for the difference in group means. |
| `lnRR_var` | Sampling variance of lnRR. |
| $k$ | Number of effect sizes included in a model. |
:::

```{r}
#| label: setup
#| include: false

pacman::p_load(
  dplyr,
  ggplot2,
  knitr,
  metafor,
  orchaRd,
  purrr,
  readr,
  tibble
)

knitr::opts_chunk$set(comment = "#>")

project_root <- normalizePath("..", mustWork = TRUE)
derived_dir <- file.path(project_root, "Data", "derived")
audit_dir <- file.path(project_root, "Data", "audits")
```

# Analysis data and model functions

We load the effect-size database created in the **Effect size calculation**. The model functions reconstruct `V` with the variables and working correlation used in the **Sampling variance-covariance matrices**. Then, we fit the random-effects structure specified in the **Overall effects**.

```{r}
#| label: load-data-and-functions

# Load the lnRR effect sizes and sampling variances.
db <- readr::read_csv(
  file.path(derived_dir, "db_effect_sizes.csv"),
  show_col_types = FALSE
)

# Construct the lnRR sampling variance-covariance matrix for any subset.
construct_lnRR_vcv <- function(data) {
  metafor::vcalc(
    vi = lnRR_var,
    cluster = study_id,
    grp1 = ex_id,
    grp2 = con_id,
    w1 = ex_n,
    w2 = c_n,
    obs = es_id,
    rho = 0.5,
    checkpd = TRUE,
    data = data
  )
}

# Reconstruct V and fit the intercept-only lnRR model for any subset.
fit_lnRR_model <- function(data) {
  V <- construct_lnRR_vcv(data)

  metafor::rma.mv(
    yi = lnRR,
    V = V,
    random = list(
      ~ 1 | study_id,
      ~ 1 | es_id,
      ~ 1 | strain
    ),
    test = "t",
    method = "REML",
    sparse = TRUE,
    data = data
  )
}

# Fit the overall model before applying any exclusions.
overall_lnRR_model <- fit_lnRR_model(db)
```

# Study-level leave-one-out analysis

Each iteration removes every effect size from one `study_id`. Because the number and arrangement of dependent effect sizes change after each exclusion, we reconstruct `V` before fitting that iteration's model.

```{r}
#| label: fit-leave-one-out-models

# Fit one model for each omitted study.
study_ids <- unique(db$study_id)

loo_models <- purrr::map(
  study_ids,
  ~ fit_lnRR_model(dplyr::filter(db, study_id != .x))
)
names(loo_models) <- study_ids

# Extract the estimate and intervals returned by each fitted model.
loo_model_results <- purrr::imap(
  loo_models,
  function(model, omitted_study) {
    result <- orchaRd::mod_results(model, group = "study_id")
    model_table <- result$mod_table
    model_data <- result$data
    model_table$name <- omitted_study
    model_data$moderator <- omitted_study
    list(mod_table = model_table, data = model_data)
  }
)

# Assemble the object used by the leave-one-out orchard plot.
loo_results <- structure(
  list(
    mod_table = dplyr::bind_rows(
      purrr::map(loo_model_results, "mod_table")
    ),
    data = dplyr::bind_rows(
      purrr::map(loo_model_results, "data")
    ),
    orig_mod_results = orchaRd::mod_results(
      overall_lnRR_model,
      group = "study_id"
    )
  ),
  class = c("orchard", "data.frame")
)

# Create a compact table with one row for each omitted study.
loo_table <- loo_results$mod_table |>
  dplyr::transmute(
    `Study removed` = as.character(name),
    k = nrow(db) - as.integer(table(db$study_id)[as.character(name)]),
    `Estimate` = estimate,
    `95% CI lower` = lowerCL,
    `95% CI upper` = upperCL
  )

loo_table |>
  knitr::kable(
    digits = 3,
    caption = "Overall lnRR after removing each study in turn."
  )
```

The orchard plot shows the refitted overall lnRR estimate for each omitted study. The dotted reference lines show the confidence limits from the model that includes all studies [@nakagawa2023b].

```{r}
#| label: leave-one-out-orchard-plot
#| fig-cap: "Study-level leave-one-out analysis for the overall lnRR."
#| fig-width: 8
#| fig-height: 8

# Plot the model estimate obtained after each study exclusion.
loo_plot <- orchaRd::orchard_leave1out(
  leave1out = loo_results,
  xlab = "lnRR",
  ylab = "Study removed",
  ci_lines_color = "grey40",
  trunk.size = 0.3,
  branch.size = 1.5,
  alpha = 0.35,
  legend.pos = "top.out"
)

loo_plot
```

```{r}
#| label: export-leave-one-out-results
#| include: false

# Export the numerical results without replacing the earlier audit file.
readr::write_csv(
  loo_table,
  file.path(audit_dir, "leave_one_out_lnRR_current.csv"),
  na = ""
)
```

::: callout-note
## Session information

```{r}
#| label: session-information

sessionInfo()
```
:::
