---
title: "Sensitivity analysis"
---

We use two sensitivity analyses to examine whether specific data-inclusion decisions influence the overall lnRR estimate. First, we exclude the four Abramova-laboratory studies together. Second, we exclude the effect sizes that required exact-zero boundary corrections.

::: callout-note
## Variable definitions

| Variable | Definition |
|:--|:--|
| `study_id` | Identifier for the primary research paper. |
| `es_id` | Unique identifier for each effect size. |
| `ex_id`, `con_id` | Identifiers for the exposed and control groups used to construct `V`. |
| `lnRR` | Log response ratio for the difference in group means. |
| `lnRR_var` | Sampling variance of lnRR. |
| `c_mean_continuity_corrected` | Indicates that the control-group mean required a boundary correction. |
| `c_sd_continuity_corrected` | Indicates that the control-group SD required a boundary correction. |
| $k$ | Number of effect sizes included in a model. |
:::

```{r}
#| label: setup
#| include: false

pacman::p_load(
  dplyr,
  ggplot2,
  knitr,
  metafor,
  orchaRd,
  patchwork,
  readr,
  tibble
)

knitr::opts_chunk$set(comment = "#>")

project_root <- normalizePath("..", mustWork = TRUE)
derived_dir <- file.path(project_root, "Data", "derived")
audit_dir <- file.path(project_root, "Data", "audits")
```

# Analysis data and model functions

We load the effect-size database created in the **Effect size calculation**. The model functions reconstruct `V` with the variables and working correlation used in the **Sampling variance-covariance matrices**. Then, we fit the random-effects structure specified in the **Overall effects**.

```{r}
#| label: load-data-and-functions

# Load the complete lnRR effect-size database.
db <- readr::read_csv(
  file.path(derived_dir, "db_effect_sizes.csv"),
  show_col_types = FALSE
)

# Construct the lnRR sampling variance-covariance matrix for any subset.
construct_lnRR_vcv <- function(data) {
  metafor::vcalc(
    vi = lnRR_var,
    cluster = study_id,
    grp1 = ex_id,
    grp2 = con_id,
    w1 = ex_n,
    w2 = c_n,
    obs = es_id,
    rho = 0.5,
    checkpd = TRUE,
    data = data
  )
}

# Reconstruct V and fit the intercept-only lnRR model for any subset.
fit_lnRR_model <- function(data) {
  V <- construct_lnRR_vcv(data)

  metafor::rma.mv(
    yi = lnRR,
    V = V,
    random = list(
      ~ 1 | study_id,
      ~ 1 | es_id,
      ~ 1 | strain
    ),
    test = "t",
    method = "REML",
    sparse = TRUE,
    data = data
  )
}

# Fit the overall lnRR model before applying either exclusion.
model_all <- fit_lnRR_model(db)

# Identify the Abramova-laboratory studies and count their current effect sizes.
abramova_studies <- db |>
  dplyr::filter(grepl("^abramova_", study_id)) |>
  dplyr::distinct(study_id) |>
  dplyr::pull(study_id)

abramova_effect_sizes <- sum(db$study_id %in% abramova_studies)
```

# Combined Abramova-laboratory exclusion

The `r length(abramova_studies)` Abramova-laboratory studies contribute `r abramova_effect_sizes` of the `r nrow(db)` effect sizes. When studies from one laboratory contribute a large proportion of the data, laboratory-specific methods, experimental conditions, or reporting practices may have a disproportionate influence on the overall estimate. To examine this possibility, we remove all four studies together, reconstruct `V`, and compare the refitted model with the model that includes all studies.

```{r}
#| label: exclude-abramova-laboratory

# Remove all studies with the Abramova study identifier.
db_without_abramova <- db |>
  dplyr::filter(!study_id %in% abramova_studies)

# Reconstruct V and refit the lnRR model after the combined exclusion.
model_without_abramova <- fit_lnRR_model(db_without_abramova)
```

We use the same lnRR scale in both panels. Panel A shows the original overall-effect model, and Panel B shows the model after excluding the four Abramova-laboratory studies.

```{r}
#| label: abramova-overall-effect-comparison
#| fig-cap: "Overall lnRR before and after the combined Abramova-laboratory exclusion. Both panels use the same effect-size scale."
#| fig-width: 12
#| fig-height: 5.5

# Set one effect-size range for both orchard plots.
shared_effect_limits <- range(
  c(
    db$lnRR,
    model_all$pi.lb,
    model_all$pi.ub,
    model_without_abramova$pi.lb,
    model_without_abramova$pi.ub
  ),
  finite = TRUE
)
effect_padding <- diff(shared_effect_limits) * 0.05
shared_effect_limits <- shared_effect_limits + c(
  -effect_padding,
  effect_padding
)
shared_effect_breaks <- pretty(shared_effect_limits, n = 5)

# Draw the overall model that includes all studies.
plot_all <- orchaRd::orchard_plot(
  model_all,
  group = "study_id",
  xlab = "lnRR",
  flip = TRUE
) +
  ggplot2::scale_x_discrete(labels = "Overall effect") +
  ggplot2::scale_y_continuous(
    limits = shared_effect_limits,
    breaks = shared_effect_breaks
  ) +
  ggplot2::labs(title = "All studies")

# Draw the model after excluding the Abramova-laboratory studies.
plot_without_abramova <- orchaRd::orchard_plot(
  model_without_abramova,
  group = "study_id",
  xlab = "lnRR",
  flip = TRUE
) +
  ggplot2::scale_x_discrete(labels = "Overall effect") +
  ggplot2::scale_y_continuous(
    limits = shared_effect_limits,
    breaks = shared_effect_breaks
  ) +
  ggplot2::labs(title = "Abramova-laboratory studies excluded")

# Place the original and sensitivity models side by side.
abramova_comparison_plot <-
  (plot_all + plot_without_abramova) +
  patchwork::plot_layout(guides = "collect") +
  patchwork::plot_annotation(tag_levels = "A") &
  ggplot2::theme(legend.position = "top")

abramova_comparison_plot
```

```{r}
#| label: abramova-model-comparison

# Compare the numerical results shown in the two panels.
abramova_comparison <- tibble::tibble(
  Model = c(
    "All studies",
    "Abramova-laboratory studies excluded"
  ),
  k = c(nrow(db), nrow(db_without_abramova)),
  Studies = c(
    dplyr::n_distinct(db$study_id),
    dplyr::n_distinct(db_without_abramova$study_id)
  ),
  Estimate = c(
    unname(stats::coef(model_all)),
    unname(stats::coef(model_without_abramova))
  ),
  `95% CI lower` = c(
    model_all$ci.lb,
    model_without_abramova$ci.lb
  ),
  `95% CI upper` = c(
    model_all$ci.ub,
    model_without_abramova$ci.ub
  )
)

abramova_comparison |>
  knitr::kable(
    digits = 3,
    caption = "Overall lnRR with and without the Abramova-laboratory studies."
  )
```

# Exact-zero effect-size exclusion

Two effect sizes required a boundary correction because the reported control-group summary statistics produced an exact-zero mean and SD. We identify these effect sizes, remove them, reconstruct `V`, and refit the multilevel intercept-only model.

```{r}
#| label: exact-zero-exclusion

# Identify rows that received both control-group boundary corrections.
exact_zero_ids <- db |>
  dplyr::filter(
    c_mean_continuity_corrected,
    c_sd_continuity_corrected
  ) |>
  dplyr::pull(es_id)

if (length(exact_zero_ids) == 0) {
  stop("No exact-zero-corrected effect sizes were identified.")
}

# Retain all other effect sizes for the sensitivity model.
db_without_exact_zero <- db |>
  dplyr::filter(!es_id %in% exact_zero_ids)

# Reconstruct V and refit the model after excluding the corrected rows.
model_without_exact_zero <- fit_lnRR_model(db_without_exact_zero)
```

We report the identifiers of the excluded effect sizes and compare the overall lnRR estimates before and after the exclusion.

```{r}
#| label: exact-zero-comparison

# List the effect sizes removed from the sensitivity model.
tibble::tibble(`Effect size excluded` = exact_zero_ids) |>
  knitr::kable(caption = "Effect sizes excluded from the sensitivity model.")

# Create the model-comparison table.
exact_zero_comparison <- tibble::tibble(
  Model = c(
    "All effect sizes",
    "Exact-zero-corrected effect sizes excluded"
  ),
  k = c(nrow(db), nrow(db_without_exact_zero)),
  Studies = c(
    dplyr::n_distinct(db$study_id),
    dplyr::n_distinct(db_without_exact_zero$study_id)
  ),
  Estimate = c(
    unname(stats::coef(model_all)),
    unname(stats::coef(model_without_exact_zero))
  ),
  `95% CI lower` = c(
    model_all$ci.lb,
    model_without_exact_zero$ci.lb
  ),
  `95% CI upper` = c(
    model_all$ci.ub,
    model_without_exact_zero$ci.ub
  )
)

exact_zero_comparison |>
  knitr::kable(
    digits = 3,
    caption = "Overall lnRR with and without the exact-zero-corrected effect sizes."
  )
```

```{r}
#| label: export-sensitivity-results
#| include: false

# Export both comparisons without replacing the earlier audit files.
readr::write_csv(
  abramova_comparison,
  file.path(audit_dir, "abramova_laboratory_exclusion_lnRR_current.csv"),
  na = ""
)
readr::write_csv(
  exact_zero_comparison,
  file.path(audit_dir, "exact_zero_exclusion_lnRR_current.csv"),
  na = ""
)
```

::: callout-note
## Session information

```{r}
#| label: session-information

sessionInfo()
```
:::
