---
title: "Alluvial plots"
---

We use study-level alluvial plots to summarize the composition of the evidence base. Each study contributes a total width of one, regardless of how many effect sizes it reports. When a study reports more than one value for a characteristic, we combine those values into a study-level coverage category.

```{r}
#| label: setup
#| include: false

pacman::p_load(
  dplyr,
  ggalluvial,
  ggplot2,
  readr,
  stringr,
  tidyr
)

project_root <- normalizePath("..", mustWork = TRUE)
derived_dir <- file.path(project_root, "Data", "derived")

# Load the effect-size database used throughout the supplement.
db <- readr::read_csv(
  file.path(derived_dir, "db_effect_sizes.csv"),
  show_col_types = FALSE
)
```

# Study-level categories

We first reduce the effect-size database to one row per study. The derived coverage variables retain whether a study reports one or several outcome types, offspring sexes, gestational exposure stages, or control conditions.

```{r}
#| label: create-study-level-data

# Combine the distinct values reported within each study.
combine_values <- function(x) {
  x <- trimws(as.character(x))
  x <- x[!is.na(x) & x != ""]
  paste(sort(unique(x)), collapse = "|")
}

study_data <- db |>
  dplyr::group_by(study_id) |>
  dplyr::summarise(
    outcome_values = combine_values(outcome_type),
    species = combine_values(sp_latin),
    strain = combine_values(strain),
    sex_values = combine_values(offspring_sex),
    gestational_stage_values = combine_values(
      gest_stage_ex_standardised
    ),
    exposure_type = combine_values(exposure_type),
    noise_type = combine_values(noise_type),
    control_values = combine_values(control_conditions),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    outcome_coverage = dplyr::case_when(
      outcome_values == "anxiety" ~ "Anxiety only",
      outcome_values == "depression" ~ "Depression only",
      outcome_values == "anxiety|depression" ~ "Anxiety and depression",
      TRUE ~ "Multiple outcome categories"
    ),
    sex_coverage = dplyr::case_when(
      sex_values == "male" ~ "Male only",
      sex_values == "female" ~ "Female only",
      sex_values == "female|male" ~ "Both sexes",
      sex_values == "mixed" ~ "Mixed/not separated",
      TRUE ~ "Multiple sex categories"
    ),
    gestational_stage_coverage = dplyr::case_when(
      stringr::str_detect(gestational_stage_values, "\\|") ~
        "Multiple stages",
      gestational_stage_values == "early_mid" ~ "Early–mid",
      gestational_stage_values == "full_gestation" ~ "Full gestation",
      gestational_stage_values == "mid_late" ~ "Mid–late",
      gestational_stage_values == "late" ~ "Late",
      TRUE ~ "Not reported"
    ),
    exposure_type = stringr::str_to_sentence(exposure_type),
    noise_type = stringr::str_to_sentence(noise_type),
    control_coverage = dplyr::case_when(
      stringr::str_detect(control_values, "\\|") ~
        "Multiple conditions",
      control_values == "ambient_sound" ~ "Ambient sound",
      control_values == "silence" ~ "Silence",
      control_values == "unclear" ~ "Unclear",
      TRUE ~ "Not reported"
    )
  )

if (nrow(study_data) != dplyr::n_distinct(db$study_id)) {
  stop("The study-level database must contain one row per study_id.")
}
```

# Study subjects and outcome coverage

The first plot connects study-level outcome coverage, species, and strain. Ribbon colour represents the offspring sexes reported by each study.

```{r}
#| label: study-subjects-alluvial
#| fig-cap: "Study-level distribution of outcome coverage, species, strain, and offspring sex. Ribbon width represents the number of studies."
#| fig-width: 10
#| fig-height: 6

# Count studies that follow the same path through the plot.
subject_flows <- study_data |>
  dplyr::count(
    outcome_coverage,
    species,
    strain,
    sex_coverage,
    name = "Studies"
  )

sex_colours <- c(
  "Male only" = "#E69F00",
  "Female only" = "#0072B2",
  "Both sexes" = "#009E73",
  "Mixed/not separated" = "#999999",
  "Multiple sex categories" = "#CC79A7"
)

study_subjects_plot <- ggplot2::ggplot(
  subject_flows,
  ggplot2::aes(
    y = Studies,
    axis1 = outcome_coverage,
    axis2 = species,
    axis3 = strain
  )
) +
  ggalluvial::geom_alluvium(
    ggplot2::aes(fill = sex_coverage),
    width = 0.22,
    alpha = 0.65
  ) +
  ggalluvial::geom_stratum(
    width = 0.22,
    fill = "#F1EFE8",
    colour = "#4A4A4A"
  ) +
  ggplot2::geom_text(
    stat = "stratum",
    ggplot2::aes(label = ggplot2::after_stat(stratum)),
    size = 3.2,
    fontface = "bold"
  ) +
  ggplot2::scale_x_discrete(
    limits = c("Outcome coverage", "Species", "Strain"),
    expand = c(0.08, 0.08)
  ) +
  ggplot2::scale_fill_manual(
    values = sex_colours,
    name = "Sex coverage"
  ) +
  ggplot2::labs(x = NULL, y = "Number of studies") +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    axis.text.y = ggplot2::element_blank(),
    axis.ticks.y = ggplot2::element_blank(),
    panel.grid = ggplot2::element_blank(),
    legend.position = "top"
  )

study_subjects_plot
```

# Prenatal-noise exposure design

The second plot connects gestational-stage coverage, exposure type, and noise type. Ribbon colour represents the control conditions reported by each study.

```{r}
#| label: exposure-design-alluvial
#| fig-cap: "Study-level distribution of prenatal-noise exposure and control designs. Ribbon width represents the number of studies."
#| fig-width: 10
#| fig-height: 6

# Count studies that follow the same exposure-design path.
exposure_flows <- study_data |>
  dplyr::count(
    gestational_stage_coverage,
    exposure_type,
    noise_type,
    control_coverage,
    name = "Studies"
  )

control_colours <- c(
  "Ambient sound" = "#56B4E9",
  "Silence" = "#009E73",
  "Unclear" = "#999999",
  "Multiple conditions" = "#CC79A7",
  "Not reported" = "#D9D9D9"
)

exposure_design_plot <- ggplot2::ggplot(
  exposure_flows,
  ggplot2::aes(
    y = Studies,
    axis1 = gestational_stage_coverage,
    axis2 = exposure_type,
    axis3 = noise_type
  )
) +
  ggalluvial::geom_alluvium(
    ggplot2::aes(fill = control_coverage),
    width = 0.22,
    alpha = 0.65
  ) +
  ggalluvial::geom_stratum(
    width = 0.22,
    fill = "#F1EFE8",
    colour = "#4A4A4A"
  ) +
  ggplot2::geom_text(
    stat = "stratum",
    ggplot2::aes(label = ggplot2::after_stat(stratum)),
    size = 3.2,
    fontface = "bold"
  ) +
  ggplot2::scale_x_discrete(
    limits = c(
      "Gestational-stage coverage",
      "Exposure type",
      "Noise type"
    ),
    expand = c(0.08, 0.08)
  ) +
  ggplot2::scale_fill_manual(
    values = control_colours,
    name = "Control conditions"
  ) +
  ggplot2::labs(x = NULL, y = "Number of studies") +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    axis.text.y = ggplot2::element_blank(),
    axis.ticks.y = ggplot2::element_blank(),
    panel.grid = ggplot2::element_blank(),
    legend.position = "top"
  )

exposure_design_plot
```

::: callout-note
## Session information

```{r}
#| label: session-information

sessionInfo()
```
:::
